---
title: "Supplementary Material: Cardiovascular risk and hippocampal-cognitive coupling in Alzheimer's disease"
author:
  - name: Sofia Fernandez-Lozano
    orcid: 0000-0003-2288-8111
    affiliations:
      - ref: bic
      - ref: ipn
  - name: Sylvia Villeneuve
    orcid: 0000-0003-2338-0467
    affiliations:
      - ref: bic
      - ref: psych
      - ref: stopad
  - name: D. Louis Collins
    orcid: 0000-0002-8432-7021
    affiliations:
      - ref: bic
      - ref: ipn
      - ref: bbme
    email: louis.collins@mcgill.ca
    corresponding: true
  - name: for the Alzheimer's Disease Neuroimaging Initiative
    affiliations: []
affiliations:
  - id: bic
    name: McConnell Brain Imaging Centre
    department: Montreal Neurological Institute
    city: Montreal
    state: Quebec
    country: Canada
  - id: ipn
    name: McGill University
    department: Department of Neurology and Neurosurgery
    city: Montreal
    state: Quebec
    country: Canada
  - id: psych
    name: McGill University
    department: Department of Psychiatry
    city: Montreal
    state: Quebec
    country: Canada
  - id: stopad
    name: Douglas Mental Health Institute Research Centre
    department: StoP-AD Centre
    city: Montreal
    state: Quebec
    country: Canada
  - id: bbme
    name: McGill University
    department: Department of Biomedical Engineering
    city: Montreal
    state: Quebec
    country: Canada
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
  pdf:
    keep-tex: true
    toc: false
    number-sections: true
    documentclass: article
    pdf-engine: xelatex
    geometry:
      - margin=1in
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage{tikz}
        \usetikzlibrary{shapes.geometric, arrows.meta, positioning}
        \usepackage{authblk}
        \renewcommand\Authfont{\normalsize}
        \renewcommand\Affilfont{\small\itshape}
        \providecommand{\orcidlink}[1]{\href{https://orcid.org/#1}{\textsuperscript{\textsf{\scriptsize iD}}}}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{float}
        \floatplacement{table}{htbp}
        \floatplacement{figure}{htbp}
        \usepackage{placeins}
        \usepackage{setspace}
        \doublespacing
        \usepackage[mathlines]{lineno}
        \linenumbers
        \usepackage{textcomp}
        \usepackage{newunicodechar}
        \newunicodechar{≤}{\ensuremath{\leq}}
        \newunicodechar{≥}{\ensuremath{\geq}}
        \newunicodechar{×}{\ensuremath{\times}}
        \newunicodechar{β}{\ensuremath{\beta}}
        \newunicodechar{η}{\ensuremath{\eta}}
        \newunicodechar{χ}{\ensuremath{\chi}}
        \newunicodechar{σ}{\ensuremath{\sigma}}
        \newunicodechar{τ}{\ensuremath{\tau}}
        \newunicodechar{γ}{\ensuremath{\gamma}}
        \newunicodechar{ζ}{\ensuremath{\zeta}}
        \newunicodechar{φ}{\ensuremath{\varphi}}
        \newunicodechar{μ}{\ensuremath{\mu}}
        \newunicodechar{ν}{\ensuremath{\nu}}
        \newunicodechar{λ}{\ensuremath{\lambda}}
        \newunicodechar{Δ}{\ensuremath{\Delta}}
        \newunicodechar{ε}{\ensuremath{\varepsilon}}
        \newunicodechar{ρ}{\ensuremath{\rho}}
        \newunicodechar{→}{\ensuremath{\rightarrow}}
        \newunicodechar{↔}{\ensuremath{\leftrightarrow}}
        \newunicodechar{≈}{\ensuremath{\approx}}
        \newunicodechar{²}{\textsuperscript{2}}
        \setlength{\LTleft}{0pt}
        \setlength{\LTright}{\fill}
knitr:
  opts_chunk:
    dev: cairo_pdf
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
csl: elsevier-vancouver.csl
---

```{r setup}
#| include: false

# Find project root and change to it so here() works
orig_wd <- getwd()
project_root <- normalizePath(
  file.path(orig_wd, "..")
)
if (!file.exists(
  file.path(project_root, ".here")
)) {
  project_root <- orig_wd
}
setwd(project_root)

library(here)
library(data.table)
library(gt)
library(ggplot2)
library(knitr)
library(patchwork)

source(here("R/utils/config.R"))
source(here("R/utils/data_io.R"))
source(here("R/utils/tables.R"))
source(here("R/utils/plotting.R"))
source(here("R/utils/consort_diagram.R"))

config <- load_config()

# -------------------------------------------------
# Load pre-computed manuscript environment
# -------------------------------------------------
env.lst <- readRDS(
  here("outputs/manuscript_env.rds")
)
list2env(env.lst, envir = environment())
rm(env.lst)

# -------------------------------------------------------------------
# gt table rendering for LaTeX/PDF output
# -------------------------------------------------------------------

knit_print.gt_tbl <- function(x, ...) {
  if (knitr::is_latex_output()) {
    latex_code <- as.character(gt::as_latex(x))

    # Fix 1: Table width
    latex_code <- gsub(
      "\\\\begin\\{tabular\\*\\}\\{\\\\linewidth\\}\\{@\\{\\\\extracolsep\\{\\\\fill\\}\\}",
      "\\\\begin{tabular}{",
      latex_code
    )
    latex_code <- gsub(
      "\\\\end\\{tabular\\*\\}", "\\\\end{tabular}",
      latex_code
    )

    # Fix 2: HTML tags to LaTeX
    latex_code <- gsub(
      "<em>([^<]+)</em>", "\\\\textit{\\1}", latex_code
    )
    latex_code <- gsub(
      "<strong>([^<]+)</strong>", "\\\\textbf{\\1}", latex_code
    )
    latex_code <- gsub(
      "<sub>([^<]+)</sub>", "\\\\textsubscript{\\1}", latex_code
    )
    latex_code <- gsub(
      "<sup>([^<]+)</sup>", "\\\\textsuperscript{\\1}", latex_code
    )

    # Fix 3: Unicode symbols to LaTeX math
    latex_code <- gsub("Δ", "$\\\\Delta$", latex_code)
    latex_code <- gsub("β", "$\\\\beta$", latex_code)
    latex_code <- gsub("λ", "$\\\\lambda$", latex_code)
    latex_code <- gsub("σ", "$\\\\sigma$", latex_code)
    latex_code <- gsub("η", "$\\\\eta$", latex_code)
    latex_code <- gsub("χ", "$\\\\chi$", latex_code)
    latex_code <- gsub("τ", "$\\\\tau$", latex_code)
    latex_code <- gsub("γ", "$\\\\gamma$", latex_code)
    latex_code <- gsub(
      "²", "\\\\textsuperscript{2}", latex_code
    )
    latex_code <- gsub("\u2264", "$\\\\leq$", latex_code)
    latex_code <- gsub("\u2265", "$\\\\geq$", latex_code)
    latex_code <- gsub("\u2260", "$\\\\neq$", latex_code)
    latex_code <- gsub("×", "$\\\\times$", latex_code)

    # Fix 4: Unicode text commands missing {} terminator
    latex_code <- gsub(
      "\\\\(texttimes|texttwosuperior|texttwosuperiorp|textendash|textemdash)([A-Za-z])",
      "\\\\\\1{}\\2", latex_code
    )

    # Fix 5: Footnote minipage width
    latex_code <- gsub(
      "\\\\begin\\{minipage\\}\\{\\\\linewidth\\}",
      "\\\\begin{minipage}[t]{0.8\\\\textwidth}\\\\raggedright\\\\small",
      latex_code
    )

    # Fix 6: Column alignment (first left, rest right)
    fix_alignment <- function(code) {
      pattern <- "(\\\\begin\\{(?:tabular|longtable)\\}\\{)([lcr]+)(\\})"
      m <- gregexpr(pattern, code, perl = TRUE)
      if (m[[1]][1] == -1) return(code)
      matches <- regmatches(code, m)[[1]]
      for (match in matches) {
        col_spec <- gsub(".*\\{([lcr]+)\\}$", "\\1", match)
        n_cols <- nchar(col_spec)
        if (n_cols > 1) {
          new_spec <- paste0(
            "l", paste(rep("r", n_cols - 1), collapse = "")
          )
          new_match <- gsub(
            "\\{[lcr]+\\}$", paste0("{", new_spec, "}"), match
          )
          code <- sub(match, new_match, code, fixed = TRUE)
        }
      }
      code
    }
    latex_code <- fix_alignment(latex_code)

    # Fix 7: Table spacing
    latex_code <- gsub(
      "\\\\begin\\{tabular\\}",
      "\\\\renewcommand{\\\\arraystretch}{1.15}\\\\begin{tabular}",
      latex_code
    )

    # Fix 8: Escape [ ] in table cells
    latex_code <- gsub(
      "\\\\addlinespace\\[([^]]+)\\]",
      "<<ADDLNSP_\\1_>>", latex_code
    )
    latex_code <- gsub(
      "\\\\\\\\\\[([0-9.]+pt)\\]",
      "<<LINEBRK_\\1_>>", latex_code
    )
    latex_code <- gsub(
      "\\\\begin\\{minipage\\}\\[([a-z])\\]",
      "<<MINIPAGE_\\1_>>", latex_code
    )
    latex_code <- gsub("\\[", "{[}", latex_code)
    latex_code <- gsub("\\]", "{]}", latex_code)
    latex_code <- gsub(
      "<<ADDLNSP_([^_]+)_>>",
      "\\\\addlinespace[\\1]", latex_code
    )
    latex_code <- gsub(
      "<<LINEBRK_([^_]+)_>>",
      "\\\\\\\\[\\1]", latex_code
    )
    latex_code <- gsub(
      "<<MINIPAGE_([a-z])_>>",
      "\\\\begin{minipage}[\\1]", latex_code
    )

    knitr::asis_output(latex_code)
  } else {
    gt:::knit_print.gt_tbl(x, ...)
  }
}
registerS3method(
  "knit_print", "gt_tbl", knit_print.gt_tbl,
  envir = asNamespace("knitr")
)

# Domain labels for cognitive measures
DOMAIN_LABELS <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

# Helper: Format-aware markdown (handles LaTeX vs HTML)
fmt_crossformat <- function(gt_tbl, columns, rows = everything()) {
  if (knitr::is_latex_output()) {
    gt::fmt(
      gt_tbl, columns = {{ columns }},
      rows = {{ rows }}, fns = function(x) {
        x <- gsub("\\*([^*]+)\\*", "\\\\textit{\\1}", x)
        x <- gsub("~([^~]+)~", "\\\\textsubscript{\\1}", x)
        x
      }
    )
  } else {
    gt::fmt_markdown(
      gt_tbl, columns = {{ columns }}, rows = {{ rows }}
    )
  }
}
```

:::{=latex}
\appendix
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}
:::

# Appendix A: Normative Transfer Validation


For transfer validation, we identified a comparable
ADNI subsample of `r adni_demo$n_total`
amyloid-negative cognitively unimpaired participants
(`r sprintf("%.0f%%", adni_demo$pct_female)` female)
within the UKB age range. Despite matching on amyloid
status and cognitive normality, the ADNI comparable
subsample was approximately
`r sprintf("%.0f", adni_demo$age_mean - ukb_demo$age_mean)`
years older on average (mean
`r sprintf("%.1f", adni_demo$age_mean)` vs
`r sprintf("%.1f", ukb_demo$age_mean)` years;
@fig-normative-validation A), concentrating in the
upper tail of the UKB age distribution. HVR z-scores
showed a systematic negative offset (mean =
`r sprintf("%.2f", comp_hvr$mean)`, SD =
`r sprintf("%.2f", comp_hvr$sd)`, 95% CI
[`r sprintf("%.2f", comp_hvr$ci_95["lower"])`,
`r sprintf("%.2f", comp_hvr$ci_95["upper"])`]),
while HC z-scores were closer to expected values
(mean = `r sprintf("%.2f", comp_hc$mean)`, SD =
`r sprintf("%.2f", comp_hc$sd)`;
@fig-normative-validation B). Age-bin-specific
calibration (@fig-normative-validation, panel B)
showed mean z-scores near zero in younger bins with
increasingly negative offset in older bins,
consistent with the age mismatch.

```{r fig-normative-validation}
#| fig-cap: "**Normative transfer validation.** (A) Age distributions: UK Biobank (UKB, yellow) vs ADNI comparable subsample (cyan); dashed lines indicate sample means. (B) Forest plot of HVR (purple) and HC (green) z-scores (mean ± 95% CI) by age bin; reference line at z = 0 indicates perfect calibration; open circles = age bins, filled circles = overall. (C) Z-score distributions for HVR (left) and HC (right) compared to the expected N(0,1) (yellow)."
#| fig-width: 6.5
#| fig-height: 5.5

stopifnot(!is.null(nv_data))
ukb_d <- nv_data$comparable_subsample$demographics$ukb_normative
adni_d <- nv_data$comparable_subsample$demographics$adni_comparable

p_age <- plot_age_distribution_comparison(
  ukb_age_mean = ukb_d$age_mean,
  ukb_age_sd = ukb_d$age_sd,
  ukb_age_min = ukb_d$age_min,
  ukb_age_max = ukb_d$age_max,
  ukb_n = ukb_d$n_total,
  adni_age_mean = adni_d$age_mean,
  adni_age_sd = adni_d$age_sd,
  adni_n = adni_d$n_total
)

p_forest <- plot_zscore_validation_forest(nv_data)

# Build z-score data.table with both measures
hvr_z.v <- comp_hvr$zscores
hc_z.v <- comp_hc$zscores
stopifnot(!is.null(hvr_z.v), !is.null(hc_z.v))
n_max <- max(length(hvr_z.v), length(hc_z.v))
zscore.dt <- data.table(
  PTID = seq_len(n_max),
  HVR_z = hvr_z.v[seq_len(n_max)],
  HC_z = hc_z.v[seq_len(n_max)]
)

p_dist <- plot_zscore_distribution(
  zscore.dt, sample_label = "Comparable"
)

(p_age | p_forest) / p_dist +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(tag_levels = "A")
```


:::{=latex}
\FloatBarrier
\newpage
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{B\arabic{table}}
\renewcommand{\thefigure}{B\arabic{figure}}
:::

# Appendix B: Methodological Validations

## CVR~mimic~ Measurement Invariance

Measurement invariance of the CVR~mimic~ model was
tested across sex in four steps (configural, metric,
scalar, strict; @tbl-mi).

```{r tbl-mi}
#| tbl-cap: "**CVR~mimic~ measurement invariance.** Configural = same structure; Metric = equal loadings; Scalar = equal intercepts; Strict = equal residuals. ΔCFI < 0.01 and ΔRMSEA < 0.015 indicate acceptable invariance."

stopifnot(!is.null(cvr_mi.res))
require_output(
  create_mi_table(
    cvr_mi.res,
    "Measurement Invariance Across Sex"
  ),
  "tbl-mi"
)
```

## EDT Linearity and Simulation Validation

The linear time specification was validated via LME
likelihood ratio tests comparing linear vs quadratic
EDT across all sex-domain models (see
@tbl-linearity).

```{r tbl-linearity}
#| tbl-cap: "**EDT linearity validation.** LME models testing linear vs quadratic EDT specification. LRT tests whether adding a quadratic EDT term significantly improves fit. ΔAIC = AIC(linear) − AIC(quadratic); positive values favor quadratic. EDT = Estimated Disease Time."

require_output(
  create_linearity_table(
    edt_linearity,
    title = "EDT Linearity Validation"
  ),
  "tbl-linearity"
)
```

```{r edt-linearity-narrative}
#| include: false

stopifnot(!is.null(edt_linearity))
lt <- as.data.table(
  edt_linearity$results_table
)
n_sig <- sum(
  lt$LRT_p < 0.05, na.rm = TRUE
)
n_total <- nrow(lt)
if (n_sig == 0) {
  edt_lin_text <- paste0(
    "No domain showed a significant",
    " quadratic EDT term",
    " (all p > 0.05), supporting the",
    " linear time specification for",
    " the LGCM."
  )
} else {
  sig_doms <- lt[LRT_p < 0.05, Domain]
  max_chisq <- max(
    lt$LRT_chisq, na.rm = TRUE
  )
  edt_lin_text <- sprintf(
    paste0(
      "%s models showed significant",
      " quadratic EDT terms",
      " (%s; max χ² = %.1f).",
      " Large test statistics reflect",
      " substantial power afforded by",
      " sample size and repeated",
      " observations. A fixed-quadratic",
      " specification was adopted:",
      " quadratic variance fixed to",
      " near zero with freely estimated",
      " quadratic means, capturing",
      " population-level curvature",
      " without individual-level",
      " quadratic variation."
    ),
    sub("^(.)", "\\U\\1",
      n_of_total.fn(n_sig, n_total),
      perl = TRUE
    ),
    paste(
      DOMAIN_LABELS[sig_doms],
      collapse = ", "
    ),
    max_chisq
  )
}
```

`r edt_lin_text`

```{r fig-edt-comparison}
#| fig-cap: "**Cognitive trajectories under calendar time versus estimated disease time.** Composite scores (Memory, Language, Executive Function) plotted against years from baseline (A) and estimated disease time (B). Under calendar time, the population-average trajectory appears flat because participants at different disease stages are pooled at each follow-up; under EDT, disease-stage alignment reveals the expected monotonic cognitive decline. Lines show loess smooths with 95% CI bands; individual trajectories in light gray. Y-axes are shared across panels for direct comparison; x-axes differ because the two metrics have different natural scales."
#| fig-width: 7
#| fig-height: 5.5

# Reshape to long format for faceting
cog_long.dt <- melt(
  cohort.dt[
    , .(PTID, YRS_from_bl, EDT,
        MEM, LAN, EXF)
  ],
  id.vars = c("PTID", "YRS_from_bl", "EDT"),
  variable.name = "Domain",
  value.name = "Score"
)
cog_long.dt <- cog_long.dt[!is.na(Score)]

domain_labs <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)
cog_long.dt[, Domain := factor(
  domain_labs[as.character(Domain)],
  levels = domain_labs
)]

# Shared y-axis limits across all panels
y_range <- range(
  cog_long.dt$Score, na.rm = TRUE
)
y_lims <- c(
  floor(y_range[1]), ceiling(y_range[2])
)

# Common plot elements
smooth_col <- "gray20"
spaghetti_alpha <- 0.05

# Calendar time (YRS) panel
p_yrs <- ggplot(
  cog_long.dt,
  aes(x = YRS_from_bl, y = Score)
) +
  geom_line(
    aes(group = PTID),
    alpha = spaghetti_alpha,
    linewidth = 0.3, color = "gray60"
  ) +
  geom_smooth(
    method = "loess", formula = y ~ x,
    color = smooth_col, fill = "gray60",
    alpha = 0.25, linewidth = 0.9
  ) +
  facet_wrap(~ Domain) +
  coord_cartesian(ylim = y_lims) +
  labs(
    x = "Years from baseline",
    y = "Composite score",
    title = "A. Calendar time (YRS)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(
      size = 9, face = "bold"
    ),
    plot.title = element_text(
      size = 10, face = "bold"
    ),
    panel.grid.minor = element_blank()
  )

# Disease time (EDT) panel
p_edt <- ggplot(
  cog_long.dt[!is.na(EDT)],
  aes(x = EDT, y = Score)
) +
  geom_line(
    aes(group = PTID),
    alpha = spaghetti_alpha,
    linewidth = 0.3, color = "gray60"
  ) +
  geom_smooth(
    method = "loess", formula = y ~ x,
    color = smooth_col, fill = "gray60",
    alpha = 0.25, linewidth = 0.9
  ) +
  facet_wrap(~ Domain) +
  coord_cartesian(ylim = y_lims) +
  labs(
    x = "Estimated disease time (years)",
    y = "Composite score",
    title = "B. Disease time (EDT)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(
      size = 9, face = "bold"
    ),
    plot.title = element_text(
      size = 10, face = "bold"
    ),
    panel.grid.minor = element_blank()
  )

p_yrs / p_edt
```

## Standard Error-Constrained LGCM Validation

A Monte Carlo simulation
(`r simulation.res$config$n_reps` replications per
condition at
N = `r paste(simulation.res$config$sample_sizes, collapse = " and N = ")`)
compared SE-constrained LGCM (residual variance
fixed to SE²) against free estimation. Parameters
were calibrated from the observed ADNI LGCM results
(see Methods). To assess robustness, the HVR effect
and slope variance were varied at 0.5×, 1×, and
1.5× their observed values, yielding 9 conditions.

```{r simulation-interpretation}
#| include: false

stopifnot(!is.null(sim_conv))
stopifnot(!is.null(sim_conv_full))

n_reps <- simulation.res$config$n_reps

# Base condition (1x/1x) summary
base_n200 <- sim_conv[n == 200]
base_n500 <- sim_conv[n == 500]

# Grid-wide summary
cov_range_se <- range(
  sim_conv_full$coverage_se * 100
)
cov_range_free <- range(
  sim_conv_full$coverage_free * 100
)
max_bias <- max(
  abs(sim_conv_full$bias_se),
  abs(sim_conv_full$bias_free)
)
```

At the base condition (observed ADNI parameters),
the SE-constrained approach converged in
`r base_n200$n_converged_se` of `r n_reps`
replications at N = 200 (free: `r base_n200$n_converged_free` of `r n_reps`)
with bias =
`r sprintf("%.1e", base_n200$bias_se)`
(free: `r sprintf("%.1e", base_n200$bias_free)`),
RMSE = `r sprintf("%.4f", base_n200$rmse_se)`
(free: `r sprintf("%.4f", base_n200$rmse_free)`),
and 95% CI coverage =
`r sprintf("%.1f", base_n200$coverage_se * 100)`%
(free:
`r sprintf("%.1f", base_n200$coverage_free * 100)`%).
At N = 500, bias was
`r sprintf("%.1e", base_n500$bias_se)`
(free: `r sprintf("%.1e", base_n500$bias_free)`),
RMSE = `r sprintf("%.4f", base_n500$rmse_se)`
(free: `r sprintf("%.4f", base_n500$rmse_free)`),
and coverage =
`r sprintf("%.1f", base_n500$coverage_se * 100)`%
(free:
`r sprintf("%.1f", base_n500$coverage_free * 100)`%).

Across all 9 grid conditions, maximum absolute bias
was `r sprintf("%.1e", max_bias)`, coverage ranged
from `r sprintf("%.1f", cov_range_se[1])`% to
`r sprintf("%.1f", cov_range_se[2])`% for
SE-constrained and
`r sprintf("%.1f", cov_range_free[1])`% to
`r sprintf("%.1f", cov_range_free[2])`% for free
estimation. Both approaches showed comparable
performance across all realistic parameter
combinations, validating the SE-constrained
specification used in the primary analyses. The
equivalence between methods under ADNI-calibrated
conditions confirms that fixing residual variance to
SE² neither introduces bias nor distorts inference,
while incorporating known measurement information
from the ADSP-PHC scoring model.

## LME Covariate Effects

Full covariate effects from the sex-stratified
LME 3-way interaction models are presented for
the FRS model (@tbl-lme-covariates) and the
CVR~mimic~ model (@tbl-cvr-lme-covariates).

```{r tbl-lme-covariates}
#| tbl-cap: "**FRS LME covariate effects.** Sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    frs_hvr, cvr_label = "FRS",
    title = "Covariate Effects (FRS Model)"
  ),
  "tbl-lme-covariates"
)
```

```{r tbl-cvr-lme-covariates}
#| tbl-cap: "**CVR~mimic~ LME covariate effects.** Sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    cvr_hvr,
    cvr_label = "CVR~mimic~",
    title = md(
      "Covariate Effects (CVR~mimic~ Model)"
    )
  ),
  "tbl-cvr-lme-covariates"
)
```


:::{=latex}
\FloatBarrier
\newpage
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{C\arabic{table}}
\renewcommand{\thefigure}{C\arabic{figure}}
:::

# Appendix C: Sensitivity Analyses

All sensitivity checks converge on the same
pattern: FRS effects trace to age-weighting,
CVR~mimic~ effects are null or isolated, and the
HVR–cognition coupling (b-path) is consistent. Main
findings are stable across alternative brain
measures (HC volume), raw scores, attrition
adjustments, EDT specification checks, and an
amyloid-negative CU control analysis. Full details
follow.

## LME Sensitivity

### HC z-score (Total Hippocampal Volume)

HC z-score effects were generally weaker than HVR
z-score effects, with fewer significant 3-way
interactions, supporting the
hippocampus-to-ventricle ratio as a more sensitive
brain aging biomarker. The LME 3-way interaction was
repeated with HC z-score instead of HVR; results
appear in @tbl-hc-comparison, with covariate effects
in @tbl-hc-covariates and @tbl-hc-covariates-cvr.

```{r tbl-hc-comparison}
#| tbl-cap: "**HC z-score sensitivity analysis.** LME 3-way interaction effects (Years $\\times$ cardiovascular risk $\\times$ HC z-score) for total hippocampal volume. FRS and CVR$_{\\text{mimic}}$ compared side-by-side. Bold indicates p < 0.05."

require_output(
  create_lme_comparison_table(
    frs_hc, cvr_hc,
    summary_dt = hc_summary.dt
  ),
  "tbl-hc-comparison"
)
```

```{r tbl-hc-covariates}
#| tbl-cap: "**FRS HC z-score covariate effects.** Sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    frs_hc, cvr_label = "FRS",
    title = "HC Covariate Effects (FRS Model)"
  ),
  "tbl-hc-covariates"
)
```

```{r tbl-hc-covariates-cvr}
#| tbl-cap: "**CVR~mimic~ HC z-score covariate effects.** Sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    cvr_hc, cvr_label = "CVR~mimic~",
    title = md(
      "HC Covariate Effects (CVR~mimic~ Model)"
    )
  ),
  "tbl-hc-covariates-cvr"
)
```

```{r hc-interpretation}
#| include: false

stopifnot(!is.null(hc_summary.dt))
hc_p <- hc_summary.dt[
  Analysis == "Primary_Stratified"
]
frs_hc_sig <- hc_p[
  CVR_Measure == "FRS" & Significant == TRUE
]
cvr_hc_sig <- hc_p[
  CVR_Measure == "CVR_mimic" &
    Significant == TRUE
]
frs_hc_desc <- if (nrow(frs_hc_sig) > 0) {
  paste(
    sprintf(
      "%s %s",
      frs_hc_sig$Sex,
      DOMAIN_LABELS[frs_hc_sig$Domain]
    ),
    collapse = " and "
  )
} else "no sex–domain combination"
cvr_hc_desc <- if (nrow(cvr_hc_sig) > 0) {
  paste(
    sprintf(
      "%s %s",
      cvr_hc_sig$Sex,
      DOMAIN_LABELS[cvr_hc_sig$Domain]
    ),
    collapse = " and "
  )
} else "any sex–domain combination"
hc_interp_text <- sprintf(
  paste0(
    "The FRS × HC × Years interaction was",
    " significant for %s, whereas",
    " CVR~mimic~ × HC × Years %s"
  ),
  frs_hc_desc,
  if (nrow(cvr_hc_sig) > 0) {
    sprintf(
      "reached significance only for %s",
      cvr_hc_desc
    )
  } else {
    paste0(
      "did not reach significance for ",
      cvr_hc_desc
    )
  }
)
```

`r hc_interp_text`. HC z-score showed weaker effects than HVR z-score, supporting use of the ratio measure.

```{r hc-vs-hvr-synthesis}
#| include: false

hc_p2 <- hc_summary.dt[
  Analysis == "Primary_Stratified"
]
hvr_p <- lme_summary.dt[
  Analysis == "Primary_Stratified"
]
n_hc_sig <- sum(
  hc_p2$Significant, na.rm = TRUE
)
n_hvr_sig <- sum(
  hvr_p$Significant, na.rm = TRUE
)
hc_synth_text <- sprintf(
  paste0(
    "Significant 3-way interactions emerged",
    " in %s models for HC z-score versus",
    " %s for HVR z-score. The weaker HC",
    " effects support the specificity of",
    " HVR: capturing ventricular expansion",
    " alongside hippocampal atrophy",
    " provides greater sensitivity to",
    " cardiovascular risk moderation."
  ),
  n_of_total.fn(n_hc_sig, 12L),
  n_of_total.fn(n_hvr_sig, 12L)
)
```

`r hc_synth_text`

### Attrition Analysis

Baseline characteristics were compared between
completers (≥4 visits) and dropouts (<4 visits);
results appear in @tbl-attrition-baseline.

```{r tbl-attrition-baseline}
#| tbl-cap: "**Attrition analysis.** Baseline characteristics of completers (≥4 visits) vs dropouts (<4 visits). Bold rows indicate p < 0.05. Completers = participants with sufficient visits for quadratic trend estimation."

require_output(
  create_attrition_table(attrition_analysis),
  "tbl-attrition-baseline"
)
```

```{r attrition-interpretation}
#| include: false

stopifnot(
  !is.null(attrition_analysis),
  !is.null(attrition_analysis$comparison_table)
)
comp_dt <- attrition_analysis$comparison_table
sig_vars <- comp_dt[p_value < 0.05, Variable]
var_display <- c(
  AGE = "age", EDUC = "education",
  FRS = "FRS", HVR_z = "HVR z-score",
  HC_z = "HC z-score", SEX = "sex",
  APOE4 = "APOE ε4 carrier status"
)
sig_labels <- vapply(
  sig_vars, function(v) {
    if (v %in% names(var_display)) {
      var_display[v]
    } else v
  }, character(1)
)
if (length(sig_vars) > 0) {
  attrition_text <- sprintf(
    paste0(
      "Significant baseline differences",
      " between completers and dropouts",
      " were observed in %s (all p < 0.05;",
      " see @tbl-attrition-baseline)."
    ),
    paste(sig_labels, collapse = ", ")
  )
} else {
  attrition_text <- paste0(
    "No significant baseline",
    " differences between completers",
    " and dropouts were observed",
    " (see @tbl-attrition-baseline),",
    " suggesting attrition was random",
    " with respect to measured variables."
  )
}
```

`r attrition_text`

```{r attrition-logistic-inline}
#| include: false

stopifnot(
  !is.null(attrition_analysis),
  !is.null(attrition_analysis$logistic)
)
log_model <- attrition_analysis$logistic
or_dt <- as.data.table(log_model$odds_ratios)
setnames(
  or_dt,
  c("2.5 %", "97.5 %"),
  c("CI_lower", "CI_upper"),
  skip_absent = TRUE
)
or_dt <- or_dt[term != "(Intercept)"]
or_dt[, sig := CI_lower > 1 | CI_upper < 1]

# Readable term labels
term_display <- c(
  FRS = "FRS", AGE = "age",
  APOE4TRUE = "APOE ε4 carrier status",
  SEXMale = "male sex",
  HVR_z = "HVR z-score",
  HC_z = "HC z-score"
)
or_dt[, term_label := fifelse(
  term %in% names(term_display),
  term_display[term], term
)]

sig_preds <- or_dt[sig == TRUE]
if (nrow(sig_preds) > 0) {
  items <- sprintf(
    "%s (OR = %.2f, 95%% CI [%.2f, %.2f])",
    sig_preds$term_label,
    sig_preds$OR,
    sig_preds$CI_lower,
    sig_preds$CI_upper
  )
  logistic_text <- sprintf(
    paste0(
      "Logistic regression predicting",
      " dropout identified %d significant",
      " predictor%s — %s."
    ),
    nrow(sig_preds),
    ifelse(nrow(sig_preds) > 1, "s", ""),
    paste(items, collapse = "; ")
  )
} else {
  logistic_text <- paste0(
    "Logistic regression predicting",
    " dropout identified no significant",
    " predictors (all 95% CIs included 1)."
  )
}

# FRS direction narrative
comp_dt <- attrition_analysis$comparison_table
frs_row <- comp_dt[
  grepl("FRS", Variable, ignore.case = TRUE)
]
frs_dir_text <- ""
if (nrow(frs_row) > 0) {
  c_mean <- frs_row$Completers_mean[1]
  d_mean <- frs_row$Dropouts_mean[1]
  if (!is.null(c_mean) && !is.null(d_mean) &&
      !is.na(c_mean) && !is.na(d_mean)) {
    direction <- if (d_mean > c_mean) {
      "higher"
    } else "lower"
    frs_dir_text <- sprintf(
      paste0(
        " Dropouts had %s baseline FRS",
        " (M = %.2f) than completers",
        " (M = %.2f), consistent with",
        " selective loss of higher-risk",
        " participants — this may",
        " attenuate observed cardiovascular",
    " risk effects."
      ),
      direction, d_mean, c_mean
    )
  }
}
```

`r logistic_text``r frs_dir_text`

### Raw Brain Measure Sensitivity

The LME 3-way interaction was repeated using raw
(un-normalized) HVR and raw HC (with total
intracranial volume [TIV] covariate),
with FRS as predictor, full-sample, no sex
stratification, and no Years² term.

```{r raw-hvr-inline}
#| include: false

stopifnot(
  !is.null(lme_raw_sens.res),
  !is.null(lme_raw_sens.res$results)
)
hvr_raw <- lme_raw_sens.res$results$HVR_raw_FRS
stopifnot(!is.null(hvr_raw))
raw_hvr_results <- list(results = hvr_raw)

hc_raw <- lme_raw_sens.res$results$HC_raw_FRS
stopifnot(!is.null(hc_raw))
raw_hc_results <- list(results = hc_raw)

fmt_raw_sig <- function(res) {
  sig.v <- c()
  nonsig.v <- c()
  for (d in c("MEM", "LAN", "EXF")) {
    r <- res$results[[d]]$interaction
    if (is.null(r)) next
    entry <- sprintf(
      "%s (β = %.4f, %s)",
      DOMAIN_LABELS[d], r$beta,
      fmt_p_inline(r$p_value)
    )
    if (r$p_value < 0.05) {
      sig.v <- c(sig.v, entry)
    } else {
      nonsig.v <- c(nonsig.v, entry)
    }
  }
  list(sig = sig.v, nonsig = nonsig.v)
}
hvr_raw_split <- fmt_raw_sig(raw_hvr_results)
hc_raw_split <- fmt_raw_sig(raw_hc_results)
raw_hvr_prose <- if (
  length(hvr_raw_split$sig) == 3
) {
  paste0(
    "Raw HVR was significant for all three",
    " domains: ",
    paste(hvr_raw_split$sig, collapse = ", ")
  )
} else if (length(hvr_raw_split$sig) > 0) {
  paste0(
    "Raw HVR was significant for ",
    paste(
      hvr_raw_split$sig, collapse = " and "
    ),
    if (length(hvr_raw_split$nonsig) > 0)
      paste0(
        ", but not ",
        paste(
          hvr_raw_split$nonsig,
          collapse = " or "
        )
      )
    else ""
  )
} else {
  "Raw HVR was not significant for any domain"
}
raw_hc_prose <- if (
  length(hc_raw_split$sig) > 0 &&
    length(hc_raw_split$nonsig) > 0
) {
  paste0(
    "Raw HC (+TIV) reached significance",
    " only for ",
    paste(
      hc_raw_split$sig, collapse = " and "
    )
  )
} else if (length(hc_raw_split$sig) == 0) {
  paste0(
    "Raw HC (+TIV) was not significant",
    " for any domain"
  )
} else {
  paste0(
    "Raw HC (+TIV) was significant for ",
    paste(hc_raw_split$sig, collapse = ", ")
  )
}
raw_meas_text <- paste0(
  raw_hvr_prose, ". ", raw_hc_prose
)

# Concordance check
sig_hvr <- sapply(
  names(raw_hvr_results$results),
  function(d) {
    r <- raw_hvr_results$results[[d]]$interaction
    !is.null(r) && r$p_value < 0.05
  }
)
if (any(sig_hvr)) {
  raw_concord_text <- sprintf(
    paste0(
      "Raw HVR × FRS was significant for",
      " %s, confirming normative z-scoring",
      " does not drive primary findings."
    ),
    paste(
      DOMAIN_LABELS[names(sig_hvr)[sig_hvr]],
      collapse = ", "
    )
  )
} else {
  raw_concord_text <- paste0(
    "Raw HVR × FRS was not significant",
    " for any domain, suggesting",
    " normative z-scoring enhances",
    " sensitivity."
  )
}

# Raw HC (+TIV) note
sig_hc <- sapply(
  names(raw_hc_results$results),
  function(d) {
    r <- raw_hc_results$results[[d]]$interaction
    !is.null(r) && r$p_value < 0.05
  }
)
raw_hc_note <- if (!any(sig_hc)) {
  paste0(
    " Raw HC (+TIV) was not significant",
    " for any domain, reinforcing the",
    " added sensitivity of the ratio",
    " measure."
  )
} else ""
```

`r raw_meas_text`. `r raw_concord_text``r raw_hc_note`

### EDT Moderation of LME Effects

A 4-way interaction (cardiovascular risk × HVR ×
EDT × Years) tested
whether cardiovascular risk effects vary by disease
stage.

```{r edt-moderation-inline}
#| include: false

build_edt_text <- function(edt_res, cvr_label) {
  stopifnot(
    !is.null(edt_res),
    !is.null(edt_res$summary_continuous)
  )
  sc <- as.data.table(edt_res$summary_continuous)
  parts <- vapply(
    c("Male", "Female"), function(sx) {
      sx_rows <- sc[Sex == sx]
      sig <- sx_rows[Significant == TRUE]
      if (nrow(sig) > 0) {
        items <- sprintf(
          "%s (β = %.4f, %s)",
          DOMAIN_LABELS[sig$Domain],
          sig$Beta,
          sapply(sig$p_value, fmt_p_inline)
        )
        sprintf(
          paste0(
            "%s × HVR × EDT (%s):",
            " significant for %s"
          ),
          cvr_label, sx,
          paste(items, collapse = "; ")
        )
      } else {
        min_p <- min(
          sx_rows$p_value, na.rm = TRUE
        )
        sprintf(
          paste0(
            "%s × HVR × EDT (%s):",
            " not significant (min %s)"
          ),
          cvr_label, sx,
          fmt_p_inline(min_p)
        )
      }
    }, character(1)
  )
  paste(parts, collapse = ". ")
}
edt_mod_text <- paste(
  build_edt_text(frs_edt, "FRS"),
  build_edt_text(cvr_edt, "CVR~mimic~"),
  sep = ". "
)
```

`r edt_mod_text`. Limited EDT moderation suggests cardiovascular risk effects on HVR-cognition coupling are relatively stable across disease stages.

### EDT Stage Sensitivity

Analyses were restricted to early and middle EDT
tertiles (excluding late-stage) to confirm effect
stability.

```{r edt-sensitivity-inline}
#| include: false

stopifnot(!is.null(edt_sens.res))
rt <- as.data.table(edt_sens.res$results_table)

meas_labels <- c(
  HVR_z = "HVR z-score",
  HC_z = "HC z-score"
)
dom_labels <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

edt_sens_parts <- vapply(
  unique(rt$Measure), function(meas) {
    mrt <- rt[Measure == meas]
    m_lab <- if (
      meas %in% names(meas_labels)
    ) meas_labels[meas] else meas
    items <- c()
    n_concord <- 0L
    for (i in seq_len(nrow(mrt))) {
      row <- mrt[i]
      d_lab <- if (
        row$Domain %in% names(dom_labels)
      ) dom_labels[row$Domain] else row$Domain
      pct <- if (
        !is.null(row$Beta_change_pct) &&
        !is.na(row$Beta_change_pct)
      ) {
        sprintf(
          ", %.1f%% change",
          row$Beta_change_pct
        )
      } else ""
      if (!is.null(row$Consistent) &&
          isTRUE(row$Consistent)) {
        n_concord <- n_concord + 1L
      }
      items <- c(items, sprintf(
        "%s (β = %.4f, %s%s)",
        d_lab, row$Beta_early_middle,
        fmt_p_inline(row$P_early_middle),
        pct
      ))
    }
    concord_note <- if (n_concord > 0) {
      sprintf(
        " (%d/%d concordant with full-sample)",
        n_concord, nrow(mrt)
      )
    } else ""
    sprintf(
      "%s: %s%s", m_lab,
      paste(items, collapse = "; "),
      concord_note
    )
  }, character(1)
)
edt_sens_text <- paste(
  edt_sens_parts, collapse = ". "
)
```

`r edt_sens_text`.

## Amyloid-Negative CU Control

### LME 3-Way Interaction

```{r aneg-lme-inline}
#| include: false

stopifnot(!is.null(aneg_lme_summary.lst))
al <- aneg_lme_summary.lst
```

The LME 3-way interaction was repeated in
`r al$n` A^--^/CU participants (pooled sex, SEX
as covariate; see Methods). Results appear in
@tbl-aneg-lme.

```{r tbl-aneg-lme}
#| tbl-cap: "**A$^{-}$/CU LME 3-way interaction.** FRS vs CVR$_{\\text{mimic}}$ effects (YRS $\\times$ CVR $\\times$ HVR z-score $\\rightarrow$ Cognition) in amyloid-negative, cognitively unimpaired participants. Pooled sex with SEX as covariate. Bold indicates p < 0.05. FDR correction applied within each predictor (3 tests)."

require_output(
  create_aneg_lme_table(
    aneg_lme_summary.lst
  ),
  "tbl-aneg-lme"
)
```

FRS did not survive FDR correction in any domain
(`r n_of_total.fn(al$frs_fdr, 3L)`;
`r paste(al$frs_details, collapse = "; ")`),
whereas CVR~mimic~ reached significance in
`r n_of_total.fn(al$cvr_fdr, 3L)` domains after
FDR correction
(`r paste(al$cvr_details, collapse = "; ")`).
This complementary reversal — FRS significant in
A^+^ but not after FDR in A^--^/CU; CVR~mimic~ not
significant after FDR in A^+^ but significant in
A^--^/CU — supports the interpretation that FRS
effects are carried by the age component of the
score, which co-varies with AD pathology
(see main text, Figure 2).

### LGCM Mediation

```{r aneg-lgcm-inline}
#| include: false

stopifnot(!is.null(aneg_lgcm_summary.lst))
s <- aneg_lgcm_summary.lst

# b-path per-domain formatted strings
aneg_b_sig_parts <- sapply(
  s$b_by_dom[sapply(
    s$b_by_dom, function(x) x$sig
  )],
  function(x) {
    sprintf(
      "%s (β ≈ %.3f, %s to %s)",
      x$domain, x$b_avg,
      fmt_p_inline(x$p_min),
      fmt_p_inline(x$p_max)
    )
  }
)
aneg_b_sig_text <- paste(
  aneg_b_sig_parts, collapse = " and "
)
aneg_b_border_parts <- sapply(
  s$b_by_dom[sapply(
    s$b_by_dom, function(x) x$borderline
  )],
  function(x) {
    sprintf(
      "%s (%s to %s)",
      x$domain,
      fmt_p_inline(x$p_min),
      fmt_p_inline(x$p_max)
    )
  }
)
aneg_b_border_text <- paste(
  aneg_b_border_parts, collapse = " and "
)
```

The A^--^/CU LGCM used a linear growth model with
YRS time metric, pooled sexes, free residuals, and
Wald/Sobel inference (see Methods;
N = `r s$n_sample`,
analytic N = `r s$n_analytic`).

```{r tbl-aneg-lgcm}
#| tbl-cap: "**A$^{-}$/CU LGCM mediation paths.** Linear LGCM mediation in amyloid-negative, cognitively unimpaired participants. a-path = predictor $\\rightarrow$ HVR slope; b-path = HVR slope $\\rightarrow$ cognitive slope; indirect = a $\\times$ b. Bold indicates p < 0.05. Inference via Wald/Sobel tests (no bootstrap)."

require_output(
  create_aneg_lgcm_table(
    aneg_lgcm_summary.lst
  ),
  "tbl-aneg-lgcm"
)
```

Neither FRS nor CVR~mimic~ predicted HVR slope in any
of the `r s$a_total` models (all a-path
`r fmt_p_inline(s$a_min_p)`; @tbl-aneg-lgcm).
This contrasts sharply with the primary A^+^ analysis,
where FRS predicted hippocampal decline in all six
models (see main text, Figure 4): removing amyloid pathology
eliminates the FRS a-path, confirming that
FRS-hippocampal associations require AD-related
neurodegeneration rather than reflecting direct
cardiovascular effects on the hippocampus. The
a-path estimates themselves are near zero
(range `r sprintf("%.3f", s$a_range[1])`–`r sprintf("%.3f", s$a_range[2])`),
indicating this is a genuinely null effect rather
than a power limitation.

The b-path (HVR slope → cognitive slope) was
significant for `r aneg_b_sig_text``r if (nchar(aneg_b_border_text) > 0) paste0(", and borderline for ", aneg_b_border_text) else ""`,
confirming that hippocampal-cognitive coupling
persists in the absence of amyloid burden. This
invariance across both A^+^ and A^--^/CU samples
and across both cardiovascular risk predictors supports HVR decline
as a general marker of neurodegeneration-cognition
coupling, independent of AD pathology or
cardiovascular risk
measurement.

No indirect effects reached significance (all Sobel
*p* > 0.5), consistent with null a-paths. The
overall pattern — null a-paths with preserved
b-paths for both FRS and CVR~mimic~ — closes the
age-confounding argument: FRS mediates through
hippocampal decline only when age co-varies with
AD-driven atrophy (A^+^), not in the general aging
population (A^--^/CU).

:::{=latex}
\FloatBarrier
:::
