---
title: "Cardiovascular risk and hippocampal-cognitive coupling in Alzheimer's disease"
author:
  - name: Sofia Fernandez-Lozano
    orcid: 0000-0003-2288-8111
    affiliations:
      - ref: bic
      - ref: ipn
  - name: Sylvia Villeneuve
    orcid: 0000-0003-2338-0467
    affiliations:
      - ref: bic
      - ref: psych
      - ref: stopad
  - name: D. Louis Collins
    orcid: 0000-0002-8432-7021
    affiliations:
      - ref: bic
      - ref: ipn
      - ref: bbme
    email: louis.collins@mcgill.ca
    corresponding: true
  - name: "for the Alzheimer's Disease Neuroimaging Initiative*"
    affiliations: []
affiliations:
  - id: bic
    name: McConnell Brain Imaging Centre
    department: Montreal Neurological Institute
    city: Montreal
    state: Quebec
    country: Canada
  - id: ipn
    name: McGill University
    department: Department of Neurology and Neurosurgery
    city: Montreal
    state: Quebec
    country: Canada
  - id: psych
    name: McGill University
    department: Department of Psychiatry
    city: Montreal
    state: Quebec
    country: Canada
  - id: stopad
    name: Douglas Mental Health Institute Research Centre
    department: StoP-AD Centre
    city: Montreal
    state: Quebec
    country: Canada
  - id: bbme
    name: McGill University
    department: Department of Biomedical Engineering
    city: Montreal
    state: Quebec
    country: Canada
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
  pdf:
    keep-tex: true
    toc: false
    number-sections: true
    documentclass: article
    pdf-engine: xelatex
    geometry:
      - margin=1in
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage{tikz}
        \usetikzlibrary{shapes.geometric, arrows.meta, positioning}
        \usepackage{authblk}
        \renewcommand\Authfont{\normalsize}
        \renewcommand\Affilfont{\small\itshape}
        \renewcommand\Authands{, }
        \renewcommand\Authand{, }
        \providecommand{\orcidlink}[1]{\href{https://orcid.org/#1}{\textsuperscript{\textsf{\scriptsize iD}}}}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{float}
        \floatplacement{table}{htbp}
        \floatplacement{figure}{htbp}
        \usepackage{placeins}
        \usepackage{setspace}
        \doublespacing
        \usepackage[mathlines]{lineno}
        \linenumbers
        \usepackage{textcomp}
        \usepackage{newunicodechar}
        \newunicodechar{≤}{\ensuremath{\leq}}
        \newunicodechar{≥}{\ensuremath{\geq}}
        \newunicodechar{×}{\ensuremath{\times}}
        \newunicodechar{β}{\ensuremath{\beta}}
        \newunicodechar{η}{\ensuremath{\eta}}
        \newunicodechar{χ}{\ensuremath{\chi}}
        \newunicodechar{σ}{\ensuremath{\sigma}}
        \newunicodechar{τ}{\ensuremath{\tau}}
        \newunicodechar{γ}{\ensuremath{\gamma}}
        \newunicodechar{ζ}{\ensuremath{\zeta}}
        \newunicodechar{φ}{\ensuremath{\varphi}}
        \newunicodechar{μ}{\ensuremath{\mu}}
        \newunicodechar{ν}{\ensuremath{\nu}}
        \newunicodechar{λ}{\ensuremath{\lambda}}
        \newunicodechar{Δ}{\ensuremath{\Delta}}
        \newunicodechar{ε}{\ensuremath{\varepsilon}}
        \newunicodechar{ρ}{\ensuremath{\rho}}
        \newunicodechar{→}{\ensuremath{\rightarrow}}
        \newunicodechar{↔}{\ensuremath{\leftrightarrow}}
        \newunicodechar{≈}{\ensuremath{\approx}}
        \newunicodechar{²}{\textsuperscript{2}}
        \setlength{\LTleft}{0pt}
        \setlength{\LTright}{\fill}
knitr:
  opts_chunk:
    dev: cairo_pdf
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
csl: elsevier-vancouver.csl
---

```{r setup}
#| include: false

# Find project root and change to it so here() works
orig_wd <- getwd()
project_root <- normalizePath(
  file.path(orig_wd, "..")
)
if (!file.exists(
  file.path(project_root, ".here")
)) {
  project_root <- orig_wd
}
setwd(project_root)

library(here)
library(data.table)
library(gt)
library(ggplot2)
library(knitr)
library(patchwork)

source(here("R/utils/config.R"))
source(here("R/utils/data_io.R"))
source(here("R/utils/tables.R"))
source(here("R/utils/plotting.R"))
source(here("R/utils/consort_diagram.R"))

config <- load_config()

# -------------------------------------------------
# Load pre-computed manuscript environment
# -------------------------------------------------
env.lst <- readRDS(
  here("outputs/manuscript_env.rds")
)
list2env(env.lst, envir = environment())
rm(env.lst)

# -------------------------------------------------------------------
# gt table rendering for LaTeX/PDF output
# -------------------------------------------------------------------

knit_print.gt_tbl <- function(x, ...) {
  if (knitr::is_latex_output()) {
    latex_code <- as.character(gt::as_latex(x))

    # Fix 1: Table width
    latex_code <- gsub(
      "\\\\begin\\{tabular\\*\\}\\{\\\\linewidth\\}\\{@\\{\\\\extracolsep\\{\\\\fill\\}\\}",
      "\\\\begin{tabular}{",
      latex_code
    )
    latex_code <- gsub(
      "\\\\end\\{tabular\\*\\}", "\\\\end{tabular}",
      latex_code
    )

    # Fix 2: HTML tags to LaTeX
    latex_code <- gsub(
      "<em>([^<]+)</em>", "\\\\textit{\\1}", latex_code
    )
    latex_code <- gsub(
      "<strong>([^<]+)</strong>", "\\\\textbf{\\1}", latex_code
    )
    latex_code <- gsub(
      "<sub>([^<]+)</sub>", "\\\\textsubscript{\\1}", latex_code
    )
    latex_code <- gsub(
      "<sup>([^<]+)</sup>", "\\\\textsuperscript{\\1}", latex_code
    )

    # Fix 3: Unicode symbols to LaTeX math
    latex_code <- gsub("Δ", "$\\\\Delta$", latex_code)
    latex_code <- gsub("β", "$\\\\beta$", latex_code)
    latex_code <- gsub("λ", "$\\\\lambda$", latex_code)
    latex_code <- gsub("σ", "$\\\\sigma$", latex_code)
    latex_code <- gsub("η", "$\\\\eta$", latex_code)
    latex_code <- gsub("χ", "$\\\\chi$", latex_code)
    latex_code <- gsub("τ", "$\\\\tau$", latex_code)
    latex_code <- gsub("γ", "$\\\\gamma$", latex_code)
    latex_code <- gsub(
      "²", "\\\\textsuperscript{2}", latex_code
    )
    latex_code <- gsub("\u2264", "$\\\\leq$", latex_code)
    latex_code <- gsub("\u2265", "$\\\\geq$", latex_code)
    latex_code <- gsub("\u2260", "$\\\\neq$", latex_code)
    latex_code <- gsub("×", "$\\\\times$", latex_code)

    # Fix 4: Unicode text commands missing {} terminator
    latex_code <- gsub(
      "\\\\(texttimes|texttwosuperior|texttwosuperiorp|textendash|textemdash)([A-Za-z])",
      "\\\\\\1{}\\2", latex_code
    )

    # Fix 5: Footnote minipage width
    latex_code <- gsub(
      "\\\\begin\\{minipage\\}\\{\\\\linewidth\\}",
      "\\\\begin{minipage}[t]{0.8\\\\textwidth}\\\\raggedright\\\\small",
      latex_code
    )

    # Fix 6: Column alignment (first left, rest right)
    fix_alignment <- function(code) {
      pattern <- "(\\\\begin\\{(?:tabular|longtable)\\}\\{)([lcr]+)(\\})"
      m <- gregexpr(pattern, code, perl = TRUE)
      if (m[[1]][1] == -1) return(code)
      matches <- regmatches(code, m)[[1]]
      for (match in matches) {
        col_spec <- gsub(".*\\{([lcr]+)\\}$", "\\1", match)
        n_cols <- nchar(col_spec)
        if (n_cols > 1) {
          new_spec <- paste0(
            "l", paste(rep("r", n_cols - 1), collapse = "")
          )
          new_match <- gsub(
            "\\{[lcr]+\\}$", paste0("{", new_spec, "}"), match
          )
          code <- sub(match, new_match, code, fixed = TRUE)
        }
      }
      code
    }
    latex_code <- fix_alignment(latex_code)

    # Fix 7: Table spacing
    latex_code <- gsub(
      "\\\\begin\\{tabular\\}",
      "\\\\renewcommand{\\\\arraystretch}{1.15}\\\\begin{tabular}",
      latex_code
    )

    # Fix 8: Escape [ ] in table cells
    latex_code <- gsub(
      "\\\\addlinespace\\[([^]]+)\\]",
      "<<ADDLNSP_\\1_>>", latex_code
    )
    latex_code <- gsub(
      "\\\\\\\\\\[([0-9.]+pt)\\]",
      "<<LINEBRK_\\1_>>", latex_code
    )
    latex_code <- gsub(
      "\\\\begin\\{minipage\\}\\[([a-z])\\]",
      "<<MINIPAGE_\\1_>>", latex_code
    )
    latex_code <- gsub("\\[", "{[}", latex_code)
    latex_code <- gsub("\\]", "{]}", latex_code)
    latex_code <- gsub(
      "<<ADDLNSP_([^_]+)_>>",
      "\\\\addlinespace[\\1]", latex_code
    )
    latex_code <- gsub(
      "<<LINEBRK_([^_]+)_>>",
      "\\\\\\\\[\\1]", latex_code
    )
    latex_code <- gsub(
      "<<MINIPAGE_([a-z])_>>",
      "\\\\begin{minipage}[\\1]", latex_code
    )

    knitr::asis_output(latex_code)
  } else {
    gt:::knit_print.gt_tbl(x, ...)
  }
}
registerS3method(
  "knit_print", "gt_tbl", knit_print.gt_tbl,
  envir = asNamespace("knitr")
)

# Domain labels for cognitive measures
DOMAIN_LABELS <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

# Helper: Format-aware markdown (handles LaTeX vs HTML)
fmt_crossformat <- function(gt_tbl, columns, rows = everything()) {
  if (knitr::is_latex_output()) {
    gt::fmt(
      gt_tbl, columns = {{ columns }},
      rows = {{ rows }}, fns = function(x) {
        x <- gsub("\\*([^*]+)\\*", "\\\\textit{\\1}", x)
        x <- gsub("~([^~]+)~", "\\\\textsubscript{\\1}", x)
        x
      }
    )
  } else {
    gt::fmt_markdown(
      gt_tbl, columns = {{ columns }}, rows = {{ rows }}
    )
  }
}
```

---

# Highlights {.unnumbered}

- Age-adjusted cardiovascular risk does not accelerate hippocampal atrophy
- Hippocampal-cognitive coupling persists regardless of how risk is measured
- FRS-brain associations in elderly samples primarily reflect age-weighting
- A reverse pattern in amyloid-negative controls supports a cognitive reserve effect
- MIMIC-based age-adjustment generalizes to any age-confounded composite score

---

# Abstract {.unnumbered}

**INTRODUCTION:**
The Framingham Risk Score (FRS) indexes
cardiovascular risk (CVR), but age-weighting may
confound associations with brain and cognitive
outcomes.

**METHODS:**
In `r n_subjects` amyloid-positive ADNI participants,
we compared FRS against a Multiple Indicators
Multiple Causes (MIMIC)-derived age-adjusted measure
(CVR~mimic~) using sex-stratified linear
mixed-effects (LME) and latent growth curve mediation
(LGCM) models of hippocampal-to-ventricle ratio
(HVR)--cognitive coupling.

**RESULTS:**
FRS predicted hippocampal atrophy in
`r n_of_total.fn(abs_stats$frs_a_sig, 6L)` LGCM
models; CVR~mimic~ in
`r n_of_total.fn(abs_stats$cvr_a_sig, 6L)`.
HVR--cognitive coupling held in
`r n_of_total.fn(abs_stats$frs_b_sig, 6L)` FRS and
`r n_of_total.fn(abs_stats$cvr_b_sig, 6L)`
CVR~mimic~ models. Indirect effects reached
significance in
`r n_of_total.fn(abs_stats$frs_ind_sig, 6L)` FRS and
`r n_of_total.fn(abs_stats$cvr_ind_sig, 6L)`
CVR~mimic~ models. LME 3-way interactions (years ×
risk × HVR) survived FDR correction in
`r n_of_total.fn(fdr_summary.lst$frs_n_fdr, 6L)` FRS
versus
`r n_of_total.fn(fdr_summary.lst$cvr_n_fdr, 6L)`
CVR~mimic~ models.

**DISCUSSION:**
FRS "effects" on hippocampal-cognitive decline
largely reflect age-related variance. Age-adjusted
measures complement FRS by isolating
cardiovascular effects from aging.

**Keywords:** Cardiovascular risk, Framingham Risk
Score, hippocampus, cognitive decline, Alzheimer's
disease, MIMIC model, age-confounding

## Research in Context {.unnumbered}

**Systematic Review:**
The Framingham Risk Score (FRS) predicts brain
atrophy and cognitive decline across cohorts
[@zsoldos_association_2020; @seshadri_stroke_2004;
@song_associations_2020; @meusel_cardiovascular_2019].
Yet age dominates the FRS
[@dagostino_general_2008; @koller_development_2012]
and accounts for most of its predictive value in
older adults
[@pencina_quantifying_2019; @ko_calibration_2020],
suggesting these associations may reflect age
confounding. No prior study has compared FRS against
an age-adjusted latent measure.

**Interpretation:**
FRS indirect effects on cognitive decline via
hippocampal atrophy primarily reflect age-weighting;
age is itself a legitimate vascular proxy
[@villeneuve_impact_2009]. Partialling out age
(CVR~mimic~) nullified the
cardiovascular-risk-to-hippocampal pathway, while
HVR-cognitive coupling persisted, indicating
coupling reflects neurodegeneration. Prior FRS-brain
reports likely conflate age-driven and
cardiovascular effects.

**Future Directions:**
Replication in independent aging cohorts with
longitudinal cardiovascular measurement is needed. The MIMIC age-adjustment framework can be
applied to any composite risk score where age
confounding is a concern. Intervention trials should
test whether cardiovascular management preserves
hippocampal-cognitive coupling using age-adjusted
endpoints rather than FRS.

---

# Background

Cardiovascular risk (CVR) factors are increasingly
recognized as modifiable contributors to cognitive
decline and dementia risk
[@livingston_dementia_2017; @livingston_dementia_2024; @ouldred_vascular_2010;
@viswanathan_vascular_2009;
@gorelick_vascular_2011;
@iadecola_neurovascular_2019].
The Framingham Risk Score (FRS), a widely used
composite measure of 10-year risk of cardiovascular
event [@dagostino_general_2008] — originally
designed and validated for midlife application — has
been
associated with reduced brain volumes, compromised
white matter integrity, and accelerated cognitive
decline across multiple aging cohorts
[@zsoldos_association_2020;
@seshadri_stroke_2004; @song_associations_2020;
@kaffashian_predictive_2011]. However, the FRS is
heavily weighted by age, with log-hazard
coefficients of 3.06 for men and 2.33 for women
[@dagostino_general_2008], making age the dominant
predictor in the composite, with different effects
by sex. In older samples where
age variance is restricted, this creates substantial
confounding between FRS and any age-related outcome
[@koller_development_2012;
@rodondi_framingham_2012].

Age is itself a strong proxy for cumulative
vascular exposure
[@villeneuve_impact_2009; @dagostino_general_2008],
and the FRS age-weighting therefore carries
legitimate clinical information. However, this
creates an interpretive challenge when FRS is used
to predict age-related outcomes: because age
simultaneously drives FRS through its formula
coefficients and independently drives
neurodegenerative outcomes through biological aging,
it acts as a common cause of both the predictor and
the outcome. Simply controlling for age in
regression models does not fully resolve this
confound, because age is not an external confounder
but an internal component of the score itself. In elderly cohorts such as ADNI,
this problem is compounded by the fact that the
ADSP-PHC-derived FRS variable is capped at 30%
10-year risk
[@dagostino_general_2008; @lee_fmnl2_2022],
compressing variance in the upper range and further
limiting sensitivity.

To address this, we employed a Multiple Indicators
Multiple Causes (MIMIC) model
[@joreskog_estimation_1975; @muthen_general_1984],
which treats age as a *cause* of cardiovascular risk
rather than an indicator. The structural equation
$\text{CVR} = \gamma \cdot \text{Age}_z + \zeta$
(where CVR denotes the latent cardiovascular risk
factor) partitions observed cardiovascular risk into
an age-explained
component ($\gamma \cdot \text{Age}_z$) and a
residual ($\zeta$) that captures
cardiovascular risk independent of age
[@zhang_bayesian_2024]. The resulting factor scores,
termed CVR~mimic~, are orthogonal to age by
construction (r ≈ 0), enabling a direct test of
whether FRS-brain associations reflect true
cardiovascular pathophysiology or age-related
confounding.

The primary objective of this study was to
determine whether FRS effects on
hippocampal-cognitive coupling persist when
age-confounding is removed. We tested three
hypotheses. First, if FRS effects on hippocampal
atrophy are driven by age-weighting, CVR~mimic~
should show a null or attenuated effect of
cardiovascular risk on hippocampal atrophy rate
(the CVR~mimic~ → hippocampal-to-ventricle ratio
[HVR] slope path) in latent growth curve mediation
models (LGCM). Second, if FRS mediation is
age-confounded, CVR~mimic~ indirect effects through
HVR decline should likewise be null. Third, even
when mediation is absent, CVR~mimic~ may still
moderate the brain-cognition relationship via a
three-way interaction (years from baseline [YRS]
× CVR [either FRS or CVR~mimic~]
× HVR) in linear mixed-effects (LME) models,
reflecting a
reserve-like mechanism whereby cardiovascular burden
modulates the coupling between hippocampal structure
and cognitive performance without directly
accelerating atrophy. All analyses were stratified
by sex, given established sex differences in
cardiovascular risk profiles, brain aging
trajectories, and AD pathophysiology
[@ferretti_sex_2018; @mielke_consideration_2022].

---

# Methods

## Study Design and Participants

Data used in the preparation of this article were
obtained from the Alzheimer's Disease Neuroimaging
Initiative (ADNI) database (adni.loni.usc.edu). The
ADNI was launched in 2003 as a public-private
partnership, led by Principal Investigator Michael W.
Weiner, MD. The primary goal of ADNI has been to test
whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological
markers, and clinical and neuropsychological
assessment can be combined to measure the progression
of mild cognitive impairment (MCI) and early
Alzheimer's disease (AD). ADNI data were downloaded
from the LONI-IDA portal between December 2025 and
February 2026 (latest download: 10 February 2026).

Data were drawn from ADNI phases 1, GO, 2, and 3.
We included only amyloid-positive participants,
defined by PET Centiloid > 25
[@klunk_centiloid_2015] across florbetapir (FBP),
florbetaben (FBB), and Pittsburgh Compound B (PIB)
tracers, or CSF amyloid-positive status via Gaussian
Mixture Model classification
[@timsina_harmonization_2024]. A participant was
classified as amyloid-positive if positivity was
established at any available timepoint (PET or CSF);
once classified, all longitudinal visits for that
participant were included. Additional inclusion
criteria required available MRI-derived hippocampal
and ventricular volumes
[@fernandez-lozano_enhanced_2025], cardiovascular
risk factor data (FRS components)
[@lee_fmnl2_2022; @lee_reliability_2023], and at
least one cognitive assessment with Alzheimer's
Disease Sequencing Project Phenotype Harmonization
Consortium (ADSP-PHC) harmonized composite scores
[@kang_genome-wide_2023]. The resulting sample
encompassed the full spectrum of amyloid-positive
individuals, including cognitively unimpaired (CU),
mild cognitive impairment (MCI), and Alzheimer's
disease dementia (AD). Participant inclusion and
exclusion flow is shown in
Figure \ref{fig-consort}.

A separate normative comparison sample of
amyloid-negative, cognitively unimpaired ADNI
participants with structural MRI within the UK
Biobank (UKB) age range (46--82 years) was used to
validate the transfer of UKB-derived z-scores to the
ADNI cohort (see Normative Transfer Validation). Note
that UKB participants are population-based without
amyloid testing; the normative GAMLSS (Generalized
Additive Models for Location, Scale, and Shape)
models
therefore reflect the general population rather than
a confirmed amyloid-negative sample. The UKB
normative reference is sex-imbalanced (56.2%
female), reflecting the source imaging cohort and
quality-control filtering
[@fernandez-lozano_hippocampal_2026]; sex-stratified
GAMLSS fitting ensures within-sex z-score
calibration is unaffected by this imbalance.

```{r fig-consort}
#| output: asis

cc <- consort_counts
tikz_code <- generate_consort_tikz(
  n_database = cc$n_database,
  n_with_amyloid = cc$n_with_amyloid,
  n_amyloid_pos = cc$n_amyloid_pos,
  n_with_mri = cc$n_with_mri,
  n_lme = cc$n_lme,
  n_lgcm = cc$n_lgcm,
  n_aneg_mri = cc$n_aneg_mri,
  n_aneg_cu_mri = cc$n_aneg_cu_mri,
  n_age_comparable = cc$n_age_comparable,
  ukb_age_range = cc$ukb_age_range,
  n_aneg_sensitivity =
    aneg_lme.res$sample_info$n_subjects
)

cat("\\begin{figure}[htbp]\n")
cat(tikz_code)
cap <- paste0(
  "\\caption{\\textbf{CONSORT flow diagram ",
  "showing participant selection.} ",
  "Left: amyloid-positive",
  " analysis track from LME cohort to LGCM ",
  "subsample. Right: amyloid-negative, ",
  "cognitively unimpaired participants split ",
  "into the A\\textsuperscript{--}/CU ",
  "sensitivity subsample and the UKB ",
  "normative comparison subsample. ",
  "ADSP-PHC = Alzheimer's Disease ",
  "Sequencing Project Phenotype Harmonization ",
  "Consortium; CL = Centiloid; CSF = ",
  "cerebrospinal fluid; CU = cognitively ",
  "unimpaired; EDT = estimated disease time; ",
  "MRI = magnetic resonance imaging; ",
  "UKB = UK Biobank.}\n"
)
cat(cap)
cat("\\label{fig-consort}\n")
cat("\\end{figure}\n")
```

## Neuroimaging: Hippocampal-to-Ventricle Ratio (HVR)

T1-weighted magnetization-prepared rapid gradient
echo (MPRAGE) sequences were acquired on 1.5T
and 3T scanners across ADNI sites using standardized
acquisition protocols
[@jack_alzheimers_2008]. Hippocampal (HC) and temporal
horn of the lateral ventricle (VC) volumes were
obtained using a convolutional neural network (CNN)
trained on manual labels
[@fernandez-lozano_enhanced_2025], which demonstrated
low failure rates in ADNI compared to multi-atlas
and patch-based alternatives. Volumes were provided
in stereotaxic space (mm$^3$) and converted to
native space via subject-specific scale factors.

We defined the hippocampal-to-ventricle ratio (HVR)
as:

$$
\text{HVR} = \frac{
  \text{HC}_L + \text{HC}_R}{
  \text{HC}_L + \text{HC}_R +
  \text{VC}_L + \text{VC}_R}
$$ {#eq-hvr}
computed on bilateral summed native-space volumes
[@schoemaker_hippocampal_ventricle_2019]. This ratio
captures hippocampal integrity relative to
surrounding cerebrospinal fluid (CSF)-filled spaces,
where hippocampal atrophy is accompanied by
compensatory expansion of the temporal horns
(i.e., ex vacuo dilation). HVR approaches 1.0 when
hippocampal volume is large relative to the
temporal horns (preserved structure) and approaches 0.0
when the hippocampus is small relative to the
temporal horns (atrophy with ventricular expansion). The
ratio inherently normalizes for head size without
requiring intracranial volume (ICV) correction, as
both numerator and denominator scale proportionally
with head size
[@fernandez-lozano_hippocampal_2026].

HVR z-scores were derived from Generalized Additive
Models for Location, Scale, and Shape
(GAMLSS; @rigby_generalized_2005) fit on UKB data
(N = 27,434 cognitively healthy adults, age 45--82).
The distribution family was Box-Cox Cole and Green
(BCCG), which models three distributional
parameters — location ($\mu$, the conditional
median), scale ($\sigma$, the coefficient of
variation), and skewness ($\nu$, the Box-Cox power) — as smooth
functions of covariates: age (penalized cubic
spline), education (linear), and intracranial cavity
volume (ICC), with separate models fit for each sex.
The resulting z-score quantifies how many standard
deviations an individual's observed HVR deviates
from the conditional median predicted by the GAMLSS
model given their age, sex, education, and head
size. Negative z-scores indicate disproportionate
hippocampal volume loss relative to the normative
reference population. These are cross-sectional
deviation scores, not measures of longitudinal
change rate.

### Normative Transfer Validation

Valid cross-cohort application of UKB-trained GAMLSS
models requires verification that z-scores are
calibrated in the target ADNI population. We
performed validation in a comparable ADNI subsample
of amyloid-negative, cognitively unimpaired
participants within the UKB age range
(46--82 years), where z-scores would be expected to
approximate a standard normal distribution (mean
zero, standard deviation one). We compared mean
z-scores, standard deviations, and 95% confidence
intervals against expected $N(0, 1)$ and visualized
age-bin-specific calibration via forest plot.

## Cardiovascular Risk Assessment

### Framingham Risk Score (FRS)

We obtained FRS values from the Alzheimer's Disease
Sequencing Project Phenotype Harmonization Consortium
(ADSP-PHC) Cardiovascular Risk Factor dataset
[@lee_fmnl2_2022; @lee_reliability_2023]. The ADSP-PHC
computed FRS using the Framingham 2008 simple
office-based 10-year cardiovascular disease (CVD)
risk score via the
`ascvd_10y_frs_simple` function from the CVrisk R
package [@castro_cvrisk_2026]. Despite the function
name, this implements the Framingham general
cardiovascular risk algorithm
[@dagostino_general_2008], not the American College
of Cardiology/American Heart Association (ACC/AHA)
atherosclerotic cardiovascular disease (ASCVD)
Pooled Cohort Equations. The score estimates
10-year cardiovascular disease risk based on sex,
age (≥ 30 years), body mass index (BMI, kg/m²),
systolic blood pressure (mmHg), use of blood
pressure medication, current smoking status, and
diabetes status. This is the "simple" FRS variant
that substitutes BMI for the lipid panel (total and
high-density lipoprotein [HDL] cholesterol) required
by the general FRS
[@dagostino_general_2008]. Age coefficients in the
log-hazard scale are 3.06 (men) and 2.33 (women),
making age the dominant component in older adults.
The ADSP-PHC-derived variable is capped at 30%
10-year risk, creating a ceiling effect in elderly
samples.

### MIMIC Model for Age-Adjusted CVR (CVR~mimic~)

We specified the CVR~mimic~ latent factor as a
Multiple Indicators Multiple Causes (MIMIC) model
[@joreskog_estimation_1975; @muthen_general_1984], in which age is an
exogenous cause of the latent cardiovascular risk
factor and the disturbance ζ captures residual
cardiovascular risk independent of age [@zhang_bayesian_2024].

#### Model Specification

$$
\begin{aligned}
\text{Structural:} \quad &
  \text{CVR} = \gamma \cdot \text{Age}_z + \zeta \\
\text{Measurement:} \quad &
  x_j = \lambda_j \cdot \text{CVR} + \delta_j,
  \quad j = 1, \ldots, 5 \\
\text{Residual covariances:} \quad &
  \text{Cov}(\delta_{\text{SBP}},
    \delta_{\text{CHOL}}) \neq 0; \;
  \text{Cov}(\delta_{\text{SBP}},
    \delta_{\text{CREAT}}) \neq 0
\end{aligned}
$$ {#eq-mimic}

The model includes five measurement indicators
spanning vascular, metabolic, and renal
cardiovascular risk domains: systolic blood pressure
(SBP; z-scored, continuous), hypertension (history of
hypertension; binary), fasting glucose (z-scored,
continuous), total cholesterol (z-scored, continuous),
and serum creatinine (z-scored, continuous). We specified two
correlated residuals: SBP ↔ Cholesterol, reflecting
shared vascular stiffness and lipid metabolism
pathways whereby arterial compliance links blood
pressure regulation to lipid profiles beyond the
general cardiovascular risk factor; and
SBP ↔ Creatinine, reflecting
the renal-cardiovascular axis whereby kidney function
directly influences blood pressure regulation through
the renin-angiotensin system, partially independent
of overall cardiovascular risk. Both modification
indices and
substantive physiological reasoning supported these
specifications.

#### Indicator Selection

We determined the final set of five indicators
through iterative model refinement, evaluating
factor loadings, modification indices, and
substantive interpretability at each step. Several
candidate indicators were evaluated but excluded.
BMI was dropped because its factor loading became
non-significant once residual covariances with
hypertension, glucose, and cholesterol were modeled,
indicating its cardiovascular risk variance is fully
captured by the other retained indicators. Diabetes, heart
disease, and stroke (ADSP-PHC binary medical history
flags) showed near-perfect tetrachoric correlations
due to a harmonization artifact in how self-reported
conditions were coded across ADNI phases, rendering
them empirically indistinguishable. Smoking status
had a non-significant factor loading, likely
reflecting survivorship bias in an elderly cohort.
Blood pressure medication showed near-complete
redundancy with hypertension status. Specific
loading estimates and model selection diagnostics
are reported in the supplement.

#### Estimation and Fit

We evaluated model fit with the robust maximum
likelihood (MLR)-scaled
comparative fit index (CFI), Tucker-Lewis index
(TLI), root mean square error of approximation
(RMSEA), and standardized root mean square residual
(SRMR).

#### Score Extraction and Age Adjustment

We extracted factor scores in three steps. First,
we obtained predicted latent factor scores (η)
from the fitted MIMIC model via empirical Bayes
modal estimation. Second, we computed age-adjusted
cardiovascular risk scores (ζ, hereafter
CVR~mimic~) as ζ = η − γ × Age~c~, where γ is
the estimated regression of the latent
cardiovascular risk factor on centered age from the
structural path; this removes the age-predicted
component of the latent variable, ensuring
CVR~mimic~ represents residual
cardiovascular risk after partialling out age at the
latent level rather than at the indicator level.
Third, we z-standardized the resulting ζ scores on
the common analysis sample (participants with
complete data for both FRS components and MIMIC model
indicators) for comparability with FRS. This latent-level residualization operates on
the measurement-error-free latent variable and is
equivalent to extracting the disturbance term of the
MIMIC structural equation [@bollen_structural_1989].
By construction, *r*(CVR~mimic~, Age) ≈ 0.

#### Measurement Invariance

We tested measurement invariance across sex in four
steps — configural, metric, scalar, and strict —
using the confirmatory factor analysis (CFA)
measurement model only (without the Age → cardiovascular risk structural
path). Measurement invariance evaluates
whether factor loadings, intercepts, and residual
variances are equivalent across groups, which is
required before interpreting sex-stratified analyses.
We evaluated invariance with the Satorra-Bentler
scaled chi-square difference test
[@satorra_scaled_2001], which correctly handles the
non-subtractable nature of MLR-scaled test
statistics, and with change in fit indices
(|ΔCFI| ≤ 0.01, ΔRMSEA ≤ 0.015). When the two
criteria disagreed, we adopted the stricter result.

### CVR~mimic~ Validation

We validated CVR~mimic~ by examining its correlation
with age (expected ≈ 0, confirming successful
age-partialling), correlation with FRS (moderate
expected, reflecting shared cardiovascular
indicators), distribution properties (approximate
standard normal), and FRS ceiling analysis
(percentage at the 30% risk cap) to document
variance compression in the elderly sample.

### Cardiovascular Risk Stratification

For trajectory visualizations, we defined CV risk
groups by median split of baseline FRS (for FRS
analyses) or baseline CVR~mimic~ (for CVR~mimic~
analyses), and similarly defined HVR groups by
median split of baseline HVR z-score. Groups were
assigned at baseline and held constant across
follow-up.

### Cardiovascular Risk Predictor Standardization

We z-scored both FRS and CVR~mimic~ on the common
analysis sample (defined above) prior to all
downstream analyses. This ensures
that regression coefficients from both linear
mixed-effects (LME) models and latent growth curve
models (LGCM) are on an identical scale (effect per
1 SD change), enabling direct comparison of effect
magnitudes between FRS and CVR~mimic~. We retained
raw (pre-standardized) FRS values for descriptive
statistics and ceiling analyses.

## Disease Time Estimation

Calendar time (years from baseline) assumes all
participants progress at the same rate, which is
unrealistic in a heterogeneous AD sample spanning
CU, MCI, and dementia stages. Estimated
Disease Time (EDT) aligns participants on a common
disease timeline by estimating each individual's
position along an underlying progression curve
with respect to AD dementia diagnosis
[@raket_statistical_2020]. We computed EDT using the
progmod R package with a simultaneous multivariate
model fitting the Alzheimer's Disease Assessment
Scale-Cognitive subscale (ADAS-Cog) 13 and the
Mini-Mental State Examination (MMSE) jointly,
sharing
a common disease time random effect across scales
[@kuhnel_simultaneous_2021]. The simultaneous
approach provides more stable individual time
estimates than univariate models because it borrows
concordant information across cognitive instruments.
EDT values were mean-centered for numerical stability
while retaining the years metric; centering does not
affect between-subject differences in disease timing.
For participants with missing EDT at later timepoints
(due to dropout), values were extrapolated using a
hybrid approach: a linear slope computed from the
individual's observed EDT values when at least two
were available, or the sex-stratified population mean
EDT slope otherwise, to minimize attrition bias from
listwise deletion.

## Cognitive Assessment

We used ADSP-PHC composite scores
[@kang_genome-wide_2023] for three cognitive domains:
memory (episodic memory composite: Wechsler Memory
Scale-Revised [WMS-R] Logical Memory, Rey Auditory
Verbal Learning Test [AVLT], ADAS-Cog word recall,
MMSE/Montreal Cognitive Assessment [MoCA] memory
items), language (language/semantic composite:
category fluency, Boston Naming Test, ADAS-Cog
naming/commands), and executive function (executive
composite: Trail Making A/B, Wechsler Adult
Intelligence Scale-Revised [WAIS-R] Digit
Span/Symbol, clock drawing). Scores were
co-calibrated across multiple cohorts via a bifactor
item banking approach, initially across ADNI, Adult
Changes in Thought (ACT), and Religious Orders
Study/Memory and Aging Project (ROS/MAP), with
subsequent addition of the National Alzheimer's
Coordinating Center (NACC). Higher composite scores
indicate better cognitive performance across all
three domains. Each score has an associated standard
error (SE) from the CFA scoring model, reflecting
measurement precision at each observation. In the
LGCM, cognitive composites enter as single indicators
with residual variances fixed to SE² (the squared
standard error), embedding known measurement error
directly from the ADSP-PHC model rather than
estimating it as a free parameter.

## Statistical Analysis

### Analytical Framework

All primary analyses (LME and LGCM) were run twice:
once with FRS and once with CVR~mimic~ as the
cardiovascular risk predictor. The key test is whether FRS effects
disappear when replaced by age-adjusted CVR~mimic~,
which would implicate age-confounding rather than
true cardiovascular pathophysiology. The central
question is whether cardiovascular risk influences
the hippocampal-cognitive relationship, and whether
the answer depends on how cardiovascular risk is
measured.

All models were sex-stratified (each sex analyzed
separately),
following the National Institutes of Health
policy on Sex as a Biological
Variable @clayton_policy_2014. Sex
differences in AD pathophysiology, cardiovascular
risk profiles, and brain aging trajectories motivate
stratification rather than inclusion of sex as a
covariate [@ferretti_sex_2018;
@mielke_consideration_2022]. Stratification allows
effect sizes, variance components, and covariate
relationships to differ freely between sexes; sex is
omitted from covariates because it defines the
strata. Covariates in all models include age at
baseline, education (years), and Apolipoprotein E
(APOE) ε4 carrier status.

Primary inferences are based on uncorrected per-test
p-values given the pre-specified, hypothesis-driven
nature of the analyses
[@rothman_no_1990; @perneger_whats_1998]. As a
supplementary check, we applied FDR correction
(Benjamini-Hochberg) within each cardiovascular risk measure across
the 6 LME tests (3 domains × 2 sexes). FRS and
CVR~mimic~ were corrected separately because they
address distinct hypotheses. Whether effects
surviving uncorrected thresholds also survive FDR is
reported in Results. The key inferential comparison
(a-path FRS vs a-path CVR~mimic~) is inherently
directional and does not require correction across
independent model comparisons
[@perneger_whats_1998]. For the LGCM mediation
models (12 tests: 2 sexes × 3 domains ×
2 predictors), no FDR correction was applied:
inference rests on the consistency of the a-path
pattern across models rather than on any single
test.

**Analytical roadmap.** The two modelling frameworks
address distinct questions. The LME tests
*moderation*: does cardiovascular risk change the
HVR-cognition
coupling over time, assessed via the 3-way
YRS (years from baseline) × CVR × HVR interaction? The LGCM tests
*mediation*: does cardiovascular risk drive cognitive
decline through
hippocampal atrophy, assessed via the indirect path
cardiovascular risk → HVR slope → cognitive slope?
Comparing FRS
and CVR~mimic~ across both frameworks isolates the
role of age-weighting in observed cardiovascular
risk-brain
associations.

**Formal predictor comparison.** Because FRS and
CVR~mimic~ were fitted on the same participants,
paired tests are required to compare coefficients.
In the LGCM, shared-seed bootstrap ensures identical
resamples for both predictors, yielding a paired
$\Delta_a = a_\text{FRS} - a_{\text{CVR}_\text{mimic}}$ with
bias-corrected 95% CI. In the LME, clustered
bootstrap (resampling subjects with replacement)
provides a paired
$\Delta\beta = \beta_\text{FRS} - \beta_{\text{CVR}_\text{mimic}}$
for the 3-way interaction, with bias-corrected
95% CI.

### Methodological Validations

#### EDT Linearity Validation

We used likelihood ratio tests (LRT) with LME models
comparing linear vs quadratic EDT specifications,
testing whether adding a quadratic EDT term improves
model fit. This complements the LGCM quadratic
convergence findings (see below).

#### Simulation Study: Standard Error-Constrained
Validation

A Monte Carlo simulation validated the
SE-constrained LGCM specification (residual variance
fixed to SE²) against free estimation (residual
variance freely estimated). This is a methodological
validation, not a power analysis — it validates the
estimation approach under known truth, not detection
rates for specific effect sizes. The SE-constraint
is a key methodological innovation: cognitive
composites from the ADSP-PHC CFA have known
observation-specific SEs, so fixing residual
variance to SE² uses this information rather
than discarding it.

Data were generated from a univariate latent growth
model with parameters calibrated from the observed
ADNI LGCM parallel process models (averaged across
all sex × domain combinations): intercept mean =
`r simulation.res$true_params$intercept_mean`
(z-scored composites), slope mean =
`r simulation.res$true_params$slope_mean` per year
EDT (population-average cognitive decline rate),
intercept variance =
`r simulation.res$true_params$intercept_var`
(between-person variability in baseline level),
slope variance =
`r simulation.res$true_params$slope_var`
(between-person variability in decline rate), HVR →
Slope true effect =
`r simulation.res$true_params$hvr_effect` (mean
b-path coupling estimate), measurement SE =
`r simulation.res$true_params$se_true` (median
ADSP-PHC composite SE, SD =
`r simulation.res$true_params$se_variation` across
observations), and EDT with mean =
`r simulation.res$true_params$edt_mean` years,
SD = `r simulation.res$true_params$edt_sd`, change =
`r simulation.res$true_params$edt_change` per visit
(observed EDT distribution and inter-visit spacing).
To assess robustness, the two most consequential
parameters — the HVR effect and slope variance —
were varied at 0.5×, 1×, and 1.5× their observed
values, yielding 9 simulation conditions. Two
approaches were compared across
`r simulation.res$config$n_reps` replications per
condition at
N = `r paste(simulation.res$config$sample_sizes, collapse = " and N = ")`:
(a) residual variance fixed to SE² (SE-constrained)
and (b) residual variance freely estimated.
Performance metrics include parameter bias (mean
difference from truth), RMSE (root mean squared
error), and 95% CI coverage probability.

### LME Growth Models with 3-Way Interaction

The model specification (per sex) is:

$$
\begin{aligned}
\text{Cognition}_{it} = & \; (\beta_0 + b_{0i}) +
  (\beta_1 + b_{1i}) \cdot \text{YRS}_{it} \\
& + \text{YRS}_{it} \times \text{CVR}_i \times
  \text{HVR}_{zit} \;\text{(all 2- and 3-way terms)} \\
& + \text{YRS}_{it}^2 \times \text{CVR}_i \times
  \text{HVR}_{zit} \;\text{(all 2- and 3-way terms)} \\
& + \boldsymbol{X}_i\boldsymbol{\gamma} + \epsilon_{it}
\end{aligned}
$$ {#eq-lme}
where the linear block (YRS × CVR × HVR) expands
to all lower-order terms (YRS × CVR, YRS × HVR,
CVR × HVR, and the individual main effects) plus
the 3-way interaction; the quadratic block
(YRS^2^ × CVR × HVR) similarly expands. R's `*` formula operator generates
each product term exactly once — two-way terms shared
across blocks are not duplicated. The quadratic interaction terms test
whether cardiovascular risk moderation of the
HVR-cognition
relationship accelerates (or decelerates) over
follow-up. Random effects include subject-specific
intercepts ($b_{0i}$) and linear slopes ($b_{1i}$).
Cardiovascular risk is entered as either FRS or
CVR~mimic~ in
parallel model sets. The hypothesis tested is whether
cardiovascular risk moderates how the HVR-cognition
relationship
evolves over time; a negative 3-way β means higher
cardiovascular risk increasingly weakens the
protective
HVR-cognition association over follow-up. Inclusion
of the quadratic time term (YRS²) was evaluated
via
likelihood ratio tests comparing nested linear and
quadratic models across all 12 specifications
(2 sexes × 2 brain measures × 3 domains); results
are reported in Appendix B (Table B2).

### LGCM Parallel Process Models

We specified LGCM with parallel HVR and cognitive
trajectories
[@bollen_latent_2006; @muniz-terrera_latent_2017] to
test whether cardiovascular risk affects the
hippocampal-cognitive
relationship via mediation (CVR → HVR slope →
Cognitive slope).

#### Model Specification

$$
\begin{aligned}
y_{it} &= \eta_{0i} + \eta_{1i} \cdot
  \text{EDT}_{it} + \epsilon_{it}
  \\[4pt]
\eta_{0i} &= \alpha_0 +
  \boldsymbol{X}_i \boldsymbol{\gamma}_0 + \zeta_{0i}
  \\
\eta_{1i} &= \alpha_1 +
  \boldsymbol{X}_i \boldsymbol{\gamma}_1 + \zeta_{1i}
\end{aligned}
$$ {#eq-lgcm}
where $y_{it}$ is the observed score (HVR z-score
  or cognitive composite) for person $i$ at occasion
  $t$; $\eta_{0i}$ and $\eta_{1i}$ are
  person-specific intercepts and slopes; EDT~it~ is
  the individually-varying time score (see below);
  $\epsilon_{it}$ is the occasion-specific residual;
  and $\boldsymbol{X}_i$ includes cardiovascular risk predictor, age,
  education, and APOE ε4 status.

Two parallel processes (HVR and cognitive composite)
are specified simultaneously, with their latent
growth factors allowed to covary or be linked by
directional paths depending on the model, across up
to 6 measurement occasions per participant. Cognitive
composite residual variances were fixed to SE² from
the ADSP-PHC CFA scoring model, embedding known
heteroscedastic measurement error directly rather
than estimating it as a free parameter; this yields
more precise latent slope estimates by separating
true variance from measurement noise. For HVR,
occasion-specific residual variances were freely
estimated; a likelihood ratio test against an
equality-constrained alternative (single residual
variance pooled across timepoints) rejected the
constraint ($\Delta\chi^2 = 35.0$, df = 5,
$p < 10^{-5}$; Male × Memory × CVR~mimic~ cell),
consistent with the biologically expected
heteroscedasticity of brain atrophy variance over
disease time [@grimm_residual_2010].

We first attempted fully random quadratic models
(6 latent factors per domain); when these did not
converge, we adopted a fixed-quadratic
specification: two additional phantom latent factors
(HVR quadratic, cognitive quadratic) with variance
fixed to near zero and free population means,
capturing population-level curvature (2 additional
free parameters) without individual-level quadratic
variation. Likelihood ratio tests comparing this
fixed-quadratic model against a purely linear model
determined whether quadratic curvature improved fit.
We also attempted to embed the full
MIMIC measurement model within the OpenMx parallel
process framework; convergence was insufficient, so
we used pre-computed ζ scores instead (see Score
Extraction above).

#### Time Metric

Slope loadings use individually-varying time scores
determined by each participant's EDT, which maps
each visit to a position on the disease progression
timeline rather than calendar time. Because
individually-varying time scores give each
participant a unique implied
covariance structure, standard fit indices (CFI,
TLI, RMSEA) that require a common saturated
reference model cannot be computed
[@grimm_growth_2016; @mehta_putting_2000].

#### Covariates and Predictors

The two cardiovascular risk predictors (FRS and
CVR~mimic~) were
mean-centered separately for males and females to
account for sex differences in cardiovascular risk
distributions;
covariates (age, education) were centered on grand
means pooled across sexes for consistency across
sex-stratified analyses; APOE ε4 carrier status
entered as a binary indicator. CVR~mimic~ entered as
pre-computed age-adjusted ζ scores from the MIMIC
model, z-standardized on the common sample. Both FRS
and CVR~mimic~ are observed covariates in the LGCM.
We regressed all covariates on both the latent
intercept and latent slope factors for each process
(HVR and cognitive) and fit the same models with FRS
and CVR~mimic~ for direct comparison of predictive
utility. Although HVR z-scores are age-normalized
via GAMLSS and CVR~mimic~ is age-orthogonal by
construction, baseline age was retained as a
covariate because cognitive composites are not
age-normalized and because normative models trained
on one cohort may not fully absorb age effects when
applied to a demographically different sample
[@sliwinski_evaluating_2010; @hoffman_persons_2009].

#### Model Fit and Inference

Missing data were handled via full information
maximum likelihood (FIML), which uses all available
observations without listwise deletion. We assessed
model fit via −2 log-likelihood (−2LL), Akaike
information criterion (AIC), and Bayesian information
criterion (BIC). For bootstrap inference, we used
case resampling with
n = `r format(lgcm_results$bootstrap$n_boot, big.mark = ",")`
iterations for all mediation paths (a, b, c',
indirect, total), reporting both percentile and
bias-corrected (BC) 95% confidence intervals. BC
intervals adjust for bias in the bootstrap
distribution and are recommended for products of
coefficients, which typically have skewed sampling
distributions [@efron_introduction_1994]. A path is
significant when its BC 95% CI excludes zero; this
is the primary inference method for all mediation
paths. We extracted individual factor scores via the
regression method (`mxFactorScores`) for
visualization of slope coupling.

#### Coupling Model

In the coupling model, a free covariance between HVR
linear slope and cognitive linear slope
(hvr~s~ ↔ cog~s~) is estimated to test whether
linear rates of hippocampal and cognitive decline
co-vary, reported as standardized correlations:

$$
r = \frac{\text{Cov}(\eta_{1,\text{hvr}},
  \eta_{1,\text{cog}})}{\sqrt{
  \text{Var}(\eta_{1,\text{hvr}}) \cdot
  \text{Var}(\eta_{1,\text{cog}})}}
$$

#### Mediation Model

In the mediation model, directional paths replace the
free slope covariance: the a-path (CVR → HVR slope)
tests whether cardiovascular risk predicts the rate
of hippocampal
decline; the b-path (HVR slope → cognitive slope)
captures hippocampal-cognitive coupling; the c'-path
(CVR → cognitive slope) captures the direct effect
bypassing hippocampal atrophy; and the indirect
effect (a × b) quantifies mediation through the
hippocampus. No free hvr~s~ ↔ cog~s~ covariance is
estimated in the mediation model because the
directional b-path already captures the slope
relationship; the coupling model uses a free
covariance without directional paths instead. A
significant total effect is not required for valid
mediation inference [@bollen_latent_2006].

### Sensitivity Analyses

#### LME Sensitivity

We conducted five sensitivity analyses within the
LME framework. First, for **HC z-score specificity**,
we repeated the 3-way interaction with total
hippocampal volume (HC z-score, normalized against
UKB norms) instead of HVR, testing whether
cardiovascular risk moderation effects are specific to the
hippocampus-to-ventricle ratio or generalize to
hippocampal volume alone. Second, for **attrition
analysis**, we compared baseline characteristics
between completers (≥ 4 visits) and dropouts
(< 4 visits) using t-tests (continuous) and χ² tests
(categorical), plus logistic regression predicting
dropout status from baseline characteristics. Third,
for **raw HVR and HC sensitivity**, we repeated the
3-way interaction using raw (un-normalized) HVR and
raw HC (with intracranial volume covariate) instead
of UKB z-scored values, to confirm findings are not
artifacts of the normative transformation. Fourth, for **EDT
moderation**, we used EDT as a moderator in a 4-way
interaction (CVR × HVR × EDT × YRS). Note that
EDT and YRS capture different
sources of variation: EDT
is a baseline moderator indexing disease stage, while
YRS is the within-person time metric capturing
change
over follow-up; their inclusion tests whether cardiovascular risk
effects on brain-cognition coupling vary by disease
stage. Fifth, for **EDT Early/Middle
sensitivity**, we restricted LME models to the first
two EDT tertiles, excluding late-stage participants
to assess stability of effects across
disease stages.

#### Amyloid-Negative CU Control

To assess whether the primary LME results are driven
by AD pathology rather than cardiovascular risk, we
repeated the 3-way interaction analysis in an
independent sample of amyloid-negative, baseline
cognitively unimpaired (A^--^/CU) participants
(N = `r aneg_lme.res$sample_info$n_subjects`). Models
were pooled across sex (with SEX as covariate) given
the smaller sample. We also fit a linear LGCM mediation
model (using years from baseline as the time metric)
in this sample;
results are reported in Appendix C.

## Software

All analyses were conducted in
R (`r R.version.string`) with
the following packages: lavaan
[@rosseel_lavaan_2012] for MIMIC model specification,
estimation (MLR + FIML), measurement invariance
testing, and factor score extraction; lme4
[@bates_fitting_2015] for LME growth models with
bobyqa optimizer and restricted maximum likelihood
(REML) estimation; OpenMx
[@neale_openmx_2016] for LGCM parallel process and
mediation models with sequential least squares
programming (SLSQP) optimizer (convergence
criterion: gradient < $10^{-5}$); progmod
[@raket_statistical_2020] for EDT computation; and
CVrisk [@castro_cvrisk_2026] for FRS computation (via
ADSP-PHC).

---

# Results

## Sample Characteristics

The LME cohort comprised
`r n_subjects` amyloid-positive participants
(`r sum(baseline.dt$SEX == "Male")` males,
`r sum(baseline.dt$SEX == "Female")` females)
with complete data for both FRS and CVR~mimic~
after common-sample filtering. Mean age was
`r sprintf("%.1f", age_mean)` ±
`r sprintf("%.1f", age_sd)` years
(range
`r sprintf("%.0f", age_min)`–`r sprintf("%.0f", age_max)`).
The diagnostic composition included `r n_cu`
cognitively unimpaired
(`r sprintf("%.0f", 100 * n_cu / n_subjects)`%),
`r n_mci` MCI
(`r sprintf("%.0f", 100 * n_mci / n_subjects)`%),
and `r n_ad` AD dementia
(`r sprintf("%.0f", 100 * n_ad / n_subjects)`%).
Participants had a median of
`r median_visits` visits
(range `r range_visits[1]`–`r range_visits[2]`)
with mean follow-up of
`r sprintf("%.1f", fu_mean)` years
(max `r sprintf("%.1f", fu_max)`).
Education averaged
`r sprintf("%.1f", educ_mean)` ±
`r sprintf("%.1f", educ_sd)` years, and
`r sprintf("%.0f", pct_apoe4)`% were APOE ε4
carriers. Baseline FRS was
`r sprintf("%.1f", frs_mean_raw)` ±
`r sprintf("%.1f", frs_sd_raw)` (raw);
CVR~mimic~ ζ was
`r sprintf("%.2f", cvr_mean)` ±
`r sprintf("%.2f", cvr_sd)` (age-adjusted
latent scores; both FRS and CVR~mimic~
z-standardized on the common sample before
analysis). The LGCM cohort included
`r n_lgcm_total` participants
(`r lgcm_n.lst$Male$n` males,
`r lgcm_n.lst$Female$n` females) with ≥2
follow-ups and available EDT values.
Mediation bootstrap samples comprised
`r med_n.lst$Male` males and
`r med_n.lst$Female` females (complete cases
required for case-resampling). Participant inclusion
and exclusion flow is shown in
Figure \ref{fig-consort}; baseline
demographics by sex appear in @tbl-demographics.

```{r tbl-demographics}
#| tbl-cap: "**Baseline cohort characteristics.** ADNI analysis cohort compared with the UK Biobank normative reference sample. Values are mean (SD) for continuous variables, % for categorical. HC = bilateral hippocampal volume; HVR = hippocampus-to-ventricle ratio; FRS = Framingham Risk Score; APOE ε4+ = carriers of at least one ε4 allele. Z-scores derived from UK Biobank GAMLSS normative models."

require_output(
  create_demographics_table(
    cohort_for_tbl.dt,
    "Baseline Demographic and Clinical Characteristics",
    normative_stats = norm_stats,
    ukb_raw = ukb_raw_stats
  ),
  "tbl-demographics"
)
```


We validated UKB-derived z-scores against an
amyloid-negative, cognitively unimpaired ADNI
subsample within the UKB age range. HVR z-scores
showed a systematic negative offset consistent
with the older ADNI age distribution; HC z-scores
were somewhat closer to expected values
(Appendix A, Figure A1).

## Cardiovascular Risk Characterization

### FRS Properties

In the ADNI cohort, FRS was strongly correlated with
age (r = `r sprintf("%.3f", frs_age_r)`). The FRS-age
partial correlation (controlling for other
cardiovascular risk factors)
remained substantial
(r = `r sprintf("%.3f", frs_partial_r)`),
confirming that age dominates FRS in older adults.
Variance decomposition showed age alone R² =
`r sprintf("%.3f", frs_var_decomp$r2_age)`,
cardiovascular risk factors alone R² =
`r sprintf("%.3f", frs_var_decomp$r2_cvrf)`, and
combined R² =
`r sprintf("%.3f", frs_var_decomp$r2_full)`,
indicating true cardiovascular risk factors
contribute modestly beyond
age. A total of
`r sprintf("%.1f", frs_ceiling_all)`% of
participants were at the 30% risk cap
(`r sprintf("%.1f", frs_ceiling_male)`% males,
`r sprintf("%.1f", frs_ceiling_female)`% females),
and `r sprintf("%.1f", frs_above20_all)`% scored
above 20, limiting statistical sensitivity even
below the formal ceiling.

### CVR~mimic~ Validation

The MIMIC model showed good fit: CFI =
`r sprintf("%.3f", mimic_fit[["cfi.robust"]])`,
TLI =
`r sprintf("%.3f", mimic_fit[["tli.robust"]])`,
RMSEA =
`r sprintf("%.3f", mimic_fit[["rmsea.robust"]])`,
SRMR =
`r sprintf("%.3f", mimic_fit[["srmr"]])`. The
structural path showed γ =
`r sprintf("%.3f", mimic_gamma)` with
R² = `r sprintf("%.3f", mimic_r2)`, indicating age
explained only
`r sprintf("%.1f", mimic_r2 * 100)`% of
cardiovascular risk variance and confirming CVR~mimic~ is largely
age-free. Factor loadings appear in
@tbl-mimic-loadings; visual validation in
@fig-cvr-validation.

```{r tbl-mimic-loadings}
#| tbl-cap: "**CVR~mimic~ model factor loadings.** λ = unstandardized factor loading; Std. λ = standardized loading; SE = standard error."

stopifnot(!is.null(cvr_model.res))
require_output(
  create_mimic_loadings_table(
    cvr_model.res,
    "MIMIC Model Factor Loadings"
  ),
  "tbl-mimic-loadings"
)
```

After age-adjustment and z-standardization,
CVR~mimic~ showed near-zero correlation with age
(r = `r sprintf("%.3f", cvr_age_cor)`, by
construction), moderate correlation with FRS
(r = `r sprintf("%.2f", cvr_frs_cor)`, reflecting
shared CV indicators), and approximate standard
normal distribution (mean =
`r sprintf("%.2f", cvr_mean)`, SD =
`r sprintf("%.2f", cvr_sd)`).

### Measurement Invariance

Measurement invariance of the CVR~mimic~ model was tested across sex in
four steps (Appendix B, Table B1). Configural
invariance was supported: CFI =
`r sprintf("%.3f", mi_cfg_fit[["CFI"]])`,
RMSEA =
`r sprintf("%.3f", mi_cfg_fit[["RMSEA"]])`.

```{r mi-inline-steps}
#| include: false

mi_comp <- list(
  metric = "configural_vs_metric",
  scalar = "metric_vs_scalar",
  strict = "scalar_vs_strict"
)
mi_labels <- c(
  metric = "Metric (equal loadings)",
  scalar = "Scalar (equal intercepts)",
  strict = "Strict (equal residuals)"
)
mi_step_parts <- vapply(
  names(mi_comp), function(step) {
    ckey <- mi_comp[[step]]
    delta <-
      cvr_mi.res$comparisons_delta[[ckey]]
    lrt <-
      cvr_mi.res$comparisons_lrt[[ckey]]
    if (is.null(delta) && is.null(lrt)) {
      return("")
    }
    d_cfi <- if (!is.null(delta$d_cfi)) {
      sprintf("%.3f", delta$d_cfi)
    } else "NA"
    d_rmsea <- if (!is.null(delta$d_rmsea)) {
      sprintf("%.3f", delta$d_rmsea)
    } else "NA"
    chi_p <- if (!is.null(lrt$p)) {
      fmt_p_inline(lrt$p)
    } else "NA"
    sprintf(
      "%s: ΔCFI = %s, ΔRMSEA = %s (%s)",
      mi_labels[step], d_cfi, d_rmsea, chi_p
    )
  }, character(1)
)
mi_steps_text <- paste(
  mi_step_parts[mi_step_parts != ""],
  collapse = "; "
)
scalar_delta <- cvr_mi.res$comparisons_delta[[
  "metric_vs_scalar"
]]
scalar_fail <- !is.null(scalar_delta) &&
  !is.null(scalar_delta$d_cfi) &&
  abs(scalar_delta$d_cfi) > 0.01
mi_verdict_text <- if (scalar_fail) {
  paste0(
    "Scalar invariance was not achieved",
    " (|ΔCFI| > 0.01), indicating",
    " intercepts differ across sex.",
    " Metric invariance (equal loadings)",
    " was supported \u2014 sufficient for",
    " sex-stratified analyses."
  )
} else {
  sprintf(
    "%s invariance was achieved.",
    mi_verdict
  )
}
```

`r mi_steps_text`. `r mi_verdict_text`

```{r fig-cvr-validation}
#| fig-cap: "**Cardiovascular risk characterization.** (A) FRS distribution by sex; dashed red line marks the 30\\% risk ceiling. (B) FRS vs Age by sex showing strong positive correlation driven by age-weighting in the FRS formula. (C) CVR$_{\\text{mimic}}$ vs Age showing near-zero correlation, confirming successful age adjustment; dashed line marks zero."
#| fig-width: 6.5
#| fig-height: 4.5

stopifnot(!is.null(cvr_scores), !is.null(cohort.dt))
  # Baseline with raw FRS percentage
  frs_raw_col <- get_parameter("frs", "column")
  cvrf_raw.dt <- fread(
    get_data_path("raw", "adsp_phc_cvrf"),
    select = c("PTID", frs_raw_col)
  )
  setnames(cvrf_raw.dt, frs_raw_col, "FRS_pct")

  baseline_merged.dt <- cohort.dt[, .(
    AGE = AGE[1], SEX = SEX[1]
  ), by = .(RID, PTID)]
  baseline_merged.dt <- merge(
    baseline_merged.dt,
    cvrf_raw.dt, by = "PTID"
  )
  merged_plot.dt <- merge(
    baseline_merged.dt,
    cvr_scores[, .(RID, CVR_mimic)],
    by = "RID"
  )

  p_frs_sex <- plot_frs_ceiling_by_sex(
    merged_plot.dt
  )
  p_frs_age <- plot_frs_validation(
    merged_plot.dt, frs_age_r = frs_age_r
  )
  p_cvr_age <- plot_cvr_validation(
    merged_plot.dt
  )

  p_frs_sex / (p_frs_age + p_cvr_age) +
    plot_layout(heights = c(1, 1)) +
    plot_annotation(
      tag_levels = "A",
      theme = theme(
        plot.tag = element_text(
          margin = margin(r = 5, b = 2)
        )
      )
    )
```


EDT linearity tests and SE-constrained LGCM
simulation validation are reported in Appendix B
(see Appendix B, Table B2).

## LME Results: FRS vs CVR~mimic~ 3-Way Interaction

### 3-Way Interaction (FRS vs CVR~mimic~)

```{r fig-lme-forest}
#| fig-cap: "**LME 3-way interaction forest plot.** FRS vs CVR$_{\\text{mimic}}$ effects (YRS $\\times$ CVR $\\times$ HVR z-score $\\rightarrow$ Cognition) by sex and cognitive domain, with an amyloid-negative cognitively unimpaired (A$^{-}$/CU) control panel (pooled sex). Filled points indicate p < 0.05; white-filled points indicate p $\\geq$ 0.05. Error bars show 95\\% CI."
#| fig-width: 6.5
#| fig-height: 3.0

plot_lme_forest_comparison(
  frs_hvr, cvr_hvr, aneg_lme.res
)
```

Full covariate effects (Age, Education, APOE ε4,
cardiovascular risk main effect, HVR z-score) are
reported in
Appendix B (Tables B3 and B4).

```{r lme-inline-results}
#| include: false

stopifnot(!is.null(frs_hvr), !is.null(cvr_hvr))
get_fdr_p <- function(cvr_meas, sex, domain) {
  row <- lme_summary.dt[
    Analysis == "Primary_Stratified" &
    CVR_Measure == cvr_meas &
    Sex == sex & Domain == domain
  ]
  if (nrow(row) > 0 &&
      !is.null(row$p_fdr)) {
    row$p_fdr[1]
  } else NA_real_
}
get_lme_result <- function(
    res, sex, domain, cvr_meas) {
  r <- res$stratified[[sex]][[domain]]$interaction
  if (!is.null(r)) {
    list(
      beta = r$beta, p = r$p_value,
      p_fdr = get_fdr_p(cvr_meas, sex, domain)
    )
  } else NULL
}
fmt_p_fdr <- function(p, p_fdr) {
  base <- fmt_p_inline(p)
  if (!is.na(p_fdr) && p < 0.05) {
    fdr_str <- if (p_fdr < 0.001) {
      "*p*~FDR~ < 0.001"
    } else {
      sprintf("*p*~FDR~ = %.3f", p_fdr)
    }
    fdr_tag <- if (p_fdr < 0.05) {
      sprintf(" [%s]", fdr_str)
    } else {
      sprintf(" [%s, n.s.]", fdr_str)
    }
    paste0(base, fdr_tag)
  } else base
}

frs_all <- list()
cvr_all <- list()
for (sx in c("Male", "Female")) {
  for (d in c("MEM", "LAN", "EXF")) {
    frs_all[[paste(sx, d)]] <-
      get_lme_result(frs_hvr, sx, d, "FRS")
    cvr_all[[paste(sx, d)]] <-
      get_lme_result(
        cvr_hvr, sx, d, "CVR_mimic"
      )
  }
}
frs_n_sig <- sum(sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
))
cvr_n_sig <- sum(sapply(
  cvr_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
))
frs_betas <- abs(sapply(
  frs_all, function(r) {
    if (!is.null(r)) r$beta else NA
  }
))
cvr_betas <- abs(sapply(
  cvr_all, function(r) {
    if (!is.null(r)) r$beta else NA
  }
))
beta_ratio <- mean(
  frs_betas, na.rm = TRUE
) / mean(cvr_betas, na.rm = TRUE)

frs_sig_keys <- names(frs_all)[sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
)]
fmt_doms <- function(keys) {
  ds <- sub("^(Male|Female) ", "", keys)
  paste(DOMAIN_LABELS[ds], collapse = ", ")
}
frs_sig_male <- grep(
  "^Male", frs_sig_keys, value = TRUE
)
frs_sig_female <- grep(
  "^Female", frs_sig_keys, value = TRUE
)
lme_male_detail <- paste0(
  fmt_doms(frs_sig_male)
)
lme_female_detail <- paste0(
  fmt_doms(frs_sig_female)
)

# Per-sex, per-domain results for inline use
lme.lst <- list()
for (sx in c("m", "f")) {
  sx_full <- ifelse(sx == "m", "Male", "Female")
  cvr_ps <- c()
  for (d in c("MEM", "LAN", "EXF")) {
    rf <- get_lme_result(
      frs_hvr, sx_full, d, "FRS"
    )
    rc <- get_lme_result(
      cvr_hvr, sx_full, d, "CVR_mimic"
    )
    pfx <- paste0(sx, "_", tolower(d))
    lme.lst[[paste0(pfx, "_frs_b")]] <-
      sprintf("%.4f", rf$beta)
    lme.lst[[paste0(pfx, "_frs_p")]] <-
      fmt_p_fdr(rf$p, rf$p_fdr)
    lme.lst[[paste0(pfx, "_cvr_b")]] <-
      sprintf("%.4f", rc$beta)
    lme.lst[[paste0(pfx, "_cvr_p")]] <-
      fmt_p_fdr(rc$p, rc$p_fdr)
    cvr_ps <- c(cvr_ps, rc$p)
  }
  lme.lst[[paste0(sx, "_cvr_min_p")]] <-
    sprintf("%.3f", min(cvr_ps))
}
list2env(lme.lst, envir = environment())

# Worked example
frs_sig <- frs_all[sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
)]
worked_text <- ""
if (length(frs_sig) > 0) {
  max_r <- frs_sig[[which.max(
    abs(sapply(frs_sig, `[[`, "beta"))
  )]]
  max_nm <- names(frs_sig)[which.max(
    abs(sapply(frs_sig, `[[`, "beta"))
  )]
  parts <- strsplit(max_nm, " ")[[1]]
  display_nm <- paste(
    parts[1], DOMAIN_LABELS[parts[2]]
  )
  worked_text <- sprintf(
    paste0(
      "The largest significant FRS 3-way",
      " effect (%s, β = %.4f) implies",
      " that over 5 years, a 1-SD increase",
      " in FRS is associated with a",
      " %.4f-unit cumulative divergence",
      " in the HVR-cognition association."
    ),
    display_nm, max_r$beta,
    max_r$beta * 5
  )
}
```

**Key pattern.** FRS produced significant 3-way
interactions (YRS × CVR × HVR) in
males
(`r lme_male_detail`) and females
(`r lme_female_detail`) —
`r n_of_total.fn(frs_n_sig, 6L)` models, all
surviving FDR correction. CVR~mimic~ showed only
`r n_of_total.fn(cvr_n_sig, 6L)` significant
interactions
(`r n_of_total.fn(fdr_summary.lst$cvr_n_fdr, 6L)`
surviving FDR), consistent with FRS effects being
primarily driven by age-weighting rather than true
cardiovascular risk (see @fig-lme-forest).

**Males.** FRS × HVR × YRS was significant for
Memory (β = `r m_mem_frs_b`, `r m_mem_frs_p`),
Language (β = `r m_lan_frs_b`, `r m_lan_frs_p`),
and Executive Function
(β = `r m_exf_frs_b`, `r m_exf_frs_p`).
CVR~mimic~ showed no significant interactions
(all *p* ≥ `r m_cvr_min_p`).

**Females.** FRS effects were also significant
across all three domains, with larger magnitudes:
Memory (β = `r f_mem_frs_b`, `r f_mem_frs_p`),
Language (β = `r f_lan_frs_b`, `r f_lan_frs_p`),
and Executive Function
(β = `r f_exf_frs_b`, `r f_exf_frs_p`).
CVR~mimic~ reached significance only for Executive
Function
(β = `r f_exf_cvr_b`, `r f_exf_cvr_p`).

FRS β values were on average
`r sprintf("%.1f", beta_ratio)`-fold larger in
absolute magnitude than CVR~mimic~ β values,
consistent with age-weighting amplifying apparent
FRS effects. `r worked_text`

```{r lme-fdr-note}
#| include: false

stopifnot(!is.null(fdr_summary.lst))
frs_changed <- fdr_summary.lst$frs_n_sig !=
  fdr_summary.lst$frs_n_fdr
cvr_changed <- fdr_summary.lst$cvr_n_sig !=
  fdr_summary.lst$cvr_n_fdr
fdr_note_text <- if (
  !frs_changed && !cvr_changed
) {
  paste0(
    "FDR correction (Benjamini-Hochberg)",
    " did not change any conclusions for",
    " either cardiovascular risk measure."
  )
} else if (!frs_changed && cvr_changed) {
  sprintf(
    paste0(
      "FDR correction did not change FRS",
      " conclusions. For CVR~mimic~, %d of",
      " %d uncorrected effects survived FDR."
    ),
    fdr_summary.lst$cvr_n_fdr,
    fdr_summary.lst$cvr_n_sig
  )
} else {
  sprintf(
    paste0(
      "After FDR correction, %s FRS and",
      " %s CVR~mimic~ uncorrected effects",
      " remained significant."
    ),
    n_of_total.fn(
      fdr_summary.lst$frs_n_fdr,
      fdr_summary.lst$frs_n_sig
    ),
    n_of_total.fn(
      fdr_summary.lst$cvr_n_fdr,
      fdr_summary.lst$cvr_n_sig
    )
  )
}
```

`r fdr_note_text`

### Formal Coefficient Comparison (LME)

To formally compare FRS and CVR~mimic~ 3-way
interaction coefficients, we computed clustered
bootstrap Δβ with bias-corrected 95% CIs.

```{r lme-comparison-boot}
#| include: false

lme_cmp <- lme_hvr_z.res$comparison
stopifnot(
  !is.null(lme_cmp), length(lme_cmp) > 0
)
lme_db.lst <- list()
for (sx in c("m", "f")) {
  sx_full <- ifelse(
    sx == "m", "Male", "Female"
  )
  for (d in c("MEM", "LAN", "EXF")) {
    k <- paste(sx_full, d, sep = "_")
    r <- lme_cmp[[k]]
    pfx <- paste0(sx, "_", tolower(d))
    lme_db.lst[[paste0(pfx, "_db")]] <-
      sprintf("%.4f", r$delta_beta)
    lme_db.lst[[paste0(pfx, "_ci")]] <-
      sprintf(
        "[%.4f, %.4f]",
        r$ci_lower, r$ci_upper
      )
  }
}
list2env(lme_db.lst, envir = environment())
```

**Males.** $\Delta\beta$ values were negative across all three
domains, reaching significance for Language
($\Delta\beta$ = `r m_lan_db`, 95% CI `r m_lan_ci`) but not
Memory ($\Delta\beta$ = `r m_mem_db`, 95% CI `r m_mem_ci`) or
Executive Function
($\Delta\beta$ = `r m_exf_db`, 95% CI `r m_exf_ci`).

**Females.** $\Delta\beta$ values were likewise negative.
Memory reached significance
($\Delta\beta$ = `r f_mem_db`, 95% CI `r f_mem_ci`), whereas
Language
($\Delta\beta$ = `r f_lan_db`, 95% CI `r f_lan_ci`) and
Executive Function
($\Delta\beta$ = `r f_exf_db`, 95% CI `r f_exf_ci`)
did not.

The A^--^/CU control analysis reversed this pattern:
CVR~mimic~ reached FDR significance while FRS did
not (@fig-lme-forest, panel C; full results in
Appendix C).

## LGCM Results

### Growth Parameters

LGCM models used EDT as the time metric with
individually-varying time scores, fit separately per
sex and cognitive domain.

```{r lgcm-growth-inline}
#| include: false

stopifnot(!is.null(lgcm_parallel$linear_results))
domains <- c("MEM", "LAN", "EXF")

lgcm_growth_text.lst <- list()
for (sx in c("Male", "Female")) {
  hvr_slopes <- c()
  cog_slopes <- c()
  for (d in domains) {
    k <- paste0(sx, "_", d)
    lr <- lgcm_parallel$linear_results[[k]]
    if (is.null(lr)) next
    s <- lr$summary
    if (is.null(s)) next
    hvr_s <- s[
      s$label == "hvr_slope_mean" |
      grepl("hvr.*slope.*mean", s$label,
        ignore.case = TRUE), ]
    cog_s <- s[
      s$label == "cog_slope_mean" |
      grepl("cog.*slope.*mean", s$label,
        ignore.case = TRUE), ]
    if (nrow(hvr_s) > 0) {
      hvr_slopes <- c(
        hvr_slopes, hvr_s$Estimate[1]
      )
    }
    if (nrow(cog_s) > 0) {
      cog_slopes <- c(
        cog_slopes, cog_s$Estimate[1]
      )
    }
  }
  lgcm_growth_text.lst[[sx]] <- if (
    length(hvr_slopes) > 0 &&
    length(cog_slopes) > 0
  ) {
    sprintf(
      paste0(
        "In %ss, mean HVR slopes were negative",
        " across all domains (range: %.4f to",
        " %.4f); cognitive slopes were also",
        " negative (range: %.4f to %.4f),",
        " confirming progressive decline in",
        " both processes."
      ),
      tolower(sx),
      min(hvr_slopes), max(hvr_slopes),
      min(cog_slopes), max(cog_slopes)
    )
  } else ""
}
```

`r lgcm_growth_text.lst$Male`
`r lgcm_growth_text.lst$Female`
Model fit appears in @tbl-mediation-fit; standard
fit indices (CFI, TLI, RMSEA) are not computable
with individually-varying time scores, so adequacy
was assessed via −2LL, AIC, BIC, and nested LRTs.

### Slope Coupling {#sec-slope-coupling}

LGCM parallel process models tested whether HVR and
cognitive linear decline rates co-vary (latent slope
coupling, hvr~s~ ↔ cog~s~) along the EDT disease
timeline.

```{r coupling-inline}
#| include: false

stopifnot(!is.null(lgcm_parallel$linear_results))
domains <- c("MEM", "LAN", "EXF")
domain_labels <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

coupling_text.lst <- list()
for (sx in c("Male", "Female")) {
  cors <- c()
  cor_names <- c()
  ps <- c()
  for (d in domains) {
    k <- paste0(sx, "_", d)
    lr <- lgcm_parallel$linear_results[[k]]
    if (!is.null(lr$coupling) &&
        !is.na(lr$coupling$p)) {
      prm <- lr$params
      v_h <- prm[
        name == "var_hvr_s", Estimate
      ]
      v_c <- prm[
        name == "var_cog_s", Estimate
      ]
      r_val <- lr$coupling$cov /
        sqrt(v_h * v_c)
      cors <- c(cors, r_val)
      cor_names <- c(
        cor_names, domain_labels[d]
      )
      ps <- c(ps, lr$coupling$p)
    }
  }
  coupling_text.lst[[sx]] <- if (
    length(cors) > 0
  ) {
    max_d <- cor_names[which.max(cors)]
    min_d <- cor_names[which.min(cors)]
    max_p <- max(ps)
    p_bound <- if (max_p < 0.001) {
      "all *p* < 0.001"
    } else {
      sprintf("max %s", fmt_p_inline(max_p))
    }
    sprintf(
      paste0(
        "In %ss, linear slope correlations",
        " (*r*) ranged from %.3f (%s) to",
        " %.3f (%s; %s)."
      ),
      tolower(sx),
      min(cors), min_d,
      max(cors), max_d, p_bound
    )
  } else ""
}
```

`r coupling_text.lst$Male`
`r coupling_text.lst$Female`
In males, faster hippocampal decline was
significantly associated with steeper cognitive
decline across all three domains; in females, the
direction was consistent but only Executive Function
reached significance, likely reflecting lower
statistical power in the smaller female subsample.
This coupling supports the b-path assumption in the
mediation framework: HVR decline predicts cognitive
decline. Latent factor scores from the parallel
process model confirm this pattern (see
@fig-coupling).

```{r fig-coupling}
#| fig-cap: "**HVR–cognitive latent slope coupling.** Females in red, Males in blue throughout. A. Individual trajectories (random 20% subsample) with population mean curves (bold; including fixed quadratic curvature where selected by LRT). 2×2 facets: HVR, Memory, Language, Executive. B. Latent slope coupling: latent cognitive decline rate (x) vs latent HVR decline rate (y) from factor scores (positive = faster decline); dashed lines show linear trends per sex; grey line shows 1:1 reference. Correlation estimates from the SEM (* p < .05, ** p < .01, *** p < .001)."
#| fig-width: 6.5
#| fig-height: 7.5

lgcm_m_input <- read_rds_safe(here(
  "data/derivatives/lgcm/lgcm_input_male.rds"
))
lgcm_f_input <- read_rds_safe(here(
  "data/derivatives/lgcm/lgcm_input_female.rds"
))
stopifnot(
  !is.null(lgcm_m_input),
  !is.null(lgcm_f_input),
  !is.null(lgcm_parallel)
)
plot_coupling_figure(
  lgcm_parallel,
  lgcm_m_input, lgcm_f_input
)
```

### Mediation: FRS vs CVR~mimic~

LGCM mediation models tested whether cardiovascular
risk effects on
cognitive decline operate through hippocampal
atrophy, comparing FRS and CVR~mimic~ as predictors
(see @fig-mediation).

```{r tbl-mediation-fit}
#| tbl-cap: "**LGCM mediation model fit indices.** Sex-stratified. -2LL = -2 log-likelihood; AIC = Akaike information criterion; BIC = Bayesian information criterion."

require_output(
  create_mediation_fit_table(
    lgcm_results,
    title = "Mediation Model Fit"
  ),
  "tbl-mediation-fit"
)
```

Mediation path diagrams showing a-, b-, c'-path
estimates and bootstrap CIs appear in
@fig-mediation.


```{r fig-mediation}
#| dev: png
#| fig-cap: "**LGCM mediation path diagrams.** Rows: Memory, Language, Executive Function. Columns: FRS, CVR$_{\\text{mimic}}$. Each panel shows Male (blue) and Female (red) path estimates. Bold = significant (bootstrap 95% CI excludes zero); faded = non-significant. Star (\\*) and bold indicate CI excludes zero. 95% bootstrap CIs shown for all paths."
#| fig-width: 6.5
#| fig-height: 8.0
#| fig-dpi: 300

panels <- build_combined_med_panels(
  lgcm_med_male, lgcm_med_female
)
wrap_plots(panels, ncol = 2) +
  plot_annotation(tag_levels = "A")
```

```{r lgcm-inline-results}
#| include: false

stopifnot(!is.null(lgcm_med_male))
stopifnot(!is.null(lgcm_med_female))

domains <- c("MEM", "LAN", "EXF")

frs_a_sig <- sum(sapply(
  c("Male", "Female"), function(sx) {
    med <- if (sx == "Male") {
      lgcm_med_male
    } else lgcm_med_female
    sapply(domains, function(d) {
      r <- med$results[[
        paste0(d, "_FRS")
      ]]$a
      !is.null(r) && ci_sig_check.fn(r)
    })
  }
))
cvr_a_sig <- sum(sapply(
  c("Male", "Female"), function(sx) {
    med <- if (sx == "Male") {
      lgcm_med_male
    } else lgcm_med_female
    sapply(domains, function(d) {
      r <- med$results[[
        paste0(d, "_CVR")
      ]]$a
      !is.null(r) && ci_sig_check.fn(r)
    })
  }
))
cvr_a_null <- 6L - cvr_a_sig

all_ind_sig <- 0L
all_ind_total <- 0L
ind_sig_labels <- c()
for (sx in c("Male", "Female")) {
  med <- if (sx == "Male") {
    lgcm_med_male
  } else lgcm_med_female
  for (d in domains) {
    for (pred in c("FRS", "CVR")) {
      k <- paste0(d, "_", pred)
      r <- med$results[[k]]
      all_ind_total <- all_ind_total + 1L
      if (!is.null(r$indirect) &&
          ci_sig_check.fn(r$indirect)) {
        all_ind_sig <- all_ind_sig + 1L
        ind_sig_labels <- c(
          ind_sig_labels,
          sprintf(
            "%s %s (%s)",
            sx, DOMAIN_LABELS[d], pred
          )
        )
      }
    }
  }
}
ind_detail <- if (length(ind_sig_labels) > 0) {
  paste0(
    " — ",
    paste(ind_sig_labels, collapse = ", ")
  )
} else ""

# Per-sex mediation scalars for inline use
med_inline.lst <- list()
for (sx in c("m", "f")) {
  sx_full <- ifelse(
    sx == "m", "Male", "Female"
  )
  med <- if (sx == "m") {
    lgcm_med_male
  } else lgcm_med_female
  frs_a.v <- c()
  frs_b.v <- c()
  frs_a_nsig <- 0L
  cvr_a_nsig <- 0L
  frs_b_nsig <- 0L
  frs_ind_nsig <- 0L
  cvr_ind_nsig <- 0L
  cvr_max_abs_a <- 0
  frs_pm.v <- c()
  for (d in domains) {
    rf <- med$results[[paste0(d, "_FRS")]]
    rc <- med$results[[paste0(d, "_CVR")]]
    if (!is.null(rf$a)) {
      frs_a.v <- c(frs_a.v, rf$a$est)
      if (ci_sig_check.fn(rf$a))
        frs_a_nsig <- frs_a_nsig + 1L
    }
    if (!is.null(rc$a)) {
      cvr_max_abs_a <- max(
        cvr_max_abs_a, abs(rc$a$est)
      )
      if (ci_sig_check.fn(rc$a))
        cvr_a_nsig <- cvr_a_nsig + 1L
    }
    if (!is.null(rf$b)) {
      frs_b.v <- c(frs_b.v, rf$b$est)
      if (ci_sig_check.fn(rf$b))
        frs_b_nsig <- frs_b_nsig + 1L
    }
    if (!is.null(rf$indirect) &&
        ci_sig_check.fn(rf$indirect))
      frs_ind_nsig <- frs_ind_nsig + 1L
    if (!is.null(rc$indirect) &&
        ci_sig_check.fn(rc$indirect))
      cvr_ind_nsig <- cvr_ind_nsig + 1L
    pm <- rf$prop_mediated
    if (!is.null(pm) && !is.na(pm))
      frs_pm.v <- c(frs_pm.v, pm)
  }
  pfx <- sx
  med_inline.lst[[paste0(pfx, "_frs_a_range")]] <-
    sprintf("%.4f to %.4f",
      min(frs_a.v), max(frs_a.v))
  med_inline.lst[[paste0(pfx, "_frs_a_nsig")]] <-
    n_of_total.fn(frs_a_nsig, 3L)
  med_inline.lst[[paste0(pfx, "_cvr_a_nsig")]] <-
    n_of_total.fn(cvr_a_nsig, 3L)
  med_inline.lst[[paste0(pfx, "_cvr_max_a")]] <-
    sprintf("%.2f", cvr_max_abs_a + 0.005)
  med_inline.lst[[paste0(pfx, "_frs_b_range")]] <-
    sprintf("%.3f to %.3f",
      min(frs_b.v), max(frs_b.v))
  med_inline.lst[[paste0(pfx, "_frs_b_nsig")]] <-
    n_of_total.fn(frs_b_nsig, 3L)
  med_inline.lst[[paste0(pfx, "_frs_ind")]] <-
    n_of_total.fn(frs_ind_nsig, 3L)
  med_inline.lst[[paste0(pfx, "_cvr_ind")]] <-
    n_of_total.fn(cvr_ind_nsig, 3L)
  # Check CVR c' significance for each domain
  cvr_cp_sig.v <- character(0)
  for (d in domains) {
    rc <- med$results[[paste0(d, "_CVR")]]
    if (!is.null(rc$cprime) &&
        ci_sig_check.fn(rc$cprime))
      cvr_cp_sig.v <- c(cvr_cp_sig.v, d)
  }
  med_inline.lst[[paste0(
    pfx, "_cvr_cp_sig_doms"
  )]] <- cvr_cp_sig.v
  frs_pm_capped.v <- pmin(frs_pm.v, 1)
  med_inline.lst[[paste0(pfx, "_frs_pm")]] <-
    if (length(frs_pm_capped.v) > 0) {
      lo <- min(frs_pm_capped.v) * 100
      hi <- max(frs_pm_capped.v) * 100
      if (hi >= 100) {
        sprintf("%.0f%%–full mediation", lo)
      } else {
        sprintf("%.0f%%–%.0f%%", lo, hi)
      }
    } else "NA"
}
list2env(
  med_inline.lst, envir = environment()
)
```

**FRS vs CVR~mimic~ across all 12 models.** The
a-path (CVR → HVR slope) showed a clear predictor
dissociation: FRS significantly predicted
hippocampal decline in
`r n_of_total.fn(frs_a_sig, 6L)` models (bootstrap
95% CIs excluding zero), whereas CVR~mimic~ did not
in any — indicating that FRS "effects" on
hippocampal decline are largely consistent with
age-weighting. By contrast, the b-path
(HVR slope → cognitive slope) was significant for
`r n_of_total.fn(abs_stats$frs_b_sig, 6L)` FRS and
`r n_of_total.fn(abs_stats$cvr_b_sig, 6L)`
CVR~mimic~ models, confirming that hippocampal-
cognitive coupling persists regardless of predictor.
The FRS indirect effect (a × b) was significant in
`r n_of_total.fn(abs_stats$frs_ind_sig, 6L)` models;
CVR~mimic~ indirect effects in
`r n_of_total.fn(abs_stats$cvr_ind_sig, 6L)`.

**Males.** FRS predicted hippocampal atrophy across
`r m_frs_a_nsig` domains (a-path β range
`r m_frs_a_range`, all bootstrap 95% CIs excluding
zero); CVR~mimic~ did not
(all |β| < `r m_cvr_max_a`). Hippocampal decline
in turn predicted cognitive decline (b-path β range
`r m_frs_b_range`, `r m_frs_b_nsig` CIs excluding
zero), and the FRS indirect effect was significant
for `r m_frs_ind` domains (proportion mediated
`r m_frs_pm`). No CVR~mimic~ indirect effects
reached significance.

**Females.** The same a-path dissociation held: FRS
was significant across `r f_frs_a_nsig` domains
(β range `r f_frs_a_range`), CVR~mimic~ was not
(all |β| < `r f_cvr_max_a`). B-path estimates were
comparable to males (β range `r f_frs_b_range`,
`r f_frs_b_nsig` CIs excluding zero), and FRS
indirect effects reached significance in
`r f_frs_ind` domains (proportion mediated
`r f_frs_pm`); CVR~mimic~ indirect effects
reached significance in `r f_cvr_ind`.
`r if (length(f_cvr_cp_sig_doms) > 0) paste0("The CVR~mimic~ ", paste(DOMAIN_LABELS[f_cvr_cp_sig_doms], collapse = " and "), " model", ifelse(length(f_cvr_cp_sig_doms) > 1, "s showed significant direct effects", " showed a significant direct effect"), " (c'), suggesting CVR~mimic~ may influence cognitive decline through a non-hippocampal pathway.") else "No CVR~mimic~ direct effects (c') reached significance."`

Complete path estimates, bootstrap CIs, and direct
(c') effects appear in @fig-mediation.

### Formal a-Path Comparison (LGCM)

Paired bootstrap with shared-seed resampling was
used to formally compare FRS and CVR~mimic~ a-path
coefficients (individual estimates shown in
@fig-mediation above).

```{r lgcm-comparison-boot}
#| include: false

lgcm_cmp <- lgcm_results$comparisons
stopifnot(
  !is.null(lgcm_cmp), length(lgcm_cmp) > 0
)
lgcm_da.lst <- list()
for (sx in c("m", "f")) {
  sx_full <- ifelse(
    sx == "m", "Male", "Female"
  )
  for (d in c("MEM", "LAN", "EXF")) {
    k <- paste(sx_full, d, sep = "_")
    r <- lgcm_cmp[[k]]
    pfx <- paste0(sx, "_", tolower(d))
    lgcm_da.lst[[paste0(pfx, "_da")]] <-
      sprintf("%.5f", r$delta_a_est)
    lgcm_da.lst[[paste0(pfx, "_da_ci")]] <-
      sprintf(
        "[%.5f, %.5f]",
        r$delta_a_ci_lower,
        r$delta_a_ci_upper
      )
    lgcm_da.lst[[paste0(pfx, "_da_sig")]] <-
      isTRUE(r$delta_a_significant)
  }
}
list2env(lgcm_da.lst, envir = environment())
n_da_sig <- sum(vapply(
  names(lgcm_cmp), function(k) {
    isTRUE(lgcm_cmp[[k]]$delta_a_significant)
  }, logical(1)
))
```

The paired difference
Δ~a~ = a~FRS~ − a~CVR~ quantifies how much larger
the FRS a-path is than the CVR~mimic~ a-path; a
significant positive Δ~a~ confirms FRS is the
stronger predictor of hippocampal atrophy rate.

**Males.** Δ~a~ was significantly positive across
all three domains — Memory
(Δ~a~ = `r m_mem_da`, 95% CI `r m_mem_da_ci`),
Language
(Δ~a~ = `r m_lan_da`, 95% CI `r m_lan_da_ci`),
and Executive Function
(Δ~a~ = `r m_exf_da`, 95% CI `r m_exf_da_ci`) —
confirming a consistent FRS advantage.

**Females.** Δ~a~ was significantly positive for
Language
(Δ~a~ = `r f_lan_da`, 95% CI `r f_lan_da_ci`);
Memory
(Δ~a~ = `r f_mem_da`, 95% CI `r f_mem_da_ci`) and
Executive Function
(Δ~a~ = `r f_exf_da`, 95% CI `r f_exf_da_ci`)
CIs narrowly included zero.

Paired bootstrap confirmed the a-path dissociation
in `r n_of_total.fn(n_da_sig, 6L)` sex × domain
models: Δ~a~
(FRS − CVR~mimic~) was positive in every case, with
95% BC CIs excluding zero in all but Female Memory
and Female Executive Function.
Male Δ~a~ estimates were consistent across domains
(mean Δ~a~ = `r sprintf("%.4f", m_da_mean)`),
indicating
a stable FRS advantage regardless of cognitive
outcome. Female Δ~a~ estimates were approximately
`r da_ratio_pct`% larger
(mean Δ~a~ = `r sprintf("%.4f", f_da_mean)`),
suggesting
the FRS-CVR~mimic~ dissociation is more pronounced
in females. These formal tests upgrade the
qualitative dissociation (FRS a-path significant in
all six models, CVR~mimic~ in none) to a
quantitative difference: FRS is a significantly
stronger predictor of hippocampal atrophy rate than
CVR~mimic~ within the mediation framework.

Sensitivity analyses — including HC z-score
comparisons, attrition analysis, raw (un-normalized)
concordance checks, EDT moderation, SE-constrained
LGCM validation, and A^--^/CU control analysis — are
reported in Appendix C.

# Discussion

## Summary of Findings

The apparent impact of cardiovascular risk on
hippocampal-cognitive trajectories depends on
whether the risk measure retains its age component.
The LGCM a-path dissociated: FRS predicted
hippocampal decline in
`r n_of_total.fn(abs_stats$frs_a_sig, 6L)` models
(@fig-mediation), CVR~mimic~ in none, indicating
age-weighting drives the FRS-hippocampal link. The
b-path (hippocampal-cognitive coupling) remained
significant for both FRS
(`r n_of_total.fn(abs_stats$frs_b_sig, 6L)`) and
CVR~mimic~
(`r n_of_total.fn(abs_stats$cvr_b_sig, 6L)`), so
hippocampal atrophy predicts cognitive decline
regardless of CVR measurement
[@kramer_longitudinal_2007; @oschwald_brain_2020; @anblagan_coupled_2018].
The LME 3-way interaction (FRS × HVR × YRS) was
significant across
`r n_of_total.fn(abs_lme$frs_m_sig, 3L)` male and
`r n_of_total.fn(abs_lme$frs_f_sig, 3L)` female
domains (all surviving FDR), while CVR~mimic~
reached significance only in
`r n_of_total.fn(abs_lme$cvr_f_sig, 3L)` female
domains
`r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s only)", abs_lme$cvr_f_sig_doms) else ""`
and
`r n_of_total.fn(abs_lme$cvr_m_sig, 3L)` male,
suggesting residual cardiovascular burden may
modulate HVR-cognition independently of
accelerated atrophy.

## FRS Age-Confounding and CVR~mimic~

CVR~mimic~ successfully removes the age component
from cardiovascular risk (near-zero age correlation;
@fig-cvr-validation C), supporting its use as an
age-independent index. Its moderate FRS correlation reflects shared
cardiovascular indicators while confirming distinct
constructs: one age-confounded, one age-orthogonal.

The FRS formula heavily weights age
[@dagostino_general_2008], and in older samples age
dominates the score
[@koller_development_2012; @pencina_quantifying_2019].
FRS values in ADNI are compressed into the upper
range (@fig-cvr-validation, panels A-B), limiting
score variance and statistical sensitivity,
particularly in females. When age is partialled out
(CVR~mimic~), the indirect pathway disappears: the
a-path becomes null, suggesting the FRS-hippocampal
association is primarily age-driven. This does not invalidate
FRS clinically; age is itself a potent marker of
cumulative vascular exposure
[@villeneuve_impact_2009]. However, FRS was
originally validated for adults aged 30–74
[@dagostino_general_2008], and ADNI participants
substantially exceed this range and often have
prevalent cardiovascular disease, raising questions
about FRS validity in this population.

The negative total cholesterol loading
(@tbl-mimic-loadings) likely reflects widespread
statin use in this elderly clinical sample: treated
individuals have lower cholesterol despite elevated
underlying risk, inverting the expected loading
direction. Prior FRS-brain reports in elderly
samples
[@zsoldos_association_2020; @seshadri_stroke_2004; @song_associations_2020; @meusel_cardiovascular_2019]
therefore likely conflate age-driven and
cardiovascular effects that are difficult to
disentangle without age-adjustment.

## Dissociating Moderation from Mediation

The two modelling frameworks test complementary
questions. The LME 3-way interaction
(YRS × CVR × HVR → Cognition) tests
moderation: does cardiovascular risk modify
HVR-cognition over time? The LGCM a-path
(CVR → HVR slope; @fig-mediation) tests mediation:
does cardiovascular risk affect cognition through
hippocampal decline?

For CVR~mimic~, the LGCM a-path was null across all
six models, suggesting age-adjusted cardiovascular
risk does not accelerate hippocampal atrophy. Yet
the LME 3-way interaction was significant in
`r n_of_total.fn(abs_lme$cvr_f_sig, 3L)` female
domains
`r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s)", abs_lme$cvr_f_sig_doms) else ""`
only, with
`r n_of_total.fn(abs_lme$cvr_m_sig, 3L)` in males.
This pattern represents moderation without mediation:
CVR~mimic~ does not drive structural decline (a-path
null) but may modify how hippocampal integrity
relates to cognitive performance over time,
consistent with a cognitive reserve framework
[@stern_cognitive_2012; @rabin_interactive_2018; @iadecola_framingham_2020].
`r if (length(f_cvr_cp_sig_doms) > 0) paste0("The significant CVR~mimic~ direct effect (c') for Female ", paste(DOMAIN_LABELS[f_cvr_cp_sig_doms], collapse = " and "), " (@fig-mediation) supports this, indicating a non-hippocampal pathway even when the a-path is null. ") else ""`The A^--^/CU analysis strengthens this: with AD
pathology absent, CVR~mimic~ reached FDR significance
while FRS did not (@fig-lme-forest), suggesting a
cardiovascular reserve effect masked by amyloid-driven
hippocampal decline in the A^+^ sample.

The HVR-cognition coupling (b-path) was consistent
across both cardiovascular risk predictors, in line with prior LGCM
coupling studies
[@muniz-terrera_latent_2017; @bollen_latent_2006].

## Sex Differences

CVR~mimic~ LME effects reached significance in
`r n_of_total.fn(abs_lme$cvr_m_sig, 3L)` male
domains and
`r n_of_total.fn(abs_lme$cvr_f_sig, 3L)` female
domains
`r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s)", abs_lme$cvr_f_sig_doms) else ""`.
The female CVR~mimic~ Executive finding is the only
CVR~mimic~ LME effect in either sex, consistent with
vascular burden selectively impairing executive
function in MCI [@villeneuve_nature_2011];
replication in independent samples is needed. Female
FRS variance compression limits statistical
sensitivity, so FRS findings in females may be
primarily age-driven. Sex differences in
cardiovascular risk profiles
[@ferretti_sex_2018; @mielke_consideration_2022],
differential APOE ε4 effects, and hormonal
influences on vascular regulation may all contribute
to this asymmetry. Covariate effects (age,
education, APOE ε4) appear in Appendix B
(Appendix B, Tables B3 and B4).

The sex asymmetry extended to the age covariate.
Despite GAMLSS age-normalization of HVR z-scores,
age → HVR intercept and slope paths remained
significant in all six male mediation models
(intercept `r age_cov.lst$m_hvr_i_range`, all
*p* < `r age_cov.lst$m_hvr_i_max_p`; slope all
*p* < `r age_cov.lst$m_hvr_s_max_p`), indicating
that UKB-trained normative models did not fully
absorb age effects in older ADNI males (consistent
with the transfer offset in Appendix A). Female age
effects on HVR were uniformly non-significant (all
*p* > `r age_cov.lst$f_hvr_i_min_p`). Male age
coefficients were larger in FRS models (mean
|β| = `r age_cov.lst$m_frs_mag`) than CVR~mimic~
models (`r age_cov.lst$m_cvr_mag`), consistent with
FRS and age competing for shared variance. Baseline
age also predicted the cognitive intercept in all
twelve models (all
*p* < `r age_cov.lst$cog_i_max_p`), justifying its
retention as a covariate.

## HVR-Cognitive Coupling

The HVR slope → Cognitive slope relationship
(b-path) was consistent regardless of whether FRS or
CVR~mimic~ was the predictor (@fig-mediation).
The b-path was significant in
`r n_of_total.fn(abs_stats$frs_b_sig, 6L)` FRS and
`r n_of_total.fn(abs_stats$cvr_b_sig, 6L)`
CVR~mimic~ models, supporting HVR decline as a
reliable marker of cognitive decline
[@kramer_longitudinal_2007; @oschwald_brain_2020]
and extending evidence that HVR captures
hippocampal integrity more sensitively than raw
volume [@schoemaker_hippocampal_ventricle_2019;
@fernandez-lozano_enhanced_2025;
@bruno_hippocampal_2016].

Parallel process coupling showed a complementary
pattern: significant in all three male models but
only Executive Function in females (one of three).
This sex asymmetry likely reflects lower power in
the smaller female subsample, since mediation
b-paths (which condition on a predictor) were
significant in
`r n_of_total.fn(abs_stats$frs_b_sig, 6L)` models
for both sexes. Language mediation b-paths were
non-significant for both sexes; parallel process
coupling reached Language significance in males but
not females, consistent with later-declining domains
showing less variance at this disease stage. The
hippocampal-cognitive coupling is not an artifact of
cardiovascular risk measurement; it persists across
predictor specifications and analysis frameworks.

## Methodological Considerations

The fixed-quadratic LGCM captures population-level
curvature but constrains individual quadratic
variation, so slope estimates primarily reflect
average rates of change. This makes the a-path null
finding conservative: any individual-level
non-linear CVR~mimic~ effects on hippocampal
trajectories would require a fully random quadratic
model to detect, further strengthening the FRS
age-weighting interpretation.

EDT linearity tests confirmed non-linear
trajectories (Appendix B, Table B2),
motivating the fixed-quadratic
specification [@newsom_longitudinal_2015]. Fully
random quadratic models failed to converge due to
near-zero variance. EDT accommodates heterogeneous
disease progression but is sensitive to the disease
model's assumptions
[@raket_statistical_2020].

Sensitivity analyses (Appendix C) showed HC z-score
effects were weaker than HVR effects
(Appendix C, Table C1), attrition patterns suggest
observed effects may underestimate true relationships
(Appendix C, Table C4), and raw-volume results
were concordant with z-scored analyses.

## Clinical Implications

In clinical trials targeting cardiovascular risk
reduction for dementia prevention, FRS remains a
valid summary of overall vascular burden, including
the age-driven component that reflects cumulative
exposure
[@villeneuve_impact_2009; @dagostino_general_2008].
When the question concerns cardiovascular effects
independent of aging, age-adjusted measures such as
CVR~mimic~ provide a necessary complement
[@ko_calibration_2020; @zhang_bayesian_2024]. HVR is
a more sensitive neuroimaging endpoint than
hippocampal volume alone, and its consistent
cognitive coupling supports use as a surrogate
outcome in intervention studies. The isolated
CVR~mimic~ Executive Function finding (one of six tests,
non-significant after FDR) requires replication
before clinical interpretation.

## Limitations

This study used a single cohort (ADNI), and
replication in other aging samples is needed.
UKB-trained GAMLSS normative models showed a
systematic negative z-score offset in the ADNI
comparable subsample due to age distribution
mismatch; within-ADNI relative comparisons remain
valid, but absolute z-scores should be interpreted
cautiously across cohorts. CVR~mimic~ indicators
were limited to available cardiovascular risk factor
data and measured cross-sectionally; richer or
longitudinal indicators could refine the construct.
FRS values in this elderly cohort cluster near the
30% 10-year ceiling, particularly in females, so
null CVR~mimic~ findings may partly reflect
insufficient variance after age removal rather than
a true absence of cardiovascular effects.
The A^--^/CU control analysis showed a reversed
FRS/CVR~mimic~ pattern, suggesting generalizability
beyond the AD continuum, but the smaller pooled-sex
subsample limits this inference.

ADAS-13 and MMSE items contribute to both EDT and
the Memory/Language composites; this overlap is
mitigated by (1) the bifactor CFA extracting
domain-specific variance, diluting any single shared
item's contribution, and (2) EDT entering as a time
metric rather than an outcome. Executive Function
has minimal item overlap (Appendix B, Figure B1).

Coefficient-comparison tests share the resampling
framework of mediation inference, without
independent validation. Individually-varying time
scores preclude CFI/TLI/RMSEA; model adequacy was
assessed via information criteria and nested LRTs.
The MIMIC model used 5 indicators; richer models may
yield different CVR~mimic~ values. Bootstrap
mediation CIs (N = `r med_n.lst$Male` males,
`r med_n.lst$Female` females) may be liberal with
these modest sample sizes.

## Conclusions

Whether cardiovascular risk appears to drive
hippocampal-cognitive decline depends on how it is
measured. FRS produces significant indirect effects
via hippocampal decline; age-adjusted CVR~mimic~
does not. Because age both marks cumulative vascular
exposure and drives neurodegeneration
[@villeneuve_impact_2009], FRS-hippocampal
associations primarily reflect age. The A^--^/CU
analysis (@fig-lme-forest; Appendix C) shows the
converse: with amyloid burden removed, CVR~mimic~
reached FDR significance while FRS did not,
consistent with a reserve-like mechanism.
Hippocampal-cognitive coupling persists across both
measures, supporting HVR as a reliable biomarker
target.

Future directions include replication in independent
aging cohorts, longitudinal cardiovascular
measurement, and intervention trials. The MIMIC age-adjustment
framework also applies to composite scores where a
constituent independently predicts the outcome.

---

# References {.unnumbered}

::: {#refs}
:::

---

# Ethics Approval Statement {.unnumbered}

Data used in preparation of this article were
obtained from the ADNI database. ADNI was approved by
the Institutional Review Boards of all participating
institutions. Written informed consent was obtained
from all participants.

# Author Contributions {.unnumbered}

**S.F-L.**: Conceptualization, Methodology, Software,
Formal analysis, Investigation, Data curation,
Visualization, Writing -- original draft.
**S.V.**: Conceptualization, Writing -- review & editing.
**D.L.C.**: Supervision, Project administration,
Writing -- review & editing. All authors read and
approved the final manuscript.

# Acknowledgments {.unnumbered}

*Data used in preparation of this article were
obtained from the Alzheimer's Disease Neuroimaging
Initiative (ADNI) database (adni.loni.usc.edu). As
such, the investigators
within the ADNI contributed to the design and
implementation of ADNI and/or provided data but did
not participate in analysis or writing of this
report. A complete listing of ADNI investigators can
be found at:
http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf*

# Sources of Funding {.unnumbered}

This work was supported by the Brain Canada
Foundation, through the Canada Brain Research Fund,
with the financial support of Health Canada; the
Canadian Institutes of Health Research (CIHR)
[Project Grant FRN 205819]; and La Fondation Famille
Louise & André Charron. The funders had no role in
study design, data collection and analysis, decision
to publish, or preparation of the manuscript.

Data collection and sharing for the Alzheimer's
Disease Neuroimaging Initiative (ADNI) is funded by
the National Institute on Aging (National Institutes
of Health Grant U19 AG024904). The grantee
organization is the Northern California Institute for
Research and Education. In the past, ADNI has also
received funding from the National Institute of
Biomedical Imaging and Bioengineering, the Canadian
Institutes of Health Research, and private sector
contributions through the Foundation for the National
Institutes of Health (FNIH) including generous
contributions from the following: AbbVie, Alzheimer's
Association; Alzheimer's Drug Discovery Foundation;
Araclon Biotech; BioClinica, Inc.; Biogen;
Bristol-Myers Squibb Company; CereSpir, Inc.;
Cogstate; Eisai Inc.; Elan Pharmaceuticals, Inc.; Eli
Lilly and Company; EuroImmun; F. Hoffmann-La Roche
Ltd and its affiliated company Genentech, Inc.;
Fujirebio; GE Healthcare; IXICO Ltd.; Janssen
Alzheimer Immunotherapy Research & Development, LLC.;
Johnson & Johnson Pharmaceutical Research &
Development LLC.; Lumosity; Lundbeck; Merck & Co.,
Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research;
Neurotrack Technologies; Novartis Pharmaceuticals
Corporation; Pfizer Inc.; Piramal Imaging; Servier;
Takeda Pharmaceutical Company; and Transition
Therapeutics.

# Disclosures {.unnumbered}

Declarations of interest: None.

# Data Availability Statement {.unnumbered}

This research was conducted using ADNI data
(adni.loni.usc.edu). Data are available to qualified
researchers through the ADNI data sharing platform.
Analysis code is available at
https://github.com/soffiafdz/hvr-cvr-mediation.
