---
title: "Age-Adjusted Cardiovascular Risk and Hippocampal-Cognitive Coupling Along the Alzheimer's Disease Continuum"
subtitle: "Disentangling FRS Age-Weighting from True CVR Effects"
author:
  - name: Sofia Fernandez-Lozano
    affiliations:
      - ref: bic
      - ref: ipn
  - name: Sylvia Villeneuve
    affiliations:
      - ref: bic
      - ref: psych
      - ref: stopad
  - name: D. Louis Collins
    affiliations:
      - ref: bic
      - ref: ipn
      - ref: bbme
    email: louis.collins@mcgill.ca
    corresponding: true
  - name: for the Alzheimer's Disease Neuroimaging Initiative
    affiliations: []
affiliations:
  - id: bic
    name: McConnell Brain Imaging Centre
    department: Montreal Neurological Institute
    city: Montreal
    state: Quebec
    country: Canada
  - id: ipn
    name: McGill University
    department: Department of Neurology and Neurosurgery
    city: Montreal
    state: Quebec
    country: Canada
  - id: psych
    name: McGill University
    department: Department of Psychiatry
    city: Montreal
    state: Quebec
    country: Canada
  - id: stopad
    name: Douglas Mental Health Institute Research Centre
    department: StoP-AD Centre
    city: Montreal
    state: Quebec
    country: Canada
  - id: bbme
    name: McGill University
    department: Department of Biomedical Engineering
    city: Montreal
    state: Quebec
    country: Canada
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: false
    number-sections: true
    documentclass: article
    pdf-engine: xelatex
    geometry:
      - margin=1in
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage{tikz}
        \usetikzlibrary{shapes.geometric, arrows.meta, positioning}
        \usepackage{authblk}
        \renewcommand\Authfont{\normalsize}
        \renewcommand\Affilfont{\small\itshape}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{float}
        \floatplacement{table}{H}
        \floatplacement{figure}{H}
        \usepackage{placeins}
        \usepackage{textcomp}
        \usepackage{newunicodechar}
        \newunicodechar{≤}{\ensuremath{\leq}}
        \newunicodechar{≥}{\ensuremath{\geq}}
        \newunicodechar{×}{\ensuremath{\times}}
        \newunicodechar{β}{\ensuremath{\beta}}
        \newunicodechar{η}{\ensuremath{\eta}}
        \newunicodechar{χ}{\ensuremath{\chi}}
        \newunicodechar{σ}{\ensuremath{\sigma}}
        \newunicodechar{τ}{\ensuremath{\tau}}
        \newunicodechar{γ}{\ensuremath{\gamma}}
        \newunicodechar{ζ}{\ensuremath{\zeta}}
        \newunicodechar{φ}{\ensuremath{\varphi}}
        \newunicodechar{μ}{\ensuremath{\mu}}
        \newunicodechar{ν}{\ensuremath{\nu}}
        \newunicodechar{λ}{\ensuremath{\lambda}}
        \newunicodechar{Δ}{\ensuremath{\Delta}}
        \newunicodechar{ε}{\ensuremath{\varepsilon}}
        \newunicodechar{ρ}{\ensuremath{\rho}}
        \newunicodechar{→}{\ensuremath{\rightarrow}}
        \newunicodechar{↔}{\ensuremath{\leftrightarrow}}
        \newunicodechar{≈}{\ensuremath{\approx}}
        \newunicodechar{²}{\textsuperscript{2}}
        \setlength{\LTleft}{0pt}
        \setlength{\LTright}{\fill}
knitr:
  opts_chunk:
    dev: cairo_pdf
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| include: false

# Find project root and change to it so here() works
orig_wd <- getwd()
project_root <- normalizePath(
  file.path(orig_wd, "..")
)
if (!file.exists(
  file.path(project_root, ".here")
)) {
  project_root <- orig_wd
}
setwd(project_root)

library(here)
library(data.table)
library(gt)
library(ggplot2)
library(knitr)
library(patchwork)

source(here("R/utils/config.R"))
source(here("R/utils/data_io.R"))
source(here("R/utils/tables.R"))
source(here("R/utils/plotting.R"))
source(here("R/utils/consort_diagram.R"))

config <- load_config()

# -------------------------------------------------
# Load pre-computed manuscript environment
# -------------------------------------------------
env.lst <- readRDS(
  here("outputs/manuscript_env.rds")
)
list2env(env.lst, envir = environment())
rm(env.lst)

# -------------------------------------------------------------------
# gt table rendering for LaTeX/PDF output
# -------------------------------------------------------------------

knit_print.gt_tbl <- function(x, ...) {
  if (knitr::is_latex_output()) {
    latex_code <- as.character(gt::as_latex(x))

    # Fix 1: Table width
    latex_code <- gsub(
      "\\\\begin\\{tabular\\*\\}\\{\\\\linewidth\\}\\{@\\{\\\\extracolsep\\{\\\\fill\\}\\}",
      "\\\\begin{tabular}{",
      latex_code
    )
    latex_code <- gsub(
      "\\\\end\\{tabular\\*\\}", "\\\\end{tabular}",
      latex_code
    )

    # Fix 2: HTML tags to LaTeX
    latex_code <- gsub(
      "<em>([^<]+)</em>", "\\\\textit{\\1}", latex_code
    )
    latex_code <- gsub(
      "<strong>([^<]+)</strong>", "\\\\textbf{\\1}", latex_code
    )
    latex_code <- gsub(
      "<sub>([^<]+)</sub>", "\\\\textsubscript{\\1}", latex_code
    )
    latex_code <- gsub(
      "<sup>([^<]+)</sup>", "\\\\textsuperscript{\\1}", latex_code
    )

    # Fix 3: Unicode symbols to LaTeX math
    latex_code <- gsub("Δ", "$\\\\Delta$", latex_code)
    latex_code <- gsub("β", "$\\\\beta$", latex_code)
    latex_code <- gsub("λ", "$\\\\lambda$", latex_code)
    latex_code <- gsub("σ", "$\\\\sigma$", latex_code)
    latex_code <- gsub("η", "$\\\\eta$", latex_code)
    latex_code <- gsub("χ", "$\\\\chi$", latex_code)
    latex_code <- gsub("τ", "$\\\\tau$", latex_code)
    latex_code <- gsub("γ", "$\\\\gamma$", latex_code)
    latex_code <- gsub(
      "²", "\\\\textsuperscript{2}", latex_code
    )
    latex_code <- gsub("\u2264", "$\\\\leq$", latex_code)
    latex_code <- gsub("\u2265", "$\\\\geq$", latex_code)
    latex_code <- gsub("\u2260", "$\\\\neq$", latex_code)
    latex_code <- gsub("×", "$\\\\times$", latex_code)

    # Fix 4: Unicode text commands missing {} terminator
    latex_code <- gsub(
      "\\\\(texttimes|texttwosuperior|texttwosuperiorp|textendash|textemdash)([A-Za-z])",
      "\\\\\\1{}\\2", latex_code
    )

    # Fix 5: Footnote minipage width
    latex_code <- gsub(
      "\\\\begin\\{minipage\\}\\{\\\\linewidth\\}",
      "\\\\begin{minipage}[t]{0.8\\\\textwidth}\\\\raggedright\\\\small",
      latex_code
    )

    # Fix 6: Column alignment (first left, rest right)
    fix_alignment <- function(code) {
      pattern <- "(\\\\begin\\{(?:tabular|longtable)\\}\\{)([lcr]+)(\\})"
      m <- gregexpr(pattern, code, perl = TRUE)
      if (m[[1]][1] == -1) return(code)
      matches <- regmatches(code, m)[[1]]
      for (match in matches) {
        col_spec <- gsub(".*\\{([lcr]+)\\}$", "\\1", match)
        n_cols <- nchar(col_spec)
        if (n_cols > 1) {
          new_spec <- paste0(
            "l", paste(rep("r", n_cols - 1), collapse = "")
          )
          new_match <- gsub(
            "\\{[lcr]+\\}$", paste0("{", new_spec, "}"), match
          )
          code <- sub(match, new_match, code, fixed = TRUE)
        }
      }
      code
    }
    latex_code <- fix_alignment(latex_code)

    # Fix 7: Table spacing
    latex_code <- gsub(
      "\\\\begin\\{tabular\\}",
      "\\\\renewcommand{\\\\arraystretch}{1.15}\\\\begin{tabular}",
      latex_code
    )

    # Fix 8: Escape [ ] in table cells
    latex_code <- gsub(
      "\\\\addlinespace\\[([^]]+)\\]",
      "<<ADDLNSP_\\1_>>", latex_code
    )
    latex_code <- gsub(
      "\\\\\\\\\\[([0-9.]+pt)\\]",
      "<<LINEBRK_\\1_>>", latex_code
    )
    latex_code <- gsub(
      "\\\\begin\\{minipage\\}\\[([a-z])\\]",
      "<<MINIPAGE_\\1_>>", latex_code
    )
    latex_code <- gsub("\\[", "{[}", latex_code)
    latex_code <- gsub("\\]", "{]}", latex_code)
    latex_code <- gsub(
      "<<ADDLNSP_([^_]+)_>>",
      "\\\\addlinespace[\\1]", latex_code
    )
    latex_code <- gsub(
      "<<LINEBRK_([^_]+)_>>",
      "\\\\\\\\[\\1]", latex_code
    )
    latex_code <- gsub(
      "<<MINIPAGE_([a-z])_>>",
      "\\\\begin{minipage}[\\1]", latex_code
    )

    knitr::asis_output(latex_code)
  } else {
    gt:::knit_print.gt_tbl(x, ...)
  }
}
registerS3method(
  "knit_print", "gt_tbl", knit_print.gt_tbl,
  envir = asNamespace("knitr")
)

# Domain labels for cognitive measures
DOMAIN_LABELS <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

# Helper: Format-aware markdown (handles LaTeX vs HTML)
fmt_crossformat <- function(gt_tbl, columns, rows = everything()) {
  if (knitr::is_latex_output()) {
    gt::fmt(
      gt_tbl, columns = {{ columns }},
      rows = {{ rows }}, fns = function(x) {
        x <- gsub("\\*([^*]+)\\*", "\\\\textit{\\1}", x)
        x <- gsub("~([^~]+)~", "\\\\textsubscript{\\1}", x)
        x
      }
    )
  } else {
    gt::fmt_markdown(
      gt_tbl, columns = {{ columns }}, rows = {{ rows }}
    )
  }
}
```

---

# Abstract {.unnumbered}

**INTRODUCTION:**
The Framingham Risk Score (FRS) is widely used to
index cardiovascular risk (CVR) in aging research,
yet its age-weighting may confound
associations with brain and cognitive outcomes.

**METHODS:**
In `r n_subjects` amyloid-positive ADNI participants,
we compared FRS with an age-adjusted CVR measure
(CVR~mimic~) from a MIMIC model. Sex-stratified
LME 3-way interaction models and LGCM mediation
models tested CVR effects on hippocampal-cognitive
coupling.

**RESULTS:**
FRS predicted hippocampal atrophy (a-path significant
in `r n_of_total.fn(abs_stats$frs_a_sig, 6L)` LGCM
models); CVR~mimic~ did not
(`r n_of_total.fn(abs_stats$cvr_a_sig, 6L)`). The
b-path (HVR slope → cognitive slope) was significant
in `r n_of_total.fn(abs_stats$frs_b_sig, 6L)` FRS
and `r n_of_total.fn(abs_stats$cvr_b_sig, 6L)`
CVR~mimic~ models, confirming hippocampal-cognitive
coupling is not an artifact of CVR measurement. FRS
indirect effects were significant in
`r n_of_total.fn(abs_stats$frs_ind_sig, 6L)` models;
CVR~mimic~ indirect effects in
`r n_of_total.fn(abs_stats$cvr_ind_sig, 6L)`. LME
3-way interactions (YRS × CVR × HVR) were significant
in
`r n_of_total.fn(abs_lme$frs_m_sig + abs_lme$frs_f_sig, 6L)`
FRS models (all surviving FDR) versus
`r n_of_total.fn(abs_lme$cvr_m_sig + abs_lme$cvr_f_sig, 6L)`
for CVR~mimic~.

**DISCUSSION:**
FRS "effects" on hippocampal-cognitive decline
largely reflect age-related variance. Age-adjusted
CVR measures complement FRS when the goal is to
isolate cardiovascular effects from aging per se.

**Keywords:** Cardiovascular risk, Framingham Risk
Score, hippocampus, cognitive decline, Alzheimer's
disease, MIMIC model, age-confounding

## Research in Context {.unnumbered}

**Systematic Review:**
The Framingham Risk Score (FRS) has been linked to
reduced brain volumes and accelerated cognitive
decline in multiple cohorts
[@zsoldos_association_2020; @seshadri_stroke_2004;
@song_associations_2020; @meusel_cardiovascular_2019]. However, FRS includes age as a
heavily weighted component
[@dagostino_general_2008; @koller_development_2012], and age accounts for
the majority of FRS predictive performance in older
adults [@pencina_quantifying_2019; @ko_calibration_2020], raising
the concern that apparent FRS-brain associations in
elderly samples may reflect age-driven confounding.
No prior study has directly compared an
age-confounded composite (FRS) against an
age-adjusted latent CVR measure within the same
mediation and moderation framework.

**Interpretation:**
FRS indirect effects on cognitive decline through
hippocampal atrophy primarily reflect its
age-weighting — age being itself a legitimate proxy
for cumulative vascular exposure
[@villeneuve_impact_2009]. Partialling out age via a
MIMIC model (CVR~mimic~) nullified the
CVR → hippocampal atrophy pathway. In contrast,
hippocampal-cognitive coupling remained significant
in most domains, indicating it reflects
neurodegeneration rather than cardiovascular risk.
Prior FRS-brain reports likely reflect a mixture of
age-driven and cardiovascular effects that
age-adjusted measures can help disentangle.

**Future Directions:**
Replication in independent aging cohorts (e.g.,
the Wisconsin Registry for Alzheimer's Prevention
[WRAP], the Australian Imaging, Biomarker and
Lifestyle Flagship Study of Ageing [AIBL]) with
longitudinal CVR measurement is
needed. The MIMIC-based age-adjustment framework
demonstrated here can be applied to any composite
risk score where age confounding is a concern.
Intervention studies should test whether CVR
reduction preserves hippocampal-cognitive coupling,
using age-adjusted CVR endpoints rather than FRS.

---

# Background

Cardiovascular risk factors are increasingly
recognized as modifiable contributors to cognitive
decline and dementia risk
[@livingston_dementia_2017; @ouldred_vascular_2010;
@viswanathan_vascular_2009;
@gorelick_vascular_2011;
@iadecola_neurovascular_2019].
The Framingham Risk Score (FRS), a widely used
composite measure of 10-year cardiovascular event
probability [@dagostino_general_2008], has been
associated with reduced brain volumes, compromised
white matter integrity, and accelerated cognitive
decline across multiple cohorts
[@zsoldos_association_2020;
@seshadri_stroke_2004; @song_associations_2020;
@kaffashian_predictive_2011]. However, the FRS is
heavily weighted by age, with log-hazard
coefficients of 3.06 for men and 2.33 for women
[@dagostino_general_2008], making age the dominant
predictor in the composite. In older samples where
age variance is restricted, this creates substantial
confounding between FRS and any age-related outcome
[@koller_development_2012;
@rodondi_framingham_2012].

Age is itself a strong proxy for cumulative
vascular exposure
[@villeneuve_impact_2009; @dagostino_general_2008],
and the FRS age-weighting therefore carries
legitimate clinical information. However, this
creates an interpretive challenge when FRS is used
to predict age-related outcomes: because age
simultaneously drives FRS through its formula
coefficients and independently drives
neurodegenerative outcomes through biological aging,
it acts as a common cause of both the predictor and
the outcome. Simply controlling for age in
regression models does not fully resolve this
confound, because age is not an external confounder
but an internal component of the score itself. In elderly cohorts such as ADNI,
this problem is compounded by the fact that the
ADSP-PHC-derived FRS variable is capped at 30%
10-year risk
[@dagostino_general_2008; @lee_fmnl2_2022],
compressing variance in the upper range and further
limiting sensitivity.

To address this, we employed a Multiple Indicators
Multiple Causes (MIMIC) model
[@joreskog_estimation_1975; @muthen_general_1984],
which treats age as a *cause* of cardiovascular risk
(CVR) rather than an indicator. The structural
equation
$\text{CVR} = \gamma \cdot \text{Age}_z + \zeta$
partitions observed CVR into an age-explained
component ($\gamma \cdot \text{Age}_z$) and a
residual ($\zeta$) that captures
cardiovascular risk independent of age
[@zhang_bayesian_2024]. The resulting factor scores,
termed CVR~mimic~, are orthogonal to age by
construction (r ≈ 0), enabling a direct test of
whether FRS-brain associations reflect true
cardiovascular pathophysiology or age-related
confounding.

The primary objective of this study was to
determine whether FRS effects on
hippocampal-cognitive coupling persist when
age-confounding is removed. We tested three
hypotheses. First, if FRS effects on hippocampal
atrophy are driven by age-weighting, CVR~mimic~
should show a null or attenuated a-path
(CVR → HVR slope) in latent growth curve mediation
models (LGCM). Second, if FRS mediation is
age-confounded, CVR~mimic~ indirect effects through
HVR decline should likewise be null. Third, even
when mediation is absent, CVR~mimic~ may still
moderate the brain-cognition relationship via a
three-way interaction (Years × CVR × HVR) in linear
mixed-effects (LME) models, reflecting a
reserve-like mechanism whereby cardiovascular burden
modulates the coupling between hippocampal structure
and cognitive performance without directly
accelerating atrophy. All analyses were stratified
by sex, given established sex differences in
cardiovascular risk profiles, brain aging
trajectories, and AD pathophysiology
[@ferretti_sex_2018; @mielke_consideration_2022].

---

# Methods

## Study Design and Participants

Data were drawn from the Alzheimer's Disease
Neuroimaging Initiative (ADNI), a multi-site
longitudinal study spanning phases ADNI-1, ADNI-GO,
ADNI-2, and ADNI-3. We included amyloid-positive
participants defined by PET Centiloid > 25
[@klunk_centiloid_2015] across florbetapir (FBP),
florbetaben (FBB), and Pittsburgh Compound B (PIB)
tracers, or CSF amyloid-positive status via Gaussian
Mixture Model classification
[@timsina_harmonization_2024]. Additional inclusion
criteria required available MRI-derived hippocampal
and ventricular volumes, cardiovascular risk factor
data (FRS components), and at least one cognitive
assessment with Alzheimer's Disease Sequencing
Project Phenotype Harmonization Consortium (ADSP-PHC)
harmonized composite scores. The resulting sample
encompassed the full spectrum of amyloid-positive
individuals, including cognitively unimpaired (CU),
mild cognitive impairment (MCI), and Alzheimer's
disease dementia (AD). Participant flow is shown in
Figure \ref{fig-consort}.

Data used in the preparation of this article were
obtained from the Alzheimer's Disease Neuroimaging
Initiative (ADNI) database (adni.loni.usc.edu). The
ADNI was launched in 2003 as a public-private
partnership, led by Principal Investigator Michael W.
Weiner, MD. The primary goal of ADNI has been to test
whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological
markers, and clinical and neuropsychological
assessment can be combined to measure the progression
of mild cognitive impairment (MCI) and early
Alzheimer's disease (AD).

A separate normative comparison sample of
amyloid-negative, cognitively unimpaired participants
with structural MRI within the UK Biobank (UKB) age
range (46--82 years) was used to validate the
transfer of UKB-derived z-scores to the ADNI cohort
(see Normative Transfer Validation).

```{r fig-consort}
#| output: asis

cc <- consort_counts
tikz_code <- generate_consort_tikz(
  n_database = cc$n_database,
  n_with_amyloid = cc$n_with_amyloid,
  n_amyloid_pos = cc$n_amyloid_pos,
  n_with_mri = cc$n_with_mri,
  n_lme = cc$n_lme,
  n_lgcm = cc$n_lgcm,
  n_aneg_mri = cc$n_aneg_mri,
  n_aneg_cu_mri = cc$n_aneg_cu_mri,
  n_age_comparable = cc$n_age_comparable,
  ukb_age_range = cc$ukb_age_range
)

cat("\\begin{figure}[H]\n")
cat(tikz_code)
cap <- paste0(
  "\\caption{CONSORT flow diagram showing ",
  "participant selection. Left: amyloid-positive",
  " analysis track from LME cohort to LGCM ",
  "subsample. Right: amyloid-negative, ",
  "cognitively unimpaired normative comparison",
  " sample used to validate UKB z-score ",
  "transfer. ADSP-PHC = Alzheimer's Disease ",
  "Sequencing Project Phenotype Harmonization ",
  "Consortium; CL = Centiloid; CSF = ",
  "cerebrospinal fluid; CU = cognitively ",
  "unimpaired; EDT = estimated disease time; ",
  "MRI = magnetic resonance imaging; ",
  "UKB = UK Biobank.}\n"
)
cat(cap)
cat("\\label{fig-consort}\n")
cat("\\end{figure}\n")
```

## Neuroimaging: Hippocampal-to-Ventricle Ratio (HVR)

T1-weighted magnetization-prepared rapid gradient
echo (MPRAGE) sequences were acquired on 1.5T
and 3T scanners across ADNI sites using standardized
acquisition protocols. Hippocampal (HC) and temporal
horn of the lateral ventricle (VC) volumes were
obtained using a convolutional neural network (CNN)
trained on manual labels from the same research group
[@fernandez-lozano_enhanced_2025], which demonstrated
low failure rates in ADNI compared to multi-atlas
and patch-based alternatives. Volumes were provided
in stereotaxic space (mm$^3$) and converted to
native space via subject-specific scale factors.

We defined the hippocampal-to-ventricle ratio (HVR)
as:

$$
\text{HVR} = \frac{
  \text{HC}_L + \text{HC}_R}{
  \text{HC}_L + \text{HC}_R +
  \text{VC}_L + \text{VC}_R}
$$ {#eq-hvr}

computed on bilateral summed native-space volumes
[@schoemaker_hippocampal-ventricle_2019]. This ratio
captures hippocampal integrity relative to
surrounding cerebrospinal fluid (CSF)-filled spaces,
where hippocampal atrophy is accompanied by
compensatory expansion of the temporal horns
(ex vacuo dilation). HVR approaches 1.0 when
hippocampal volume is large relative to the
ventricles (preserved structure) and approaches 0.0
when the hippocampus is small relative to the
ventricles (atrophy with ventricular expansion). The
ratio inherently normalizes for head size without
requiring intracranial volume (ICV) correction, as
both numerator and denominator scale proportionally
with head size.

HVR z-scores were derived from Generalized Additive
Models for Location, Scale, and Shape
(GAMLSS; @rigby_generalized_2005) fit on UKB data
(N = 27,434 cognitively healthy adults, age 45--83).
The distribution family was Box-Cox Cole and Green
(BCCG), which models three distributional
parameters — location ($\mu$, the conditional
median), scale ($\sigma$, coefficient of variation),
and skewness ($\nu$, Box-Cox power) — as smooth
functions of covariates: age (penalized cubic
spline), education (linear), and intracranial cavity
volume (ICC), with separate models fit for each sex.
The resulting z-score quantifies how many standard
deviations an individual's observed HVR deviates
from the conditional median predicted by the GAMLSS
model given their age, sex, education, and head
size. Negative z-scores indicate disproportionate
hippocampal volume loss relative to the normative
reference population. These are cross-sectional
deviation scores, not measures of longitudinal
change rate.

### Normative Transfer Validation

Valid cross-cohort application of UKB-trained GAMLSS
models requires verification that z-scores are
calibrated in the target ADNI population. We
performed validation in a comparable ADNI subsample
of amyloid-negative, cognitively unimpaired
participants within the UKB age range
(46--82 years), where z-scores would be expected to
approximate $N(0, 1)$. We compared mean z-scores,
standard deviations, and 95% confidence intervals
against expected $N(0, 1)$ and visualized
age-bin-specific calibration via forest plot.

## Cardiovascular Risk Assessment

### Framingham Risk Score (FRS)

We obtained FRS values from the Alzheimer's Disease
Sequencing Project Phenotype Harmonization Consortium
(ADSP-PHC) Cardiovascular Risk Factor dataset
[@lee_fmnl2_2022; @lee_reliability_2023]. The ADSP-PHC
computed FRS using the Framingham 2008 simple
office-based 10-year cardiovascular disease (CVD)
risk score via the
`ascvd_10y_frs_simple` function from the CVrisk R
package [@castro_cvrisk_2026]. Despite the function
name, this implements the Framingham general
cardiovascular risk algorithm
[@dagostino_general_2008], not the American College
of Cardiology/American Heart Association (ACC/AHA)
atherosclerotic cardiovascular disease (ASCVD)
Pooled Cohort Equations. The score estimates
10-year cardiovascular disease risk based on sex,
age (≥ 30 years), body mass index (BMI, kg/m²),
systolic blood pressure (mmHg), use of blood
pressure medication, current smoking status, and
diabetes status. This is the "simple" FRS variant
that substitutes BMI for the lipid panel (total and
high-density lipoprotein [HDL] cholesterol) required
by the general FRS
[@dagostino_general_2008]. Age coefficients in the
log-hazard scale are 3.06 (men) and 2.33 (women),
making age the dominant component in older adults.
The ADSP-PHC-derived variable is capped at 30%
10-year risk, creating a ceiling effect in elderly
samples.

### MIMIC Model for Age-Adjusted CVR (CVR~mimic~)

We specified the CVR~mimic~ latent factor as a
Multiple Indicators Multiple Causes (MIMIC) model
[@joreskog_estimation_1975; @muthen_general_1984], in which age is an
exogenous cause of the latent cardiovascular risk
factor and the disturbance ζ captures residual CVR
independent of age [@zhang_bayesian_2024].

#### Model Specification

$$
\begin{aligned}
\text{Structural:} \quad &
  \text{CVR} = \gamma \cdot \text{Age}_z + \zeta \\
\text{Measurement:} \quad &
  x_j = \lambda_j \cdot \text{CVR} + \delta_j,
  \quad j = 1, \ldots, 5 \\
\text{Residual covariances:} \quad &
  \text{Cov}(\delta_{\text{SBP}},
    \delta_{\text{CHOL}}) \neq 0; \;
  \text{Cov}(\delta_{\text{SBP}},
    \delta_{\text{CREAT}}) \neq 0
\end{aligned}
$$ {#eq-mimic}

The model includes five measurement indicators:
systolic blood pressure (SBP; z-scored, continuous),
hypertension (history of hypertension; binary),
fasting glucose (z-scored, continuous), total
cholesterol (z-scored, continuous), and serum
creatinine (z-scored, continuous). We specified two
correlated residuals: SBP ↔ Cholesterol, reflecting
shared vascular stiffness and lipid metabolism
pathways whereby arterial compliance links blood
pressure regulation to lipid profiles beyond the
general CVR factor; and SBP ↔ Creatinine, reflecting
the renal-cardiovascular axis whereby kidney function
directly influences blood pressure regulation through
the renin-angiotensin system, partially independent
of overall CVR. Both modification indices and
substantive reasoning supported these
specifications.

#### Indicator Selection

We determined the final set of five indicators
through iterative model refinement, evaluating
factor loadings, modification indices, and
substantive interpretability at each step. Several
candidate indicators were evaluated but excluded.
BMI was dropped because its factor loading became
non-significant once residual covariances with
hypertension, glucose, and cholesterol were modeled,
indicating its cardiovascular risk variance is fully
captured by the retained indicators. Diabetes, heart
disease, and stroke (ADSP-PHC binary medical history
flags) showed near-perfect tetrachoric correlations
due to a harmonization artifact in how self-reported
conditions were coded across ADNI phases, rendering
them empirically indistinguishable. Smoking status
had a non-significant factor loading, likely
reflecting survivorship bias in an elderly cohort.
Blood pressure medication showed near-complete
redundancy with hypertension status. Specific
loading estimates and model selection diagnostics
are reported in the supplement.

#### Estimation and Fit

We evaluated model fit with the robust maximum
likelihood (MLR)-scaled
comparative fit index (CFI), Tucker-Lewis index
(TLI), root mean square error of approximation
(RMSEA), and standardized root mean square residual
(SRMR).

#### Score Extraction and Age Adjustment

We extracted factor scores in three steps. First,
we obtained predicted latent factor scores (η)
from the fitted MIMIC model via empirical Bayes
modal estimation. Second, we computed age-adjusted
CVR scores (ζ) as ζ = η − γ × Age~c~, where γ
is the estimated regression of CVR on centered age
from the structural path; this removes the
age-predicted component of the latent variable,
ensuring CVR~mimic~ represents residual
cardiovascular risk after partialling out age at the
latent level rather than at the indicator level.
Third, we z-standardized the resulting ζ scores on
the common analysis sample for comparability with
FRS. This latent-level residualization operates on
the measurement-error-free latent variable and is
equivalent to extracting the disturbance term of the
MIMIC structural equation [@bollen_structural_1989].
By construction, *r*(CVR~mimic~, Age) ≈ 0.

#### Measurement Invariance

We tested measurement invariance (MI) across sex in
four steps — configural, metric, scalar, and
strict — using the confirmatory factor analysis
(CFA) measurement model only
(without the Age → CVR structural path). MI
evaluates whether factor loadings, intercepts, and
residual variances are equivalent across groups,
which is required before interpreting sex-stratified
analyses. We evaluated MI with the Satorra-Bentler
scaled chi-square difference test
[@satorra_scaled_2001], which correctly handles the
non-subtractable nature of MLR-scaled test
statistics, and with change in fit indices
(|ΔCFI| ≤ 0.01, ΔRMSEA ≤ 0.015). When the two
criteria disagreed, we adopted the stricter result.

### CVR~mimic~ Validation

We validated CVR~mimic~ by examining its correlation
with age (expected ≈ 0, confirming successful
age-partialling), correlation with FRS (moderate
expected, reflecting shared cardiovascular
indicators), distribution properties (approximate
standard normal), and FRS ceiling analysis
(percentage at the 30% risk cap) to document
variance compression in the elderly sample.

### CV Risk Stratification

For trajectory visualizations, we defined CV risk
groups by median split of baseline FRS (for FRS
analyses) or baseline CVR~mimic~ (for CVR~mimic~
analyses), and similarly defined HVR groups by
median split of baseline HVR z-score. Groups were
assigned at baseline and held constant across
follow-up.

### CVR Predictor Standardization

We z-scored both FRS and CVR~mimic~ on a common
analysis sample — participants with complete data
for both FRS components and MIMIC model indicators
— prior to all downstream analyses. This ensures
that regression coefficients from both linear
mixed-effects (LME) models and latent growth curve
models (LGCM) are on an identical scale (effect per
1 SD change), enabling direct comparison of effect
magnitudes between FRS and CVR~mimic~. We retained
raw (pre-standardized) FRS values for descriptive
statistics and ceiling analyses.

## Disease Time Estimation

Calendar time (years from baseline) assumes all
participants progress at the same rate, which is
unrealistic in a heterogeneous AD sample spanning
cognitively unimpaired (CU), mild cognitive
impairment (MCI), and dementia stages. Estimated
Disease Time (EDT) aligns participants on a common
disease timeline by estimating each individual's
position along an underlying progression curve
[@raket_statistical_2020]. We computed EDT using the
progmod R package with a simultaneous multivariate
model fitting the Alzheimer's Disease Assessment
Scale-Cognitive subscale (ADAS-Cog) 13 and the
Mini-Mental State Examination (MMSE) jointly,
sharing
a common disease time random effect across scales
[@kuhnel_simultaneous_2021]. The simultaneous
approach provides more stable individual time
estimates than univariate models because it borrows
concordant information across cognitive instruments.
EDT values were mean-centered for numerical stability
while retaining the years metric; centering does not
affect between-subject differences in disease timing.
For participants with missing EDT at later timepoints
(due to dropout), values were extrapolated using a
hybrid approach: a linear slope computed from the
individual's observed EDT values when at least two
were available, or the sex-stratified population mean
EDT slope otherwise, to minimize attrition bias from
listwise deletion.

## Cognitive Assessment

We used ADSP-PHC composite scores
[@kang_genome-wide_2023] for three cognitive domains:
memory (episodic memory composite: Wechsler Memory
Scale-Revised [WMS-R] Logical Memory, Rey Auditory
Verbal Learning Test [AVLT], ADAS-Cog word recall,
MMSE/Montreal Cognitive Assessment [MoCA] memory
items), language (language/semantic composite:
category fluency, Boston Naming Test, ADAS-Cog
naming/commands), and executive function (executive
composite: Trail Making A/B, Wechsler Adult
Intelligence Scale-Revised [WAIS-R] Digit
Span/Symbol, clock drawing). Scores were
co-calibrated across multiple cohorts via a bifactor
item banking approach, initially across ADNI, Adult
Changes in Thought (ACT), and Religious Orders
Study/Memory and Aging Project (ROS/MAP), with
subsequent addition of the National Alzheimer's
Coordinating Center (NACC). Higher composite scores
indicate better cognitive performance across all
three domains. Each score has an associated standard
error (SE) from the CFA scoring model, reflecting
measurement precision at each observation. In the
LGCM, cognitive composites enter as single indicators
with residual variances fixed to SE² (the squared
standard error), embedding known measurement error
directly from the ADSP-PHC model rather than
estimating it as a free parameter.

## Statistical Analysis

### Analytical Framework

All primary analyses (LME and LGCM) were run twice:
once with FRS and once with CVR~mimic~ as the CVR
predictor. The key diagnostic is whether FRS effects
disappear when replaced by age-adjusted CVR~mimic~,
which would implicate age-confounding rather than
true cardiovascular pathophysiology. The central
question is whether CVR influences the
hippocampal-cognitive relationship, and whether the
answer depends on how CVR is measured.

All models were sex-stratified (no pooled sample),
following the National Institutes of Health (NIH)
policy on Sex as a Biological
Variable (SABV; @clayton_policy_2014). Sex
differences in AD pathophysiology, cardiovascular
risk profiles, and brain aging trajectories motivate
stratification rather than inclusion of sex as a
covariate [@ferretti_sex_2018;
@mielke_consideration_2022]. Stratification allows
effect sizes, variance components, and covariate
relationships to differ freely between sexes; sex is
omitted from covariates because it defines the
strata. Covariates in all models include age at
baseline, education (years), and Apolipoprotein E
(APOE) ε4 carrier status.

Primary inferences are based on uncorrected per-test
p-values given the pre-specified, hypothesis-driven
nature of the analyses
[@rothman_no_1990; @perneger_whats_1998]. As a
supplementary check, we applied FDR correction
(Benjamini-Hochberg) within each CVR measure across
the 6 LME tests (3 domains × 2 sexes). FRS and
CVR~mimic~ were corrected separately because they
address distinct hypotheses. Whether effects
surviving uncorrected thresholds also survive FDR is
reported in Results. The key inferential comparison
(a-path FRS vs a-path CVR~mimic~) is inherently
directional and does not require correction across
independent model comparisons
[@perneger_whats_1998]. For the LGCM mediation
models (12 tests: 2 sexes × 3 domains ×
2 predictors), no FDR correction was applied:
inference rests on the consistency of the a-path
pattern across models rather than on any single
test (see Results).

**Analytical roadmap.** The two modelling frameworks
address distinct questions. The LME tests
*moderation*: does CVR change the HVR-cognition
coupling over time, assessed via the 3-way
YRS × CVR × HVR interaction? The LGCM tests
*mediation*: does CVR drive cognitive decline through
hippocampal atrophy, assessed via the indirect path
CVR → HVR slope → cognitive slope? Comparing FRS
and CVR~mimic~ across both frameworks isolates the
role of age-weighting in observed CVR-brain
associations.

**Formal predictor comparison.** Because FRS and
CVR~mimic~ were fitted on the same participants,
paired tests are required to compare coefficients.
In the LGCM, shared-seed bootstrap ensures identical
resamples for both predictors, yielding a paired
$\Delta_a = a_\text{FRS} - a_\text{CVR}$ with
bias-corrected 95% CI. In the LME, clustered
bootstrap (resampling subjects with replacement)
provides a paired
$\Delta\beta = \beta_\text{FRS} - \beta_\text{CVR}$
for the 3-way interaction, with bias-corrected
95% CI.

### Methodological Validations

#### EDT Linearity Validation

We used likelihood ratio tests (LRT) with LME models
comparing linear vs quadratic EDT specifications,
testing whether adding a quadratic EDT term improves
model fit. This complements the LGCM quadratic
convergence findings (see below).

#### Simulation Study: SE-Constrained Validation

A Monte Carlo simulation validated the
SE-constrained LGCM specification (residual variance
fixed to SE²) against free estimation (residual
variance freely estimated). This is a methodological
validation, not a power analysis — it validates the
estimation approach under known truth, not detection
rates for specific effect sizes. The SE-constraint
is a key methodological innovation: normative
z-scores from GAMLSS models have known SEs, so
fixing residual variance to SE² uses this
information rather than discarding it.

Data were generated from a univariate latent growth
model with known true parameters: intercept mean =
`r simulation.res$true_params$intercept_mean`, slope
mean = `r simulation.res$true_params$slope_mean` per
year EDT, intercept variance =
`r simulation.res$true_params$intercept_var`, slope
variance = `r simulation.res$true_params$slope_var`,
HVR → Slope true effect =
`r simulation.res$true_params$hvr_effect`,
measurement SE =
`r simulation.res$true_params$se_true`
(SD = `r simulation.res$true_params$se_variation`
across observations), and EDT with mean =
`r simulation.res$true_params$edt_mean` years,
SD = `r simulation.res$true_params$edt_sd`, change =
`r simulation.res$true_params$edt_change` per visit.
Two approaches were compared across
`r simulation.res$config$n_reps` replications at
N = `r paste(simulation.res$config$sample_sizes, collapse = " and N = ")`:
(a) residual variance fixed to SE² (SE-constrained)
and (b) residual variance freely estimated.
Performance metrics include parameter bias (mean
difference from truth), RMSE (root mean squared
error), and 95% CI coverage probability.

### LME Growth Models with 3-Way Interaction

The model specification (per sex) is:

$$
\begin{aligned}
\text{Cognition}_{it} = & \; (\beta_0 + b_{0i}) +
  (\beta_1 + b_{1i}) \cdot \text{YRS}_{it} \\
& + \text{YRS}_{it} \times \text{CVR}_i \times
  \text{HVR}_{zit} \;\text{(all 2- and 3-way terms)} \\
& + \text{YRS}_{it}^2 \times \text{CVR}_i \times
  \text{HVR}_{zit} \;\text{(all 2- and 3-way terms)} \\
& + \boldsymbol{X}_i\boldsymbol{\gamma} + \epsilon_{it}
\end{aligned}
$$ {#eq-lme}

where the linear block ($\text{YRS} \times \text{CVR}
\times \text{HVR}$) expands to YRS×CVR, YRS×HVR, and
YRS×CVR×HVR, and the quadratic block
($\text{YRS}^2 \times \text{CVR} \times \text{HVR}$)
expands to YRS², YRS²×CVR, YRS²×HVR, and
YRS²×CVR×HVR. The quadratic interaction terms test
whether CVR moderation of the HVR-cognition
relationship accelerates (or decelerates) over
follow-up. Random effects include subject-specific
intercepts ($b_{0i}$) and linear slopes ($b_{1i}$).
CVR is entered as either FRS or CVR~mimic~ in
parallel model sets. The hypothesis tested is whether
CVR moderates how the HVR-cognition relationship
evolves over time; a negative 3-way β means higher
CVR increasingly weakens the protective
HVR-cognition association over follow-up. Inclusion
of the quadratic time term (YRS²) was evaluated via
likelihood ratio tests comparing nested linear and
quadratic models across all 12 specifications
(2 sexes × 2 brain measures × 3 domains); results
are reported in Appendix B (see @tbl-linearity).

### LGCM Parallel Process Models

We specified LGCM with parallel HVR and cognitive
trajectories
[@bollen_latent_2006; @muniz-terrera_latent_2017] to
test whether CVR affects the hippocampal-cognitive
relationship via mediation (CVR → HVR slope →
Cognitive slope).

#### Model Specification

$$
\begin{aligned}
y_{it} &= \eta_{0i} + \eta_{1i} \cdot
  \text{EDT}_{it} + \epsilon_{it},
  \quad \text{Var}(\epsilon_{it}) = \text{SE}_{it}^2
  \\[4pt]
\eta_{0i} &= \alpha_0 +
  \boldsymbol{X}_i \boldsymbol{\gamma}_0 + \zeta_{0i}
  \\
\eta_{1i} &= \alpha_1 +
  \boldsymbol{X}_i \boldsymbol{\gamma}_1 + \zeta_{1i}
\end{aligned}
$$ {#eq-lgcm}

  where $y_{it}$ is the observed score (HVR z-score
  or cognitive composite) for person $i$ at occasion
  $t$; $\eta_{0i}$ and $\eta_{1i}$ are
  person-specific intercepts and slopes; EDT~it~ is
  the individually-varying time score (see below);
  $\epsilon_{it}$ has variance fixed to the squared
  standard error from the upstream scoring model;
  and $\boldsymbol{X}_i$ includes CVR predictor, age,
  education, and APOE ε4 status.

Two parallel processes (HVR and cognitive composite)
are specified simultaneously, with their latent
growth factors allowed to covary or linked by
directional paths depending on the model, across up
to 6 measurement occasions per participant. Cognitive
composite residual variances were fixed to SE² from
the ADSP-PHC CFA scoring model, embedding known
heteroscedastic measurement error directly rather
than estimating it as a free parameter; this yields
more precise latent slope estimates by separating
true variance from measurement noise.

We first attempted fully random quadratic models
(6 latent factors per domain); when these did not
converge, we adopted a fixed-quadratic
specification: two additional phantom latent factors
(HVR quadratic, cognitive quadratic) with variance
fixed to near zero and free population means,
capturing population-level curvature (2 additional
free parameters) without individual-level quadratic
variation. Likelihood ratio tests comparing this
fixed-quadratic model against a purely linear model
determined whether quadratic curvature improved fit
(see Results). We also attempted to embed the full
MIMIC measurement model within the OpenMx parallel
process framework; convergence was insufficient, so
we used pre-computed ζ scores instead (see Score
Extraction above).

#### Time Metric

Slope loadings use individually-varying time scores
(ITVS) determined by each participant's EDT, which
maps each visit to a position on the disease
progression timeline rather than calendar time.
Because ITVS give each participant a unique implied
covariance structure, standard fit indices (CFI,
TLI, RMSEA) that require a common saturated
reference model cannot be computed
[@grimm_growth_2016; @mehta_putting_2000].

#### Covariates and Predictors

CVR predictors (FRS and CVR~mimic~) were
mean-centered within each sex stratum to account for
sex differences in cardiovascular risk distributions;
covariates (age, education) were centered on grand
means pooled across sexes for consistency across
sex-stratified analyses. CVR~mimic~ entered as
pre-computed age-adjusted ζ scores from the MIMIC
model, z-standardized on the common sample. Both FRS
and CVR~mimic~ are observed covariates in the LGCM.
We regressed all covariates on both the latent
intercept and latent slope factors for each process
(HVR and cognitive) and fit the same models with FRS
and CVR~mimic~ for direct comparison of predictive
utility.

#### Model Fit and Inference

Missing data were handled via full information
maximum likelihood (FIML), which uses all available
observations without listwise deletion. We assessed
model fit via −2 log-likelihood (−2LL), Akaike
information criterion (AIC), and Bayesian information
criterion (BIC). For bootstrap inference, we used
case resampling with
n = `r format(lgcm_results$bootstrap$n_boot, big.mark = ",")`
iterations for all mediation paths (a, b, c',
indirect, total), reporting both percentile and
bias-corrected (BC) 95% confidence intervals. BC
intervals adjust for bias in the bootstrap
distribution and are recommended for products of
coefficients, which typically have skewed sampling
distributions [@efron_introduction_1994]. A path is
significant when its BC 95% CI excludes zero; this
is the primary inference method for all mediation
paths. We extracted individual factor scores via the
regression method (`mxFactorScores`) for
visualization of slope coupling.

#### Coupling Model

In the coupling model, a free covariance between HVR
linear slope and cognitive linear slope
(hvr~s~ ↔ cog~s~) is estimated to test whether
linear rates of hippocampal and cognitive decline
co-vary, reported as standardized correlations:

$$
r = \frac{\text{Cov}(\eta_{1,\text{hvr}},
  \eta_{1,\text{cog}})}{\sqrt{
  \text{Var}(\eta_{1,\text{hvr}}) \cdot
  \text{Var}(\eta_{1,\text{cog}})}}
$$

#### Mediation Model

In the mediation model, directional paths replace the
free slope covariance: the a-path (CVR → HVR slope)
tests whether CVR predicts the rate of hippocampal
decline; the b-path (HVR slope → cognitive slope)
captures hippocampal-cognitive coupling; the c'-path
(CVR → cognitive slope) captures the direct effect
bypassing hippocampal atrophy; and the indirect
effect (a × b) quantifies mediation through the
hippocampus. No free hvr~s~ ↔ cog~s~ covariance is
estimated in the mediation model because the
directional b-path already captures the slope
relationship; the coupling model uses a free
covariance without directional paths instead. A
significant total effect is not required for valid
mediation inference [@bollen_latent_2006].

### Sensitivity Analyses

#### LME Sensitivity

We conducted five sensitivity analyses within the
LME framework. First, for **HC z-score specificity**,
we repeated the 3-way interaction with total
hippocampal volume (HC z-score, normalized against
UKB norms) instead of HVR, testing whether CVR
moderation effects are specific to the
hippocampus-to-ventricle ratio or generalize to
hippocampal volume alone. Second, for **attrition
analysis**, we compared baseline characteristics
between completers (≥ 4 visits) and dropouts
(< 4 visits) using t-tests (continuous) and χ² tests
(categorical), plus logistic regression predicting
dropout status from baseline characteristics. Third,
for **raw HVR and HC sensitivity**, we repeated the
3-way interaction using raw (un-normalized) HVR and
raw HC (with intracranial volume covariate) instead
of UKB z-scored values, to confirm findings are not
artifacts of the normative transformation. Fourth, for **EDT
moderation**, we used EDT as a moderator in a 4-way
interaction (CVR × HVR × EDT × YRS), testing whether
CVR effects on brain-cognition coupling vary by
disease stage. Fifth, for **EDT Early/Middle
sensitivity**, we restricted LME models to the first
two EDT tertiles, excluding late-stage participants
to assess stability of effects across
disease stages.

## Software

All analyses were conducted in
R (`r R.version.string`) with
the following packages: lavaan
[@rosseel_lavaan_2012] for MIMIC model specification,
estimation (MLR + FIML), measurement invariance
testing, and factor score extraction; lme4
[@bates_fitting_2015] for LME growth models with
bobyqa optimizer and restricted maximum likelihood
(REML) estimation; OpenMx
[@neale_openmx_2016] for LGCM parallel process and
mediation models with sequential least squares
programming (SLSQP) optimizer (convergence
criterion: gradient < $10^{-5}$); progmod
[@raket_statistical_2020] for EDT computation; and
CVrisk [@castro_cvrisk_2026] for FRS computation (via
ADSP-PHC).

---

# Results

## Sample Characteristics

The LME cohort comprised
`r n_subjects` amyloid-positive participants
(`r sum(baseline.dt$SEX == "Male")` males,
`r sum(baseline.dt$SEX == "Female")` females)
with complete data for both FRS and CVR~mimic~
after common-sample filtering. Mean age was
`r sprintf("%.1f", age_mean)` ±
`r sprintf("%.1f", age_sd)` years
(range
`r sprintf("%.0f", age_min)`–`r sprintf("%.0f", age_max)`).
The diagnostic composition included `r n_cu`
cognitively unimpaired
(`r sprintf("%.0f", 100 * n_cu / n_subjects)`%),
`r n_mci` MCI
(`r sprintf("%.0f", 100 * n_mci / n_subjects)`%),
and `r n_ad` AD dementia
(`r sprintf("%.0f", 100 * n_ad / n_subjects)`%).
Participants had a median of
`r median_visits` visits
(range `r range_visits[1]`–`r range_visits[2]`)
with mean follow-up of
`r sprintf("%.1f", fu_mean)` years
(max `r sprintf("%.1f", fu_max)`).
Education averaged
`r sprintf("%.1f", educ_mean)` ±
`r sprintf("%.1f", educ_sd)` years, and
`r sprintf("%.0f", pct_apoe4)`% were APOE ε4
carriers. Baseline FRS was
`r sprintf("%.1f", frs_mean_raw)` ±
`r sprintf("%.1f", frs_sd_raw)` (raw);
CVR~mimic~ ζ was
`r sprintf("%.2f", cvr_mean)` ±
`r sprintf("%.2f", cvr_sd)` (age-adjusted
latent scores; both FRS and CVR~mimic~
z-standardized on the common sample before
analysis). The LGCM cohort included
`r n_lgcm_total` participants
(`r lgcm_n.lst$Male$n` males,
`r lgcm_n.lst$Female$n` females) with ≥2
occasions and available EDT values.
Mediation bootstrap samples comprised
`r med_n.lst$Male` males and
`r med_n.lst$Female` females (complete cases
required for case-resampling). Participant flow is
shown in Figure \ref{fig-consort}; baseline
demographics by sex appear in @tbl-demographics.

```{r tbl-demographics}
#| tbl-cap: "**Baseline cohort characteristics.** ADNI analysis cohort compared with the UK Biobank normative reference sample. Values are mean (SD) for continuous variables, % for categorical. HC = bilateral hippocampal volume; HVR = hippocampus-to-ventricle ratio; FRS = Framingham Risk Score; APOE ε4+ = carriers of at least one ε4 allele. Z-scores derived from UK Biobank GAMLSS normative models."

require_output(
  create_demographics_table(
    cohort_for_tbl.dt,
    "Baseline Demographic and Clinical Characteristics",
    normative_stats = norm_stats,
    ukb_raw = ukb_raw_stats
  ),
  "tbl-demographics"
)
```


We validated UKB-derived z-scores against an
amyloid-negative, cognitively unimpaired ADNI
subsample within the UKB age range. HVR z-scores
showed a systematic negative offset consistent
with the older ADNI age distribution; HC z-scores
were closer to expected values
(Appendix A, see @fig-normative-validation).

## CVR Characterization

### FRS Properties

FRS was strongly correlated with age (Pearson r =
`r sprintf("%.3f", frs_age_r)`, Spearman ρ =
`r sprintf("%.3f", frs_age_rho)`). The FRS-age
partial correlation (controlling for other
cardiovascular risk factors [CVRF])
remained substantial
(r = `r sprintf("%.3f", frs_partial_r)`),
confirming that age dominates FRS in older adults.
Variance decomposition showed age alone R² =
`r sprintf("%.3f", frs_var_decomp$r2_age)`,
CVRF alone R² =
`r sprintf("%.3f", frs_var_decomp$r2_cvrf)`, and
combined R² =
`r sprintf("%.3f", frs_var_decomp$r2_full)`,
indicating true CVRF contribute modestly beyond
age. A total of
`r sprintf("%.1f", frs_ceiling_all)`% of
participants were at the 30% risk cap
(`r sprintf("%.1f", frs_ceiling_male)`% males,
`r sprintf("%.1f", frs_ceiling_female)`% females),
and `r sprintf("%.1f", frs_above20_all)`% scored
above 20, limiting statistical sensitivity even
below the formal ceiling.

### CVR~mimic~ Validation

The MIMIC model showed good fit: CFI =
`r sprintf("%.3f", mimic_fit[["cfi.robust"]])`,
TLI =
`r sprintf("%.3f", mimic_fit[["tli.robust"]])`,
RMSEA =
`r sprintf("%.3f", mimic_fit[["rmsea.robust"]])`,
SRMR =
`r sprintf("%.3f", mimic_fit[["srmr"]])`. The
structural path showed γ =
`r sprintf("%.3f", mimic_gamma)` with
R² = `r sprintf("%.3f", mimic_r2)`, indicating age
explained only
`r sprintf("%.1f", mimic_r2 * 100)`% of CVR
variance and confirming CVR~mimic~ is largely
age-free. Factor loadings appear in
@tbl-mimic-loadings; visual validation in
@fig-cvr-validation.

```{r tbl-mimic-loadings}
#| tbl-cap: "**CVR MIMIC model factor loadings.** λ = unstandardized factor loading; Std. λ = standardized loading; SE = standard error."

stopifnot(!is.null(cvr_model.res))
require_output(
  create_mimic_loadings_table(
    cvr_model.res,
    "MIMIC Model Factor Loadings"
  ),
  "tbl-mimic-loadings"
)
```

After age-adjustment and z-standardization,
CVR~mimic~ showed near-zero correlation with age
(r = `r sprintf("%.3f", cvr_age_cor)`, by
construction), moderate correlation with FRS
(r = `r sprintf("%.2f", cvr_frs_cor)`, reflecting
shared CV indicators), and approximate standard
normal distribution (mean =
`r sprintf("%.2f", cvr_mean)`, SD =
`r sprintf("%.2f", cvr_sd)`).

### Measurement Invariance

MI of the CVR MIMIC model was tested across sex in
four steps (Appendix B, @tbl-mi). Configural
invariance was supported: CFI =
`r sprintf("%.3f", mi_cfg_fit[["CFI"]])`,
RMSEA =
`r sprintf("%.3f", mi_cfg_fit[["RMSEA"]])`.

```{r mi-inline-steps}
#| include: false

mi_comp <- list(
  metric = "configural_vs_metric",
  scalar = "metric_vs_scalar",
  strict = "scalar_vs_strict"
)
mi_labels <- c(
  metric = "Metric (equal loadings)",
  scalar = "Scalar (equal intercepts)",
  strict = "Strict (equal residuals)"
)
mi_step_parts <- vapply(
  names(mi_comp), function(step) {
    ckey <- mi_comp[[step]]
    delta <-
      cvr_mi.res$comparisons_delta[[ckey]]
    lrt <-
      cvr_mi.res$comparisons_lrt[[ckey]]
    if (is.null(delta) && is.null(lrt)) {
      return("")
    }
    d_cfi <- if (!is.null(delta$d_cfi)) {
      sprintf("%.3f", delta$d_cfi)
    } else "NA"
    d_rmsea <- if (!is.null(delta$d_rmsea)) {
      sprintf("%.3f", delta$d_rmsea)
    } else "NA"
    chi_p <- if (!is.null(lrt$p)) {
      fmt_p_inline(lrt$p)
    } else "NA"
    sprintf(
      "%s: ΔCFI = %s, ΔRMSEA = %s (%s)",
      mi_labels[step], d_cfi, d_rmsea, chi_p
    )
  }, character(1)
)
mi_steps_text <- paste(
  mi_step_parts[mi_step_parts != ""],
  collapse = "; "
)
scalar_delta <- cvr_mi.res$comparisons_delta[[
  "metric_vs_scalar"
]]
scalar_fail <- !is.null(scalar_delta) &&
  !is.null(scalar_delta$d_cfi) &&
  abs(scalar_delta$d_cfi) > 0.01
mi_verdict_text <- if (scalar_fail) {
  paste0(
    "Scalar invariance was not achieved",
    " (|ΔCFI| > 0.01), indicating",
    " intercepts differ across sex.",
    " Metric invariance (equal loadings)",
    " was supported \u2014 sufficient for",
    " sex-stratified analyses."
  )
} else {
  sprintf(
    "%s invariance was achieved.",
    mi_verdict
  )
}
```

`r mi_steps_text`. `r mi_verdict_text`

```{r fig-cvr-validation}
#| fig-cap: "**CVR characterization.** (A) FRS distribution by sex; dashed red line marks the 30% risk ceiling. (B) FRS vs Age showing strong positive correlation driven by age-weighting in the FRS formula. (C) CVR~mimic~ vs Age showing near-zero correlation, confirming successful age adjustment; dashed line marks zero."
#| fig-width: 6.5
#| fig-height: 4.5

stopifnot(!is.null(cvr_scores), !is.null(cohort.dt))
  # Baseline with raw FRS percentage
  frs_raw_col <- get_parameter("frs", "column")
  cvrf_raw.dt <- fread(
    get_data_path("raw", "adsp_phc_cvrf"),
    select = c("PTID", frs_raw_col)
  )
  setnames(cvrf_raw.dt, frs_raw_col, "FRS_pct")

  baseline_merged.dt <- cohort.dt[, .(
    AGE = AGE[1], SEX = SEX[1]
  ), by = .(RID, PTID)]
  baseline_merged.dt <- merge(
    baseline_merged.dt,
    cvrf_raw.dt, by = "PTID"
  )
  merged_plot.dt <- merge(
    baseline_merged.dt,
    cvr_scores[, .(RID, CVR_mimic)],
    by = "RID"
  )

  p_frs_sex <- plot_frs_ceiling_by_sex(
    merged_plot.dt
  )
  p_frs_age <- plot_frs_validation(
    merged_plot.dt, frs_age_r = frs_age_r
  )
  p_cvr_age <- plot_cvr_validation(
    merged_plot.dt
  )

  p_frs_sex / (p_frs_age + p_cvr_age) +
    plot_layout(heights = c(1, 1)) +
    plot_annotation(tag_levels = "A")
```


EDT linearity tests and SE-constrained LGCM
simulation validation are reported in Appendix B
(see @tbl-linearity).

## LME Results: FRS vs CVR~mimic~ 3-Way Interaction

### 3-Way Interaction (FRS vs CVR~mimic~)

```{r fig-lme-forest}
#| fig-cap: "**LME 3-way interaction forest plot.** FRS vs CVR~mimic~ effects (YRS × CVR × HVR z-score → Cognition) by sex and cognitive domain. Filled points indicate p < 0.05; white-filled points indicate p ≥ 0.05. Error bars show 95% CI."
#| fig-width: 6.5
#| fig-height: 2.5

plot_lme_forest_comparison(frs_hvr, cvr_hvr)
```

Full covariate effects (Age, Education, APOE ε4,
CVR main effect, HVR z-score) are reported in
Appendix B (@tbl-lme-covariates,
@tbl-cvr-lme-covariates).

```{r lme-inline-results}
#| include: false

stopifnot(!is.null(frs_hvr), !is.null(cvr_hvr))
get_fdr_p <- function(cvr_meas, sex, domain) {
  row <- lme_summary.dt[
    Analysis == "Primary_Stratified" &
    CVR_Measure == cvr_meas &
    Sex == sex & Domain == domain
  ]
  if (nrow(row) > 0 &&
      !is.null(row$p_fdr)) {
    row$p_fdr[1]
  } else NA_real_
}
get_lme_result <- function(
    res, sex, domain, cvr_meas) {
  r <- res$stratified[[sex]][[domain]]$interaction
  if (!is.null(r)) {
    list(
      beta = r$beta, p = r$p_value,
      p_fdr = get_fdr_p(cvr_meas, sex, domain)
    )
  } else NULL
}
fmt_p_fdr <- function(p, p_fdr) {
  base <- fmt_p_inline(p)
  if (!is.na(p_fdr) && p < 0.05) {
    fdr_tag <- if (p_fdr < 0.05) {
      sprintf(
        " [*p*~FDR~ = %s]",
        fmt_p_inline(p_fdr)
      )
    } else {
      sprintf(
        " [*p*~FDR~ = %s, n.s.]",
        fmt_p_inline(p_fdr)
      )
    }
    paste0(base, fdr_tag)
  } else base
}

frs_all <- list()
cvr_all <- list()
for (sx in c("Male", "Female")) {
  for (d in c("MEM", "LAN", "EXF")) {
    frs_all[[paste(sx, d)]] <-
      get_lme_result(frs_hvr, sx, d, "FRS")
    cvr_all[[paste(sx, d)]] <-
      get_lme_result(
        cvr_hvr, sx, d, "CVR_mimic"
      )
  }
}
frs_n_sig <- sum(sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
))
cvr_n_sig <- sum(sapply(
  cvr_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
))
frs_betas <- abs(sapply(
  frs_all, function(r) {
    if (!is.null(r)) r$beta else NA
  }
))
cvr_betas <- abs(sapply(
  cvr_all, function(r) {
    if (!is.null(r)) r$beta else NA
  }
))
beta_ratio <- mean(
  frs_betas, na.rm = TRUE
) / mean(cvr_betas, na.rm = TRUE)

frs_sig_keys <- names(frs_all)[sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
)]
fmt_doms <- function(keys) {
  ds <- sub("^(Male|Female) ", "", keys)
  paste(DOMAIN_LABELS[ds], collapse = ", ")
}
frs_sig_male <- grep(
  "^Male", frs_sig_keys, value = TRUE
)
frs_sig_female <- grep(
  "^Female", frs_sig_keys, value = TRUE
)
lme_male_detail <- paste0(
  fmt_doms(frs_sig_male)
)
lme_female_detail <- paste0(
  fmt_doms(frs_sig_female)
)

# Per-sex, per-domain results for inline use
lme.lst <- list()
for (sx in c("m", "f")) {
  sx_full <- ifelse(sx == "m", "Male", "Female")
  cvr_ps <- c()
  for (d in c("MEM", "LAN", "EXF")) {
    rf <- get_lme_result(
      frs_hvr, sx_full, d, "FRS"
    )
    rc <- get_lme_result(
      cvr_hvr, sx_full, d, "CVR_mimic"
    )
    pfx <- paste0(sx, "_", tolower(d))
    lme.lst[[paste0(pfx, "_frs_b")]] <-
      sprintf("%.4f", rf$beta)
    lme.lst[[paste0(pfx, "_frs_p")]] <-
      fmt_p_fdr(rf$p, rf$p_fdr)
    lme.lst[[paste0(pfx, "_cvr_b")]] <-
      sprintf("%.4f", rc$beta)
    lme.lst[[paste0(pfx, "_cvr_p")]] <-
      fmt_p_fdr(rc$p, rc$p_fdr)
    cvr_ps <- c(cvr_ps, rc$p)
  }
  lme.lst[[paste0(sx, "_cvr_min_p")]] <-
    sprintf("%.2f", min(cvr_ps))
}
list2env(lme.lst, envir = environment())

# Worked example
frs_sig <- frs_all[sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
)]
worked_text <- ""
if (length(frs_sig) > 0) {
  max_r <- frs_sig[[which.max(
    abs(sapply(frs_sig, `[[`, "beta"))
  )]]
  max_nm <- names(frs_sig)[which.max(
    abs(sapply(frs_sig, `[[`, "beta"))
  )]
  parts <- strsplit(max_nm, " ")[[1]]
  display_nm <- paste(
    parts[1], DOMAIN_LABELS[parts[2]]
  )
  worked_text <- sprintf(
    paste0(
      "The largest significant FRS 3-way",
      " effect (%s, β = %.4f) implies",
      " that over 5 years, a 1-SD increase",
      " in FRS is associated with a",
      " %.4f-unit cumulative divergence",
      " in the HVR-cognition association."
    ),
    display_nm, max_r$beta,
    max_r$beta * 5
  )
}
```

**Key pattern.** FRS produced significant 3-way
interactions (YRS × CVR × HVR) in males
(`r lme_male_detail`) and females
(`r lme_female_detail`) —
`r n_of_total.fn(frs_n_sig, 6L)` models, all
surviving FDR correction. CVR~mimic~ showed
`r n_of_total.fn(cvr_n_sig, 6L)` significant
interactions
(`r n_of_total.fn(fdr_summary.lst$cvr_n_fdr, 6L)`
surviving FDR), consistent with FRS effects being
primarily driven by age-weighting rather than true
cardiovascular risk (see @fig-lme-forest).

**Males.** FRS × HVR × YRS was significant for
Memory (β = `r m_mem_frs_b`, `r m_mem_frs_p`),
Language (β = `r m_lan_frs_b`, `r m_lan_frs_p`),
and Executive Function
(β = `r m_exf_frs_b`, `r m_exf_frs_p`).
CVR~mimic~ showed no significant interactions
(all *p* > `r m_cvr_min_p`).

**Females.** FRS effects were also significant
across all three domains, with larger magnitudes:
Memory (β = `r f_mem_frs_b`, `r f_mem_frs_p`),
Language (β = `r f_lan_frs_b`, `r f_lan_frs_p`),
and Executive Function
(β = `r f_exf_frs_b`, `r f_exf_frs_p`).
CVR~mimic~ reached significance only for Executive
Function
(β = `r f_exf_cvr_b`, `r f_exf_cvr_p`).

FRS β values were on average
`r sprintf("%.1f", beta_ratio)`-fold larger in
absolute magnitude than CVR~mimic~ β values,
consistent with age-weighting amplifying apparent
FRS effects. `r worked_text`

```{r lme-fdr-note}
#| include: false

stopifnot(!is.null(fdr_summary.lst))
frs_changed <- fdr_summary.lst$frs_n_sig !=
  fdr_summary.lst$frs_n_fdr
cvr_changed <- fdr_summary.lst$cvr_n_sig !=
  fdr_summary.lst$cvr_n_fdr
fdr_note_text <- if (
  !frs_changed && !cvr_changed
) {
  paste0(
    "FDR correction (Benjamini-Hochberg)",
    " did not change any conclusions for",
    " either CVR measure."
  )
} else if (!frs_changed && cvr_changed) {
  sprintf(
    paste0(
      "FDR correction did not change FRS",
      " conclusions. For CVR~mimic~, %d of",
      " %d uncorrected effects survived FDR."
    ),
    fdr_summary.lst$cvr_n_fdr,
    fdr_summary.lst$cvr_n_sig
  )
} else {
  sprintf(
    paste0(
      "After FDR correction, %s FRS and",
      " %s CVR~mimic~ uncorrected effects",
      " remained significant."
    ),
    n_of_total.fn(
      fdr_summary.lst$frs_n_fdr,
      fdr_summary.lst$frs_n_sig
    ),
    n_of_total.fn(
      fdr_summary.lst$cvr_n_fdr,
      fdr_summary.lst$cvr_n_sig
    )
  )
}
```

`r fdr_note_text`

### Formal Coefficient Comparison (LME)

To formally compare FRS and CVR~mimic~ 3-way
interaction coefficients, we computed clustered
bootstrap Δβ with bias-corrected 95% CIs.

```{r lme-comparison-boot}
#| include: false

lme_cmp <- lme_hvr_z.res$comparison
stopifnot(
  !is.null(lme_cmp), length(lme_cmp) > 0
)
lme_db.lst <- list()
for (sx in c("m", "f")) {
  sx_full <- ifelse(
    sx == "m", "Male", "Female"
  )
  for (d in c("MEM", "LAN", "EXF")) {
    k <- paste(sx_full, d, sep = "_")
    r <- lme_cmp[[k]]
    pfx <- paste0(sx, "_", tolower(d))
    lme_db.lst[[paste0(pfx, "_db")]] <-
      sprintf("%.4f", r$delta_beta)
    lme_db.lst[[paste0(pfx, "_ci")]] <-
      sprintf(
        "[%.4f, %.4f]",
        r$ci_lower, r$ci_upper
      )
  }
}
list2env(lme_db.lst, envir = environment())
```

**Males.** Δβ values were negative across all three
domains, reaching significance for Language
(Δβ = `r m_lan_db`, 95% CI `r m_lan_ci`) but not
Memory (Δβ = `r m_mem_db`, 95% CI `r m_mem_ci`) or
Executive Function
(Δβ = `r m_exf_db`, 95% CI `r m_exf_ci`).

**Females.** Δβ values were likewise negative.
Memory reached significance
(Δβ = `r f_mem_db`, 95% CI `r f_mem_ci`), whereas
Language
(Δβ = `r f_lan_db`, 95% CI `r f_lan_ci`) and
Executive Function
(Δβ = `r f_exf_db`, 95% CI `r f_exf_ci`)
did not.

## LGCM Results

### Growth Parameters

LGCM models used EDT as the time metric with
individually-varying time scores, fit separately per
sex and cognitive domain.

```{r lgcm-growth-inline}
#| include: false

stopifnot(!is.null(lgcm_parallel$linear_results))
domains <- c("MEM", "LAN", "EXF")

lgcm_growth_text.lst <- list()
for (sx in c("Male", "Female")) {
  hvr_slopes <- c()
  cog_slopes <- c()
  for (d in domains) {
    k <- paste0(sx, "_", d)
    lr <- lgcm_parallel$linear_results[[k]]
    if (is.null(lr)) next
    s <- lr$summary
    if (is.null(s)) next
    hvr_s <- s[
      s$label == "hvr_slope_mean" |
      grepl("hvr.*slope.*mean", s$label,
        ignore.case = TRUE), ]
    cog_s <- s[
      s$label == "cog_slope_mean" |
      grepl("cog.*slope.*mean", s$label,
        ignore.case = TRUE), ]
    if (nrow(hvr_s) > 0) {
      hvr_slopes <- c(
        hvr_slopes, hvr_s$Estimate[1]
      )
    }
    if (nrow(cog_s) > 0) {
      cog_slopes <- c(
        cog_slopes, cog_s$Estimate[1]
      )
    }
  }
  lgcm_growth_text.lst[[sx]] <- if (
    length(hvr_slopes) > 0 &&
    length(cog_slopes) > 0
  ) {
    sprintf(
      paste0(
        "In %ss, mean HVR slopes were negative",
        " across all domains (range: %.4f to",
        " %.4f); cognitive slopes were also",
        " negative (range: %.4f to %.4f),",
        " confirming progressive decline in",
        " both processes."
      ),
      tolower(sx),
      min(hvr_slopes), max(hvr_slopes),
      min(cog_slopes), max(cog_slopes)
    )
  } else ""
}
```

`r lgcm_growth_text.lst$Male`
`r lgcm_growth_text.lst$Female`
Model fit appears in @tbl-mediation-fit; standard
fit indices (CFI, TLI, RMSEA) are not computable
with individually-varying time scores, so adequacy
was assessed via −2LL, AIC, BIC, and nested LRTs.

### Formal a-Path Comparison (LGCM)

Paired bootstrap with shared-seed resampling was
used to compare FRS and CVR~mimic~ a-path
coefficients.

```{r lgcm-comparison-boot}
#| include: false

lgcm_cmp <- lgcm_results$comparisons
stopifnot(
  !is.null(lgcm_cmp), length(lgcm_cmp) > 0
)
lgcm_da.lst <- list()
for (sx in c("m", "f")) {
  sx_full <- ifelse(
    sx == "m", "Male", "Female"
  )
  for (d in c("MEM", "LAN", "EXF")) {
    k <- paste(sx_full, d, sep = "_")
    r <- lgcm_cmp[[k]]
    pfx <- paste0(sx, "_", tolower(d))
    lgcm_da.lst[[paste0(pfx, "_da")]] <-
      sprintf("%.5f", r$delta_a_est)
    lgcm_da.lst[[paste0(pfx, "_da_ci")]] <-
      sprintf(
        "[%.5f, %.5f]",
        r$delta_a_ci_lower,
        r$delta_a_ci_upper
      )
    lgcm_da.lst[[paste0(pfx, "_da_sig")]] <-
      isTRUE(r$delta_a_significant)
  }
}
list2env(lgcm_da.lst, envir = environment())
n_da_sig <- sum(vapply(
  names(lgcm_cmp), function(k) {
    isTRUE(lgcm_cmp[[k]]$delta_a_significant)
  }, logical(1)
))
```

The paired difference
Δ~a~ = a~FRS~ − a~CVR~ quantifies how much larger
the FRS a-path is than the CVR~mimic~ a-path; a
significant positive Δ~a~ confirms FRS is the
stronger predictor of hippocampal atrophy rate.

**Males.** Δ~a~ was significantly positive across
all three domains — Memory
(Δ~a~ = `r m_mem_da`, 95% CI `r m_mem_da_ci`),
Language
(Δ~a~ = `r m_lan_da`, 95% CI `r m_lan_da_ci`),
and Executive Function
(Δ~a~ = `r m_exf_da`, 95% CI `r m_exf_da_ci`) —
confirming a consistent FRS advantage.

**Females.** Δ~a~ was significant for Language
(Δ~a~ = `r f_lan_da`, 95% CI `r f_lan_da_ci`)
and Executive Function
(Δ~a~ = `r f_exf_da`, 95% CI `r f_exf_da_ci`),
but the Memory CI narrowly included zero
(Δ~a~ = `r f_mem_da`, 95% CI `r f_mem_da_ci`).

Paired bootstrap confirmed the a-path dissociation
in `r n_of_total.fn(n_da_sig, 6L)` sex × domain
models: Δ~a~
(FRS − CVR~mimic~) was positive in every case, with
95% BC CIs excluding zero in all but Female Memory.
Male Δ~a~ estimates were consistent across domains
(mean = `r sprintf("%.4f", m_da_mean)`), indicating
a stable FRS advantage regardless of cognitive
outcome. Female Δ~a~ estimates were approximately
`r da_ratio_pct`% larger
(mean = `r sprintf("%.4f", f_da_mean)`), suggesting
the FRS-CVR~mimic~ dissociation is more pronounced
in females. These formal tests upgrade the
qualitative dissociation (FRS a-path significant in
all six models, CVR~mimic~ in none) to a
quantitative difference: FRS is a significantly
stronger predictor of hippocampal atrophy rate than
CVR~mimic~ within the mediation framework.

### Slope Coupling {#sec-slope-coupling}

LGCM parallel process models tested whether HVR and
cognitive linear decline rates co-vary (latent slope
coupling, hvr~s~ ↔ cog~s~) along the EDT disease
timeline.

```{r coupling-inline}
#| include: false

stopifnot(!is.null(lgcm_parallel$linear_results))
domains <- c("MEM", "LAN", "EXF")
domain_labels <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

coupling_text.lst <- list()
for (sx in c("Male", "Female")) {
  cors <- c()
  cor_names <- c()
  ps <- c()
  for (d in domains) {
    k <- paste0(sx, "_", d)
    lr <- lgcm_parallel$linear_results[[k]]
    if (!is.null(lr$coupling) &&
        !is.na(lr$coupling$p)) {
      prm <- lr$params
      v_h <- prm[
        name == "var_hvr_s", Estimate
      ]
      v_c <- prm[
        name == "var_cog_s", Estimate
      ]
      r_val <- lr$coupling$cov /
        sqrt(v_h * v_c)
      cors <- c(cors, r_val)
      cor_names <- c(
        cor_names, domain_labels[d]
      )
      ps <- c(ps, lr$coupling$p)
    }
  }
  coupling_text.lst[[sx]] <- if (
    length(cors) > 0
  ) {
    max_d <- cor_names[which.max(cors)]
    min_d <- cor_names[which.min(cors)]
    max_p <- max(ps)
    p_bound <- if (max_p < 0.001) {
      "all *p* < 0.001"
    } else {
      sprintf("max %s", fmt_p_inline(max_p))
    }
    sprintf(
      paste0(
        "In %ss, linear slope correlations",
        " (*r*) ranged from %.3f (%s) to",
        " %.3f (%s; %s)."
      ),
      tolower(sx),
      min(cors), min_d,
      max(cors), max_d, p_bound
    )
  } else ""
}
```

`r coupling_text.lst$Male`
`r coupling_text.lst$Female`
In males, faster hippocampal decline was
significantly associated with steeper cognitive
decline across all three domains; in females, the
direction was consistent but only Executive Function
reached significance, likely reflecting lower
statistical power in the smaller female subsample.
This coupling supports the b-path assumption in the
mediation framework: HVR decline predicts cognitive
decline. Latent factor scores from the parallel
process model confirm this pattern (see
@fig-coupling).

```{r fig-coupling}
#| fig-cap: "**HVR–cognitive latent slope coupling.** Females in red, Males in blue throughout. A. Individual trajectories (random 20% subsample) with population mean curves (bold; including fixed quadratic curvature where selected by LRT). 2×2 facets: HVR, Memory, Language, Executive. B. Latent slope coupling: latent cognitive decline rate (x) vs latent HVR decline rate (y) from factor scores (positive = faster decline); solid lines show linear trends per sex; grey line shows 1:1 reference. Correlation estimates from the SEM (* p < .05, ** p < .01, *** p < .001)."
#| fig-width: 6.5
#| fig-height: 7.5

lgcm_m_input <- read_rds_safe(here(
  "data/derivatives/lgcm/lgcm_input_male.rds"
))
lgcm_f_input <- read_rds_safe(here(
  "data/derivatives/lgcm/lgcm_input_female.rds"
))
stopifnot(
  !is.null(lgcm_m_input),
  !is.null(lgcm_f_input),
  !is.null(lgcm_parallel)
)
plot_coupling_figure(
  lgcm_parallel,
  lgcm_m_input, lgcm_f_input
)
```

### Mediation: FRS vs CVR~mimic~

LGCM mediation models tested whether CVR effects on
cognitive decline operate through hippocampal
atrophy, comparing FRS and CVR~mimic~ as predictors
(see @fig-mediation).

```{r tbl-mediation-fit}
#| tbl-cap: "**LGCM mediation model fit indices.** Sex-stratified. -2LL = -2 log-likelihood; AIC = Akaike information criterion; BIC = Bayesian information criterion."

require_output(
  create_mediation_fit_table(
    lgcm_results,
    title = "Mediation Model Fit"
  ),
  "tbl-mediation-fit"
)
```

Mediation path diagrams showing a-, b-, c'-path
estimates and bootstrap CIs appear in
@fig-mediation.


```{r fig-mediation}
#| dev: png
#| fig-cap: "**LGCM mediation path diagrams.** Rows: Memory, Language, Executive Function. Columns: FRS, CVR~MIMIC~. Each panel shows Male (blue) and Female (red) path estimates. Solid/bold = significant (bootstrap 95% CI excludes zero); faded = non-significant. Star (\\*) indicates CI excludes zero. 95% bootstrap CIs shown for all paths."
#| fig-width: 6.5
#| fig-height: 8.0
#| fig-dpi: 300

panels <- build_combined_med_panels(
  lgcm_med_male, lgcm_med_female
)
wrap_plots(panels, ncol = 2) +
  plot_annotation(tag_levels = "A")
```

```{r lgcm-inline-results}
#| include: false

stopifnot(!is.null(lgcm_med_male))
stopifnot(!is.null(lgcm_med_female))

domains <- c("MEM", "LAN", "EXF")

frs_a_sig <- sum(sapply(
  c("Male", "Female"), function(sx) {
    med <- if (sx == "Male") {
      lgcm_med_male
    } else lgcm_med_female
    sapply(domains, function(d) {
      r <- med$results[[
        paste0(d, "_FRS")
      ]]$a
      !is.null(r) && ci_sig_check.fn(r)
    })
  }
))
cvr_a_sig <- sum(sapply(
  c("Male", "Female"), function(sx) {
    med <- if (sx == "Male") {
      lgcm_med_male
    } else lgcm_med_female
    sapply(domains, function(d) {
      r <- med$results[[
        paste0(d, "_CVR")
      ]]$a
      !is.null(r) && ci_sig_check.fn(r)
    })
  }
))
cvr_a_null <- 6L - cvr_a_sig

all_ind_sig <- 0L
all_ind_total <- 0L
ind_sig_labels <- c()
for (sx in c("Male", "Female")) {
  med <- if (sx == "Male") {
    lgcm_med_male
  } else lgcm_med_female
  for (d in domains) {
    for (pred in c("FRS", "CVR")) {
      k <- paste0(d, "_", pred)
      r <- med$results[[k]]
      all_ind_total <- all_ind_total + 1L
      if (!is.null(r$indirect) &&
          ci_sig_check.fn(r$indirect)) {
        all_ind_sig <- all_ind_sig + 1L
        ind_sig_labels <- c(
          ind_sig_labels,
          sprintf(
            "%s %s (%s)",
            sx, DOMAIN_LABELS[d], pred
          )
        )
      }
    }
  }
}
ind_detail <- if (length(ind_sig_labels) > 0) {
  paste0(
    " — ",
    paste(ind_sig_labels, collapse = ", ")
  )
} else ""

# Per-sex mediation scalars for inline use
med_inline.lst <- list()
for (sx in c("m", "f")) {
  sx_full <- ifelse(
    sx == "m", "Male", "Female"
  )
  med <- if (sx == "m") {
    lgcm_med_male
  } else lgcm_med_female
  frs_a.v <- c()
  frs_b.v <- c()
  frs_a_nsig <- 0L
  cvr_a_nsig <- 0L
  frs_b_nsig <- 0L
  frs_ind_nsig <- 0L
  cvr_ind_nsig <- 0L
  cvr_max_abs_a <- 0
  for (d in domains) {
    rf <- med$results[[paste0(d, "_FRS")]]
    rc <- med$results[[paste0(d, "_CVR")]]
    if (!is.null(rf$a)) {
      frs_a.v <- c(frs_a.v, rf$a$est)
      if (ci_sig_check.fn(rf$a))
        frs_a_nsig <- frs_a_nsig + 1L
    }
    if (!is.null(rc$a)) {
      cvr_max_abs_a <- max(
        cvr_max_abs_a, abs(rc$a$est)
      )
      if (ci_sig_check.fn(rc$a))
        cvr_a_nsig <- cvr_a_nsig + 1L
    }
    if (!is.null(rf$b)) {
      frs_b.v <- c(frs_b.v, rf$b$est)
      if (ci_sig_check.fn(rf$b))
        frs_b_nsig <- frs_b_nsig + 1L
    }
    if (!is.null(rf$indirect) &&
        ci_sig_check.fn(rf$indirect))
      frs_ind_nsig <- frs_ind_nsig + 1L
    if (!is.null(rc$indirect) &&
        ci_sig_check.fn(rc$indirect))
      cvr_ind_nsig <- cvr_ind_nsig + 1L
  }
  pfx <- sx
  med_inline.lst[[paste0(pfx, "_frs_a_range")]] <-
    sprintf("%.4f to %.4f",
      min(frs_a.v), max(frs_a.v))
  med_inline.lst[[paste0(pfx, "_frs_a_nsig")]] <-
    n_of_total.fn(frs_a_nsig, 3L)
  med_inline.lst[[paste0(pfx, "_cvr_a_nsig")]] <-
    n_of_total.fn(cvr_a_nsig, 3L)
  med_inline.lst[[paste0(pfx, "_cvr_max_a")]] <-
    sprintf("%.2f", cvr_max_abs_a + 0.005)
  med_inline.lst[[paste0(pfx, "_frs_b_range")]] <-
    sprintf("%.3f to %.3f",
      min(frs_b.v), max(frs_b.v))
  med_inline.lst[[paste0(pfx, "_frs_b_nsig")]] <-
    n_of_total.fn(frs_b_nsig, 3L)
  med_inline.lst[[paste0(pfx, "_frs_ind")]] <-
    n_of_total.fn(frs_ind_nsig, 3L)
  med_inline.lst[[paste0(pfx, "_cvr_ind")]] <-
    n_of_total.fn(cvr_ind_nsig, 3L)
}
list2env(
  med_inline.lst, envir = environment()
)
```

**FRS vs CVR~mimic~ across all 12 models.** The
a-path (CVR → HVR slope) showed a clear predictor
dissociation: FRS significantly predicted
hippocampal decline in
`r n_of_total.fn(frs_a_sig, 6L)` models (bootstrap
95% CIs excluding zero), whereas CVR~mimic~ did not
in any — indicating that FRS "effects" on
hippocampal decline are largely consistent with
age-weighting. By contrast, the b-path
(HVR slope → cognitive slope) was significant for
`r n_of_total.fn(abs_stats$frs_b_sig, 6L)` FRS and
`r n_of_total.fn(abs_stats$cvr_b_sig, 6L)`
CVR~mimic~ models, confirming that hippocampal-
cognitive coupling persists regardless of predictor.
The FRS indirect effect (a × b) was significant in
`r n_of_total.fn(abs_stats$frs_ind_sig, 6L)` models;
CVR~mimic~ indirect effects in
`r n_of_total.fn(abs_stats$cvr_ind_sig, 6L)`.

**Males.** FRS predicted hippocampal atrophy across
`r m_frs_a_nsig` domains (a-path β range
`r m_frs_a_range`, all bootstrap 95% CIs excluding
zero); CVR~mimic~ did not
(all |β| < `r m_cvr_max_a`). Hippocampal decline
in turn predicted cognitive decline (b-path β range
`r m_frs_b_range`, `r m_frs_b_nsig` CIs excluding
zero), and the FRS indirect effect was significant
for `r m_frs_ind` domains. No CVR~mimic~ indirect
effects reached significance.

**Females.** The same a-path dissociation held: FRS
was significant across `r f_frs_a_nsig` domains
(β range `r f_frs_a_range`), CVR~mimic~ was not
(all |β| < `r f_cvr_max_a`). B-path estimates were
comparable to males (β range `r f_frs_b_range`,
`r f_frs_b_nsig` CIs excluding zero), and FRS
indirect effects reached significance in
`r f_frs_ind` domains; CVR~mimic~ indirect effects
reached significance in `r f_cvr_ind`. Notably,
the CVR~mimic~ Memory model showed a significant
direct effect (c' =
`r sprintf("%.4f", lgcm_med_female$results$MEM_CVR$cprime$est)`,
95% CI
[`r sprintf("%.4f", lgcm_med_female$results$MEM_CVR$cprime$boot_ci_lower)`,
`r sprintf("%.4f", lgcm_med_female$results$MEM_CVR$cprime$boot_ci_upper)`]),
suggesting CVR~mimic~ may influence memory decline
through a non-hippocampal pathway.

Complete path estimates, bootstrap CIs, and direct
(c') effects appear in @fig-mediation.


Sensitivity analyses — including HC z-score
comparisons, attrition analysis, raw (un-normalized)
concordance checks, EDT moderation, and SE-constrained
LGCM validation — are reported in Appendix C.

# Discussion

## Summary of Findings

The apparent impact of cardiovascular risk on
hippocampal-cognitive trajectories depends on
whether the risk measure retains its age component.
Three findings converge on this conclusion. First,
the LGCM a-path dissociation: FRS predicted
hippocampal decline in
`r n_of_total.fn(abs_stats$frs_a_sig, 6L)` models
(@fig-mediation), whereas CVR~mimic~ did not in any,
demonstrating that the FRS-hippocampal link is
driven by age-weighting rather than cardiovascular
pathophysiology. Second, despite this predictor
dissociation, the b-path (hippocampal-cognitive
coupling) remained significant for both FRS
(`r n_of_total.fn(abs_stats$frs_b_sig, 6L)`) and
CVR~mimic~
(`r n_of_total.fn(abs_stats$cvr_b_sig, 6L)`),
confirming that hippocampal decline reliably
predicts cognitive decline regardless of CVR measure
[@kramer_longitudinal_2007; @oschwald_brain_2020; @anblagan_coupled_2018].
Third, the LME 3-way interaction (FRS × HVR × YRS)
was significant across
`r n_of_total.fn(abs_lme$frs_m_sig, 3L)` male and
`r n_of_total.fn(abs_lme$frs_f_sig, 3L)` female
domains (all surviving FDR), while CVR~mimic~ was
significant in
`r n_of_total.fn(abs_lme$cvr_f_sig, 3L)` female
domains
`r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s only)", abs_lme$cvr_f_sig_doms) else ""`
and
`r n_of_total.fn(abs_lme$cvr_m_sig, 3L)` male —
suggesting that while FRS effects on hippocampal
trajectories are primarily age-driven, residual
cardiovascular burden may still modulate the
HVR-cognition relationship through pathways
independent of accelerated atrophy.

## FRS Age-Confounding and CVR~mimic~

CVR~mimic~ successfully removes the age component
from cardiovascular risk (near-zero age correlation;
@fig-cvr-validation C), supporting its use as an
age-independent CVR index for disambiguating
age-driven vs CVR-driven brain effects. The moderate
CVR~mimic~-FRS correlation reflects shared
cardiovascular indicators while confirming the
measures capture distinct constructs — one
age-confounded, one age-orthogonal.

The FRS formula explicitly includes age with large
coefficients (3.06 men, 2.33 women in log-hazard
scale) [@dagostino_general_2008], and in older
samples age dominates the FRS score
[@koller_development_2012; @pencina_quantifying_2019].
FRS values in ADNI are compressed into the upper
range (see @fig-cvr-validation, panels A-B),
limiting score variance and statistical sensitivity,
particularly in females. When age is partialled out
(CVR~mimic~), the indirect pathway
(CVR → HVR → Cognition) disappears — the a-path is
null, suggesting the FRS-hippocampal association in
this sample is primarily carried by the age component
of FRS rather than by cardiovascular risk factors
per se. This does not invalidate FRS as a clinical
tool — age is itself a potent marker of cumulative
vascular exposure
[@villeneuve_impact_2009; @dagostino_general_2008] —
but it does mean that FRS-brain associations in
elderly cohorts cannot be straightforwardly
attributed to modifiable cardiovascular factors.

The negative total cholesterol loading (see
@tbl-mimic-loadings) likely reflects widespread
statin use in this elderly clinical sample: treated
individuals have lower cholesterol despite elevated
underlying cardiovascular risk, inverting the
expected direction of association with the latent
CVR factor. Common-sample standardization (z-scoring
both FRS and CVR~mimic~ on the same N with complete
data for both measures) ensures regression
coefficients express effects per 1 SD change on an
identical scale — without this step, the a-path
dissociation could reflect scaling artifacts rather
than substantive predictor properties. Prior studies
reporting FRS-brain relationships in elderly samples
[@zsoldos_association_2020; @seshadri_stroke_2004; @song_associations_2020; @meusel_cardiovascular_2019]
may therefore reflect a mixture of age-driven and
cardiovascular effects that is difficult to
disentangle without age-adjustment.

## Dissociating Moderation from Mediation

As outlined in the analytical roadmap, the two
modelling frameworks test complementary questions.
The LME 3-way interaction
(YRS × CVR × HVR → Cognition) tests moderation:
does CVR modify how the HVR-cognition relationship
evolves over time? The LGCM a-path
(CVR → HVR slope; @fig-mediation) tests mediation:
does CVR affect cognition through hippocampal
decline?

A different pattern emerged for CVR~mimic~: the
LGCM a-path was null across all six models,
suggesting age-adjusted CVR does not accelerate
hippocampal atrophy. Yet the LME 3-way interaction
was significant in
`r n_of_total.fn(abs_lme$cvr_f_sig, 3L)` female
domains
`r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s)", abs_lme$cvr_f_sig_doms) else ""`
only, with
`r n_of_total.fn(abs_lme$cvr_m_sig, 3L)` significant
in males. This dissociation represents moderation
without mediation through hippocampal volume:
CVR~mimic~ does not accelerate hippocampal atrophy
(a-path null), yet it may modify how hippocampal
integrity relates to cognitive performance over
time. This pattern is consistent with a cognitive
reserve framework
[@stern_cognitive_2012; @rabin_interactive_2018; @iadecola_framingham_2020]:
cardiovascular risk may reduce the brain's capacity
to maintain cognitive function at a given level of
structural integrity, without directly driving
structural decline. This interpretation is further
supported by the significant CVR~mimic~ direct
effect (c') for Female Memory (@fig-mediation),
which indicates CVR~mimic~ can influence cognitive
decline through a non-hippocampal pathway even when
the a-path is null. However, the CVR~mimic~ LME
evidence rests on a single domain in one sex, and
this isolated finding should be interpreted
cautiously until replicated.

The HVR-cognition coupling (b-path) was consistent
across CVR predictors, in line with prior LGCM
coupling studies
[@muniz-terrera_latent_2017; @bollen_latent_2006].

## Sex Differences

CVR~mimic~ LME effects reached significance in
`r n_of_total.fn(abs_lme$cvr_m_sig, 3L)` male
domains and
`r n_of_total.fn(abs_lme$cvr_f_sig, 3L)` female
domains
`r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s)", abs_lme$cvr_f_sig_doms) else ""`.
The female CVR~mimic~ Executive finding is notable as
the only CVR~mimic~ LME effect observed in either sex
and warrants targeted investigation in independent
samples. Female FRS variance compression limits
statistical sensitivity — the FRS findings in females
may be primarily driven by age-weighting. Sex
differences in cardiovascular risk profiles
[@ferretti_sex_2018; @mielke_consideration_2022],
differential APOE ε4 effects, and hormonal
influences on vascular regulation may all contribute
to the observed sex asymmetry in CVR-brain
relationships. Covariates (age, education, APOE ε4)
exhibited expected effects across all models
(Appendix B, @tbl-lme-covariates and
@tbl-cvr-lme-covariates).

## HVR-Cognitive Coupling

The HVR slope → Cognitive slope relationship
(b-path) was consistent regardless of whether FRS or
CVR~mimic~ was the predictor (@fig-mediation).
The b-path was significant in
`r n_of_total.fn(abs_stats$frs_b_sig, 6L)` FRS and
`r n_of_total.fn(abs_stats$cvr_b_sig, 6L)`
CVR~mimic~ models, supporting HVR decline as
a reliable marker of cognitive decline
[@kramer_longitudinal_2007; @oschwald_brain_2020]
and extending earlier evidence that HVR captures
hippocampal integrity more sensitively than raw
volume [@schoemaker_hippocampal-ventricle_2019;
@fernandez-lozano_enhanced_2025;
@bruno_hippocampal_2016].

Parallel process coupling showed a complementary
pattern: significant in all three male models but
only Executive Function in females (one of three).
This sex asymmetry likely reflects lower statistical
power in the smaller female subsample rather than a
biological absence of coupling, given that mediation
b-paths (which condition on a predictor) were
significant in
`r n_of_total.fn(abs_stats$frs_b_sig, 6L)` models
for both sexes. Language mediation b-paths were
non-significant for both sexes, though Language
parallel process coupling reached significance in
males (consistent with @sec-slope-coupling) but not
females — a pattern compatible with language being
a later-declining domain with less variance to
capture at this disease stage. The
hippocampal-cognitive coupling is not an artifact of
CVR measurement — it persists across predictor
specifications and analysis frameworks.

## Methodological Considerations

The fixed-quadratic LGCM captures population-level
curvature but constrains individual quadratic
variation, so slope estimates primarily reflect
average rates of change. This makes the a-path null
finding conservative — if CVR~mimic~ had
individual-level non-linear effects on hippocampal
trajectories, these would require a fully random
quadratic model to detect, further strengthening the
conclusion that the FRS a-path reflects
age-weighting.

EDT linearity tests confirmed significant non-linear
trajectories across all models (Appendix B,
@tbl-linearity), motivating the fixed-quadratic
specification [@newsom_longitudinal_2015]. Fully
random quadratic models failed to converge due to
near-zero quadratic variance. EDT as the time metric
accounts for heterogeneous disease progression,
though it introduces progmod dependence.

Sensitivity analyses (Appendix C) showed HC z-score
effects were weaker than HVR effects
(@tbl-hc-comparison), attrition patterns suggest
observed effects may underestimate true relationships
(@tbl-attrition-baseline), and raw-volume results
were concordant with z-scored analyses.

## Clinical Implications

In clinical trials targeting cardiovascular risk
reduction for dementia prevention, FRS remains a
valid summary of overall vascular burden — including
the age-driven component that itself reflects
cumulative exposure
[@villeneuve_impact_2009; @dagostino_general_2008].
However, when the research question concerns
cardiovascular effects independent of aging,
age-adjusted CVR measures such as CVR~mimic~ provide
a necessary complement
[@ko_calibration_2020; @zhang_bayesian_2024]. HVR
provides a more sensitive neuroimaging endpoint than
hippocampal volume alone, and its consistent
cognitive coupling supports use as a surrogate
outcome in intervention studies. The isolated
CVR~mimic~ Executive Function finding (one of six tests,
non-significant after FDR) requires replication
before clinical interpretation.

## Limitations

This study used a single cohort (ADNI), and
replication in other aging samples is needed.
UKB-trained GAMLSS normative models showed a
systematic negative z-score offset in the ADNI
comparable subsample due to age distribution
mismatch; within-ADNI relative comparisons remain
valid, but absolute z-scores should be interpreted
cautiously across cohorts. CVR~mimic~ indicators
were limited to available CVRF data; additional
indicators could improve the latent CVR measure.
CVR was measured cross-sectionally, and the
amyloid-positive sample limits generalizability to
individuals without AD pathology.

Our coefficient comparison tests share the same
resampling framework as primary mediation inference,
with no independent validation.
Individually-varying time scores preclude CFI, TLI,
and RMSEA computation — model adequacy was assessed
via information criteria and nested LRTs only. The
MIMIC model used only 5 indicators; richer models
may yield different CVR~mimic~ values. Bootstrap
mediation CIs (N = `r med_n.lst$Male` males,
`r med_n.lst$Female` females) may be liberal with
these modest sample sizes.

## Conclusions

How CVR is measured determines whether it appears
to influence hippocampal-cognitive trajectories: FRS
produces significant indirect effects through
hippocampal decline, but age-adjusted CVR~mimic~
does not. Because age is both a legitimate marker of
cumulative vascular exposure and an independent
driver of neurodegeneration
[@villeneuve_impact_2009], the most parsimonious
explanation is that FRS-hippocampal associations in
this elderly sample are primarily carried by the age
component of the score. Suggestive single-domain evidence
(Female Executive LME 3-way) hints that age-adjusted
CVR may still modulate the brain-cognition
relationship through reserve-like mechanisms. The
hippocampal-cognitive coupling persists regardless of
CVR measurement, supporting HVR as a reliable
biomarker target.

Future directions include replication in independent
aging cohorts (e.g., WRAP, AIBL), longitudinal CVR
measurement to capture within-person CVR
trajectories, and intervention studies testing
whether CVR reduction preserves the HVR-cognition
coupling. Beyond cardiovascular risk, the
MIMIC-based age-adjustment framework presented here
applies to any composite score whose constituent
parts include a variable that independently predicts
the outcome.

---

# Ethics Approval Statement {.unnumbered}

Data used in preparation of this article were
obtained from the ADNI database. ADNI was approved by
the Institutional Review Boards of all participating
institutions. Written informed consent was obtained
from all participants.

# Author Contributions {.unnumbered}

S.F-L. conceived the study, developed the analysis
pipeline, performed all analyses, and wrote the
manuscript. S.V. provided critical review of the
study design and interpretation. D.L.C. supervised
the project, provided critical review, and approved
the final manuscript.

# Acknowledgments {.unnumbered}

Data collection and sharing for this project was
funded by the Alzheimer's Disease Neuroimaging
Initiative (ADNI) (National Institutes of Health
Grant U01 AG024904) and DOD ADNI (Department of
Defense award number W81XWH-12-2-0012). ADNI is
funded by the National Institute on Aging, the
National Institute of Biomedical Imaging and
Bioengineering, and through generous contributions
from the following: Alzheimer's Association;
Alzheimer's Drug Discovery Foundation; Araclon
Biotech; BioClinica, Inc.; Biogen Idec Inc.;
Bristol-Myers Squibb Company; Eisai Inc.; Elan
Pharmaceuticals, Inc.; Eli Lilly and Company;
EuroImmun; F. Hoffmann-La Roche Ltd and its
affiliated company Genentech, Inc.; Fujirebio; GE
Healthcare; IXICO Ltd.; Janssen Alzheimer
Immunotherapy Research & Development, LLC.; Johnson &
Johnson Pharmaceutical Research & Development LLC.;
Medpace, Inc.; Merck & Co., Inc.; Meso Scale
Diagnostics, LLC.; NeuroRx Research; Neurotrack
Technologies; Novartis Pharmaceuticals Corporation;
Pfizer Inc.; Piramal Imaging; Servier; Synarc Inc.;
and Takeda Pharmaceutical Company. The Canadian
Institutes of Health Research is providing funds to
support ADNI clinical sites in Canada. Private sector
contributions are facilitated by the Foundation for
the National Institutes of Health (www.fnih.org). The
grantee organization is the Northern California
Institute for Research and Education, and the study
is coordinated by the Alzheimer's Disease Cooperative
Study at the University of California, San Diego.
ADNI data are disseminated by the Laboratory for
Neuro Imaging at the University of Southern
California.

*Data used in preparation of this article were
obtained from the ADNI database
(adni.loni.usc.edu). As such, the investigators
within the ADNI contributed to the design and
implementation of ADNI and/or provided data but did
not participate in analysis or writing of this
report. A complete listing of ADNI investigators can
be found at:
http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf*

# Conflict of Interest Statement {.unnumbered}

The authors declare no conflicts of interest.

# Data Availability Statement {.unnumbered}

This research was conducted using ADNI data
(adni.loni.usc.edu). Data are available to qualified
researchers through the ADNI data sharing platform.
Analysis code is available at
https://github.com/soffiafdz/hvr-cvr-mediation.

\newpage

# Supplementary Material {.unnumbered}

Appendices A--C are provided as supplementary
material and contain additional tables, figures, and
analyses referenced in the main text.

:::{=latex}
\appendix
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}
:::

# Normative Transfer Validation


For transfer validation, we identified a comparable
ADNI subsample of `r adni_demo$n_total`
amyloid-negative cognitively unimpaired participants
(`r sprintf("%.0f%%", adni_demo$pct_female)` female)
within the UKB age range. Despite matching on amyloid
status and cognitive normality, the ADNI comparable
subsample was approximately
`r sprintf("%.0f", adni_demo$age_mean - ukb_demo$age_mean)`
years older on average (mean
`r sprintf("%.1f", adni_demo$age_mean)` vs
`r sprintf("%.1f", ukb_demo$age_mean)` years;
@fig-normative-validation A), concentrating in the
upper tail of the UKB age distribution. HVR z-scores
showed a systematic negative offset (mean =
`r sprintf("%.2f", comp_hvr$mean)`, SD =
`r sprintf("%.2f", comp_hvr$sd)`, 95% CI
[`r sprintf("%.2f", comp_hvr$ci_95["lower"])`,
`r sprintf("%.2f", comp_hvr$ci_95["upper"])`]),
while HC z-scores were closer to expected values
(mean = `r sprintf("%.2f", comp_hc$mean)`, SD =
`r sprintf("%.2f", comp_hc$sd)`;
@fig-normative-validation B). Age-bin-specific
calibration (@fig-normative-validation, panel B)
showed mean z-scores near zero in younger bins with
increasingly negative offset in older bins,
consistent with the age mismatch.

```{r fig-normative-validation}
#| fig-cap: "**Normative transfer validation.** (A) Age distributions: UK Biobank normative sample (yellow) vs ADNI comparable subsample (cyan); dashed lines indicate sample means. (B) Forest plot of HVR (purple) and HC (green) z-scores (mean ± 95% CI) by age bin; reference line at z = 0 indicates perfect calibration; open circles = age bins, filled circles = overall. (C) Z-score distributions for HVR (left) and HC (right) compared to the expected N(0,1) (yellow)."
#| fig-width: 6.5
#| fig-height: 6

stopifnot(!is.null(nv_data))
ukb_d <- nv_data$comparable_subsample$demographics$ukb_normative
adni_d <- nv_data$comparable_subsample$demographics$adni_comparable

p_age <- plot_age_distribution_comparison(
  ukb_age_mean = ukb_d$age_mean,
  ukb_age_sd = ukb_d$age_sd,
  ukb_age_min = ukb_d$age_min,
  ukb_age_max = ukb_d$age_max,
  ukb_n = ukb_d$n_total,
  adni_age_mean = adni_d$age_mean,
  adni_age_sd = adni_d$age_sd,
  adni_n = adni_d$n_total
)

p_forest <- plot_zscore_validation_forest(nv_data)

# Build z-score data.table with both measures
hvr_z.v <- comp_hvr$zscores
hc_z.v <- comp_hc$zscores
stopifnot(!is.null(hvr_z.v), !is.null(hc_z.v))
n_max <- max(length(hvr_z.v), length(hc_z.v))
zscore.dt <- data.table(
  PTID = seq_len(n_max),
  HVR_z = hvr_z.v[seq_len(n_max)],
  HC_z = hc_z.v[seq_len(n_max)]
)

p_dist <- plot_zscore_distribution(
  zscore.dt, sample_label = "Comparable"
)

(p_age | p_forest) / p_dist +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(tag_levels = "A")
```


:::{=latex}
\FloatBarrier
\newpage
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{B\arabic{table}}
\renewcommand{\thefigure}{B\arabic{figure}}
:::

# Methodological Validations

## CVR MIMIC Measurement Invariance

Measurement invariance of the CVR MIMIC model was
tested across sex in four steps (configural, metric,
scalar, strict; @tbl-mi).

```{r tbl-mi}
#| tbl-cap: "**CVR MIMIC measurement invariance.** Configural = same structure; Metric = equal loadings; Scalar = equal intercepts; Strict = equal residuals. ΔCFI < 0.01 and ΔRMSEA < 0.015 indicate acceptable invariance."

stopifnot(!is.null(cvr_mi.res))
require_output(
  create_mi_table(
    cvr_mi.res,
    "Measurement Invariance Across Sex"
  ),
  "tbl-mi"
)
```

## EDT Linearity and Simulation Validation

The linear time specification was validated via LME
likelihood ratio tests comparing linear vs quadratic
EDT across all sex-domain models (see
@tbl-linearity).

```{r tbl-linearity}
#| tbl-cap: "**EDT linearity validation.** LME models testing linear vs quadratic EDT specification. LRT tests whether adding a quadratic EDT term significantly improves fit. ΔAIC = AIC(linear) − AIC(quadratic); positive values favor quadratic. EDT = Estimated Disease Time."

require_output(
  create_linearity_table(
    edt_linearity,
    title = "EDT Linearity Validation"
  ),
  "tbl-linearity"
)
```

```{r edt-linearity-narrative}
#| include: false

stopifnot(!is.null(edt_linearity))
lt <- as.data.table(
  edt_linearity$results_table
)
n_sig <- sum(
  lt$LRT_p < 0.05, na.rm = TRUE
)
n_total <- nrow(lt)
if (n_sig == 0) {
  edt_lin_text <- paste0(
    "No domain showed a significant",
    " quadratic EDT term",
    " (all p > 0.05), supporting the",
    " linear time specification for",
    " the LGCM."
  )
} else {
  sig_doms <- lt[LRT_p < 0.05, Domain]
  max_chisq <- max(
    lt$LRT_chisq, na.rm = TRUE
  )
  edt_lin_text <- sprintf(
    paste0(
      "%s models showed significant",
      " quadratic EDT terms",
      " (%s; max χ² = %.1f).",
      " Large test statistics reflect",
      " substantial power afforded by",
      " sample size and repeated",
      " observations. A linear",
      " specification was adopted as a",
      " pragmatic compromise given",
      " convergence failure of quadratic",
      " parallel process LGCMs",
      " (see Methods)."
    ),
    sub("^(.)", "\\U\\1",
      n_of_total.fn(n_sig, n_total),
      perl = TRUE
    ),
    paste(
      DOMAIN_LABELS[sig_doms],
      collapse = ", "
    ),
    max_chisq
  )
}
```

`r edt_lin_text`

## SE-Constrained LGCM Validation

A Monte Carlo simulation
(`r simulation.res$config$n_reps` replications at
N = `r paste(simulation.res$config$sample_sizes, collapse = " and N = ")`)
compared SE-constrained LGCM (residual variance
fixed to SE²) against free estimation.

```{r simulation-interpretation}
#| include: false

stopifnot(!is.null(sim_conv))
fmt_sci <- function(x) {
  if (abs(x) < 1e-10) return("0")
  exp <- floor(log10(abs(x)))
  mant <- x / 10^exp
  sprintf("%.1f × 10^%d^", mant, exp)
}
sim_text_parts <- vapply(
  seq_len(nrow(sim_conv)), function(i) {
    n_reps <- simulation.res$config$n_reps
    sprintf(
      paste0(
        "At N = %d, the SE-constrained",
        " approach converged in %d of %d",
        " replications (free estimation:",
        " %d of %d) with bias = %s",
        " (free: %s), RMSE = %.3f",
        " (free: %.3f), and 95%% CI",
        " coverage = %.1f%%",
        " (free: %.1f%%)"
      ),
      sim_conv$n[i],
      sim_conv$n_converged_se[i], n_reps,
      sim_conv$n_converged_free[i], n_reps,
      fmt_sci(sim_conv$bias_se[i]),
      fmt_sci(sim_conv$bias_free[i]),
      sim_conv$rmse_se[i],
      sim_conv$rmse_free[i],
      sim_conv$coverage_se[i] * 100,
      sim_conv$coverage_free[i] * 100
    )
  }, character(1)
)
sim_text <- paste(
  sim_text_parts, collapse = ". "
)
```

`r sim_text`. Both approaches showed comparable
performance, validating the SE-constrained
specification used in the primary analyses.

## LME Covariate Effects

Full covariate effects from the sex-stratified
LME 3-way interaction models are presented for
the FRS model (@tbl-lme-covariates) and the
CVR~mimic~ model (@tbl-cvr-lme-covariates).

```{r tbl-lme-covariates}
#| tbl-cap: "**FRS LME covariate effects.** Sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    frs_hvr, cvr_label = "FRS",
    title = "Covariate Effects (FRS Model)"
  ),
  "tbl-lme-covariates"
)
```

```{r tbl-cvr-lme-covariates}
#| tbl-cap: "**CVR~mimic~ LME covariate effects.** Sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    cvr_hvr,
    cvr_label = "CVR~mimic~",
    title = md(
      "Covariate Effects (CVR~mimic~ Model)"
    )
  ),
  "tbl-cvr-lme-covariates"
)
```


:::{=latex}
\FloatBarrier
\newpage
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{C\arabic{table}}
\renewcommand{\thefigure}{C\arabic{figure}}
:::

# Sensitivity Analyses

All sensitivity checks converge on the same
pattern: FRS effects trace to age-weighting,
CVR~mimic~ effects are null or isolated, and the
HVR–cognition coupling (b-path) is consistent. Main
findings are stable across alternative brain
measures (HC volume), raw scores, attrition
adjustments, and EDT specification checks. Full
details follow.

## LME Sensitivity

### HC z-score (Total Hippocampal Volume)

HC z-score effects were generally weaker than HVR
z-score effects, with fewer significant 3-way
interactions, supporting the
hippocampus-to-ventricle ratio as a more sensitive
brain aging biomarker. The LME 3-way interaction was
repeated with HC z-score instead of HVR; results
appear in @tbl-hc-comparison, with covariate effects
in @tbl-hc-covariates and @tbl-hc-covariates-cvr.

```{r tbl-hc-comparison}
#| tbl-cap: "**HC z-score sensitivity analysis.** LME 3-way interaction effects (YRS × CVR × HC z-score) for total hippocampal volume. FRS and CVR~mimic~ compared side-by-side. Bold indicates p < 0.05."

require_output(
  create_lme_comparison_table(
    frs_hc, cvr_hc,
    summary_dt = hc_summary.dt
  ),
  "tbl-hc-comparison"
)
```

```{r tbl-hc-covariates}
#| tbl-cap: "**FRS HC z-score covariate effects.** Sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    frs_hc, cvr_label = "FRS",
    title = "HC Covariate Effects (FRS Model)"
  ),
  "tbl-hc-covariates"
)
```

```{r tbl-hc-covariates-cvr}
#| tbl-cap: "**CVR~mimic~ HC z-score covariate effects.** Sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    cvr_hc, cvr_label = "CVR~mimic~",
    title = md(
      "HC Covariate Effects (CVR~mimic~ Model)"
    )
  ),
  "tbl-hc-covariates-cvr"
)
```

```{r hc-interpretation}
#| include: false

stopifnot(!is.null(hc_summary.dt))
hc_p <- hc_summary.dt[
  Analysis == "Primary_Stratified"
]
frs_hc_sig <- hc_p[
  CVR_Measure == "FRS" & Significant == TRUE
]
cvr_hc_sig <- hc_p[
  CVR_Measure == "CVR_mimic" &
    Significant == TRUE
]
frs_hc_desc <- if (nrow(frs_hc_sig) > 0) {
  paste(
    sprintf(
      "%s %s",
      frs_hc_sig$Sex,
      DOMAIN_LABELS[frs_hc_sig$Domain]
    ),
    collapse = " and "
  )
} else "no sex–domain combination"
cvr_hc_desc <- if (nrow(cvr_hc_sig) > 0) {
  paste(
    sprintf(
      "%s %s",
      cvr_hc_sig$Sex,
      DOMAIN_LABELS[cvr_hc_sig$Domain]
    ),
    collapse = " and "
  )
} else "any sex–domain combination"
hc_interp_text <- sprintf(
  paste0(
    "The FRS × HC × YRS interaction was",
    " significant for %s, whereas",
    " CVR~mimic~ × HC × YRS %s"
  ),
  frs_hc_desc,
  if (nrow(cvr_hc_sig) > 0) {
    sprintf(
      "reached significance only for %s",
      cvr_hc_desc
    )
  } else {
    paste0(
      "did not reach significance for ",
      cvr_hc_desc
    )
  }
)
```

`r hc_interp_text`. HC z-score showed weaker effects than HVR z-score, supporting use of the ratio measure.

```{r hc-vs-hvr-synthesis}
#| include: false

hc_p2 <- hc_summary.dt[
  Analysis == "Primary_Stratified"
]
hvr_p <- lme_summary.dt[
  Analysis == "Primary_Stratified"
]
n_hc_sig <- sum(
  hc_p2$Significant, na.rm = TRUE
)
n_hvr_sig <- sum(
  hvr_p$Significant, na.rm = TRUE
)
hc_synth_text <- sprintf(
  paste0(
    "Significant 3-way interactions emerged",
    " in %s models for HC z-score versus",
    " %s for HVR z-score. The weaker HC",
    " effects support the specificity of",
    " HVR: capturing ventricular expansion",
    " alongside hippocampal atrophy",
    " provides greater sensitivity to",
    " CVR moderation."
  ),
  n_of_total.fn(n_hc_sig, 12L),
  n_of_total.fn(n_hvr_sig, 12L)
)
```

`r hc_synth_text`

### Attrition Analysis

Baseline characteristics were compared between
completers (≥4 visits) and dropouts (<4 visits);
results appear in @tbl-attrition-baseline.

```{r tbl-attrition-baseline}
#| tbl-cap: "**Attrition analysis.** Baseline characteristics of completers (≥4 visits) vs dropouts (<4 visits). Bold rows indicate p < 0.05. Completers = participants with sufficient visits for quadratic trend estimation."

require_output(
  create_attrition_table(attrition_analysis),
  "tbl-attrition-baseline"
)
```

```{r attrition-interpretation}
#| include: false

stopifnot(
  !is.null(attrition_analysis),
  !is.null(attrition_analysis$comparison_table)
)
comp_dt <- attrition_analysis$comparison_table
sig_vars <- comp_dt[p_value < 0.05, Variable]
var_display <- c(
  AGE = "age", EDUC = "education",
  FRS = "FRS", HVR_z = "HVR z-score",
  HC_z = "HC z-score", SEX = "sex",
  APOE4 = "APOE ε4 carrier status"
)
sig_labels <- vapply(
  sig_vars, function(v) {
    if (v %in% names(var_display)) {
      var_display[v]
    } else v
  }, character(1)
)
if (length(sig_vars) > 0) {
  attrition_text <- sprintf(
    paste0(
      "Significant baseline differences",
      " between completers and dropouts",
      " were observed in %s (all p < 0.05;",
      " see @tbl-attrition-baseline)."
    ),
    paste(sig_labels, collapse = ", ")
  )
} else {
  attrition_text <- paste0(
    "No significant baseline",
    " differences between completers",
    " and dropouts were observed",
    " (see @tbl-attrition-baseline),",
    " suggesting attrition was random",
    " with respect to measured variables."
  )
}
```

`r attrition_text`

```{r attrition-logistic-inline}
#| include: false

stopifnot(
  !is.null(attrition_analysis),
  !is.null(attrition_analysis$logistic)
)
log_model <- attrition_analysis$logistic
or_dt <- as.data.table(log_model$odds_ratios)
setnames(
  or_dt,
  c("2.5 %", "97.5 %"),
  c("CI_lower", "CI_upper"),
  skip_absent = TRUE
)
or_dt <- or_dt[term != "(Intercept)"]
or_dt[, sig := CI_lower > 1 | CI_upper < 1]

# Readable term labels
term_display <- c(
  FRS = "FRS", AGE = "age",
  APOE4TRUE = "APOE ε4 carrier status",
  SEXMale = "male sex",
  HVR_z = "HVR z-score",
  HC_z = "HC z-score"
)
or_dt[, term_label := fifelse(
  term %in% names(term_display),
  term_display[term], term
)]

sig_preds <- or_dt[sig == TRUE]
if (nrow(sig_preds) > 0) {
  items <- sprintf(
    "%s (OR = %.2f, 95%% CI [%.2f, %.2f])",
    sig_preds$term_label,
    sig_preds$OR,
    sig_preds$CI_lower,
    sig_preds$CI_upper
  )
  logistic_text <- sprintf(
    paste0(
      "Logistic regression predicting",
      " dropout identified %d significant",
      " predictor%s — %s."
    ),
    nrow(sig_preds),
    ifelse(nrow(sig_preds) > 1, "s", ""),
    paste(items, collapse = "; ")
  )
} else {
  logistic_text <- paste0(
    "Logistic regression predicting",
    " dropout identified no significant",
    " predictors (all 95% CIs included 1)."
  )
}

# FRS direction narrative
comp_dt <- attrition_analysis$comparison_table
frs_row <- comp_dt[
  grepl("FRS", Variable, ignore.case = TRUE)
]
frs_dir_text <- ""
if (nrow(frs_row) > 0) {
  c_mean <- frs_row$Completers_mean[1]
  d_mean <- frs_row$Dropouts_mean[1]
  if (!is.null(c_mean) && !is.null(d_mean) &&
      !is.na(c_mean) && !is.na(d_mean)) {
    direction <- if (d_mean > c_mean) {
      "higher"
    } else "lower"
    frs_dir_text <- sprintf(
      paste0(
        " Dropouts had %s baseline FRS",
        " (M = %.2f) than completers",
        " (M = %.2f), consistent with",
        " selective loss of higher-risk",
        " participants — this may",
        " attenuate observed CVR effects."
      ),
      direction, d_mean, c_mean
    )
  }
}
```

`r logistic_text``r frs_dir_text`

### Raw Brain Measure Sensitivity

The LME 3-way interaction was repeated using raw
(un-normalized) HVR and raw HC (with total
intracranial volume [TIV] covariate),
with FRS as predictor, full-sample, no sex
stratification, and no YRS² term.

```{r raw-hvr-inline}
#| include: false

stopifnot(
  !is.null(lme_raw_sens.res),
  !is.null(lme_raw_sens.res$results)
)
hvr_raw <- lme_raw_sens.res$results$HVR_raw_FRS
stopifnot(!is.null(hvr_raw))
raw_hvr_results <- list(results = hvr_raw)

hc_raw <- lme_raw_sens.res$results$HC_raw_FRS
stopifnot(!is.null(hc_raw))
raw_hc_results <- list(results = hc_raw)

fmt_raw_sig <- function(res) {
  sig.v <- c()
  nonsig.v <- c()
  for (d in c("MEM", "LAN", "EXF")) {
    r <- res$results[[d]]$interaction
    if (is.null(r)) next
    entry <- sprintf(
      "%s (β = %.4f, %s)",
      DOMAIN_LABELS[d], r$beta,
      fmt_p_inline(r$p_value)
    )
    if (r$p_value < 0.05) {
      sig.v <- c(sig.v, entry)
    } else {
      nonsig.v <- c(nonsig.v, entry)
    }
  }
  list(sig = sig.v, nonsig = nonsig.v)
}
hvr_raw_split <- fmt_raw_sig(raw_hvr_results)
hc_raw_split <- fmt_raw_sig(raw_hc_results)
raw_hvr_prose <- if (
  length(hvr_raw_split$sig) == 3
) {
  paste0(
    "Raw HVR was significant for all three",
    " domains: ",
    paste(hvr_raw_split$sig, collapse = ", ")
  )
} else if (length(hvr_raw_split$sig) > 0) {
  paste0(
    "Raw HVR was significant for ",
    paste(
      hvr_raw_split$sig, collapse = " and "
    ),
    if (length(hvr_raw_split$nonsig) > 0)
      paste0(
        ", but not ",
        paste(
          hvr_raw_split$nonsig,
          collapse = " or "
        )
      )
    else ""
  )
} else {
  "Raw HVR was not significant for any domain"
}
raw_hc_prose <- if (
  length(hc_raw_split$sig) > 0 &&
    length(hc_raw_split$nonsig) > 0
) {
  paste0(
    "Raw HC (+TIV) reached significance",
    " only for ",
    paste(
      hc_raw_split$sig, collapse = " and "
    )
  )
} else if (length(hc_raw_split$sig) == 0) {
  paste0(
    "Raw HC (+TIV) was not significant",
    " for any domain"
  )
} else {
  paste0(
    "Raw HC (+TIV) was significant for ",
    paste(hc_raw_split$sig, collapse = ", ")
  )
}
raw_meas_text <- paste0(
  raw_hvr_prose, ". ", raw_hc_prose
)

# Concordance check
sig_hvr <- sapply(
  names(raw_hvr_results$results),
  function(d) {
    r <- raw_hvr_results$results[[d]]$interaction
    !is.null(r) && r$p_value < 0.05
  }
)
if (any(sig_hvr)) {
  raw_concord_text <- sprintf(
    paste0(
      "Raw HVR × FRS was significant for",
      " %s, confirming normative z-scoring",
      " does not drive primary findings."
    ),
    paste(
      DOMAIN_LABELS[names(sig_hvr)[sig_hvr]],
      collapse = ", "
    )
  )
} else {
  raw_concord_text <- paste0(
    "Raw HVR × FRS was not significant",
    " for any domain, suggesting",
    " normative z-scoring enhances",
    " sensitivity."
  )
}

# Raw HC (+TIV) note
sig_hc <- sapply(
  names(raw_hc_results$results),
  function(d) {
    r <- raw_hc_results$results[[d]]$interaction
    !is.null(r) && r$p_value < 0.05
  }
)
raw_hc_note <- if (!any(sig_hc)) {
  paste0(
    " Raw HC (+TIV) was not significant",
    " for any domain, reinforcing the",
    " added sensitivity of the ratio",
    " measure."
  )
} else ""
```

`r raw_meas_text`. `r raw_concord_text``r raw_hc_note`

### EDT Moderation of LME Effects

A 4-way interaction (CVR × HVR × EDT × YRS) tested
whether CVR effects vary by disease stage.

```{r edt-moderation-inline}
#| include: false

build_edt_text <- function(edt_res, cvr_label) {
  stopifnot(
    !is.null(edt_res),
    !is.null(edt_res$summary_continuous)
  )
  sc <- as.data.table(edt_res$summary_continuous)
  parts <- vapply(
    c("Male", "Female"), function(sx) {
      sx_rows <- sc[Sex == sx]
      sig <- sx_rows[Significant == TRUE]
      if (nrow(sig) > 0) {
        items <- sprintf(
          "%s (β = %.4f, %s)",
          DOMAIN_LABELS[sig$Domain],
          sig$Beta,
          sapply(sig$p_value, fmt_p_inline)
        )
        sprintf(
          paste0(
            "%s × HVR × EDT (%s):",
            " significant for %s"
          ),
          cvr_label, sx,
          paste(items, collapse = "; ")
        )
      } else {
        min_p <- min(
          sx_rows$p_value, na.rm = TRUE
        )
        sprintf(
          paste0(
            "%s × HVR × EDT (%s):",
            " not significant (min %s)"
          ),
          cvr_label, sx,
          fmt_p_inline(min_p)
        )
      }
    }, character(1)
  )
  paste(parts, collapse = ". ")
}
edt_mod_text <- paste(
  build_edt_text(frs_edt, "FRS"),
  build_edt_text(cvr_edt, "CVR~mimic~"),
  sep = ". "
)
```

`r edt_mod_text`. Limited EDT moderation suggests CVR effects on HVR-cognition coupling are relatively stable across disease stages.

### EDT Stage Sensitivity

Analyses were restricted to early and middle EDT
tertiles (excluding late-stage) to confirm effect
stability.

```{r edt-sensitivity-inline}
#| include: false

stopifnot(!is.null(edt_sens.res))
rt <- as.data.table(edt_sens.res$results_table)

meas_labels <- c(
  HVR_z = "HVR z-score",
  HC_z = "HC z-score"
)
dom_labels <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

edt_sens_parts <- vapply(
  unique(rt$Measure), function(meas) {
    mrt <- rt[Measure == meas]
    m_lab <- if (
      meas %in% names(meas_labels)
    ) meas_labels[meas] else meas
    items <- c()
    n_concord <- 0L
    for (i in seq_len(nrow(mrt))) {
      row <- mrt[i]
      d_lab <- if (
        row$Domain %in% names(dom_labels)
      ) dom_labels[row$Domain] else row$Domain
      pct <- if (
        !is.null(row$Beta_change_pct) &&
        !is.na(row$Beta_change_pct)
      ) {
        sprintf(
          ", %.1f%% change",
          row$Beta_change_pct
        )
      } else ""
      if (!is.null(row$Consistent) &&
          isTRUE(row$Consistent)) {
        n_concord <- n_concord + 1L
      }
      items <- c(items, sprintf(
        "%s (β = %.4f, %s%s)",
        d_lab, row$Beta_early_middle,
        fmt_p_inline(row$P_early_middle),
        pct
      ))
    }
    concord_note <- if (n_concord > 0) {
      sprintf(
        " (%d/%d concordant with full-sample)",
        n_concord, nrow(mrt)
      )
    } else ""
    sprintf(
      "%s: %s%s", m_lab,
      paste(items, collapse = "; "),
      concord_note
    )
  }, character(1)
)
edt_sens_text <- paste(
  edt_sens_parts, collapse = ". "
)
```

`r edt_sens_text`.

:::{=latex}
\FloatBarrier
:::

# References {.unnumbered}

::: {#refs}
:::
