---
title: "Age-Adjusted Cardiovascular Risk and Hippocampal-Cognitive Coupling Along the Alzheimer's Disease Continuum"
subtitle: "Disentangling FRS Age-Weighting from True CVR Effects"
author:
  - name: Sofia Fernandez-Lozano
    affiliations:
      - ref: bic
      - ref: ipn
  - name: Sylvia Villeneuve
    affiliations:
      - ref: bic
      - ref: psych
      - ref: stopad
  - name: Louis D. Collins
    affiliations:
      - ref: bic
      - ref: ipn
      - ref: bbme
    email: louis.collins@mcgill.ca
    corresponding: true
affiliations:
  - id: bic
    name: McConnell Brain Imaging Centre
    department: Montreal Neurological Institute
    city: Montreal
    state: Quebec
    country: Canada
  - id: ipn
    name: McGill University
    department: Department of Neurology and Neurosurgery
    city: Montreal
    state: Quebec
    country: Canada
  - id: psych
    name: McGill University
    department: Department of Psychiatry
    city: Montreal
    state: Quebec
    country: Canada
  - id: stopad
    name: Douglas Mental Health Institute Research Centre
    department: StoP-AD Centre
    city: Montreal
    state: Quebec
    country: Canada
  - id: bbme
    name: McGill University
    department: Department of Biomedical Engineering
    city: Montreal
    state: Quebec
    country: Canada
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: false
    number-sections: true
    documentclass: article
    pdf-engine: xelatex
    geometry:
      - margin=1in
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage{tikz}
        \usetikzlibrary{shapes.geometric, arrows.meta, positioning}
        \usepackage{authblk}
        \renewcommand\Authfont{\normalsize}
        \renewcommand\Affilfont{\small\itshape}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{float}
        \floatplacement{table}{H}
        \floatplacement{figure}{H}
        \usepackage{placeins}
        \usepackage{textcomp}
        \usepackage{newunicodechar}
        \newunicodechar{≤}{\ensuremath{\leq}}
        \newunicodechar{≥}{\ensuremath{\geq}}
        \newunicodechar{×}{\ensuremath{\times}}
        \newunicodechar{β}{\ensuremath{\beta}}
        \newunicodechar{η}{\ensuremath{\eta}}
        \newunicodechar{χ}{\ensuremath{\chi}}
        \newunicodechar{σ}{\ensuremath{\sigma}}
        \newunicodechar{τ}{\ensuremath{\tau}}
        \newunicodechar{γ}{\ensuremath{\gamma}}
        \newunicodechar{ζ}{\ensuremath{\zeta}}
        \newunicodechar{φ}{\ensuremath{\varphi}}
        \newunicodechar{μ}{\ensuremath{\mu}}
        \newunicodechar{ν}{\ensuremath{\nu}}
        \newunicodechar{λ}{\ensuremath{\lambda}}
        \newunicodechar{Δ}{\ensuremath{\Delta}}
        \newunicodechar{ε}{\ensuremath{\varepsilon}}
        \newunicodechar{ρ}{\ensuremath{\rho}}
        \newunicodechar{→}{\ensuremath{\rightarrow}}
        \newunicodechar{↔}{\ensuremath{\leftrightarrow}}
        \newunicodechar{≈}{\ensuremath{\approx}}
        \newunicodechar{²}{\textsuperscript{2}}
        \setlength{\LTleft}{0pt}
        \setlength{\LTright}{\fill}
knitr:
  opts_chunk:
    dev: cairo_pdf
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| include: false

# Find project root and change to it so here() works
orig_wd <- getwd()
project_root <- normalizePath(
  file.path(orig_wd, "..")
)
if (!file.exists(
  file.path(project_root, ".here")
)) {
  project_root <- orig_wd
}
setwd(project_root)

library(here)
library(data.table)
library(gt)
library(ggplot2)
library(knitr)
library(patchwork)

source(here("R/utils/config.R"))
source(here("R/utils/data_io.R"))
source(here("R/utils/tables.R"))
source(here("R/utils/plotting.R"))
source(here("R/utils/consort_diagram.R"))

config <- load_config()

# -------------------------------------------------
# Load pre-computed manuscript environment
# -------------------------------------------------
env.lst <- readRDS(
  here("outputs/manuscript_env.rds")
)
list2env(env.lst, envir = environment())
rm(env.lst)

# -------------------------------------------------------------------
# gt table rendering for LaTeX/PDF output
# -------------------------------------------------------------------

knit_print.gt_tbl <- function(x, ...) {
  if (knitr::is_latex_output()) {
    latex_code <- as.character(gt::as_latex(x))

    # Fix 1: Table width
    latex_code <- gsub(
      "\\\\begin\\{tabular\\*\\}\\{\\\\linewidth\\}\\{@\\{\\\\extracolsep\\{\\\\fill\\}\\}",
      "\\\\begin{tabular}{",
      latex_code
    )
    latex_code <- gsub(
      "\\\\end\\{tabular\\*\\}", "\\\\end{tabular}",
      latex_code
    )

    # Fix 2: HTML tags to LaTeX
    latex_code <- gsub(
      "<em>([^<]+)</em>", "\\\\textit{\\1}", latex_code
    )
    latex_code <- gsub(
      "<strong>([^<]+)</strong>", "\\\\textbf{\\1}", latex_code
    )
    latex_code <- gsub(
      "<sub>([^<]+)</sub>", "\\\\textsubscript{\\1}", latex_code
    )
    latex_code <- gsub(
      "<sup>([^<]+)</sup>", "\\\\textsuperscript{\\1}", latex_code
    )

    # Fix 3: Unicode symbols to LaTeX math
    latex_code <- gsub("\u0394", "$\\\\Delta$", latex_code)
    latex_code <- gsub("\u03B2", "$\\\\beta$", latex_code)
    latex_code <- gsub("\u03BB", "$\\\\lambda$", latex_code)
    latex_code <- gsub("\u03C3", "$\\\\sigma$", latex_code)
    latex_code <- gsub("\u03B7", "$\\\\eta$", latex_code)
    latex_code <- gsub("\u03C7", "$\\\\chi$", latex_code)
    latex_code <- gsub("\u03C4", "$\\\\tau$", latex_code)
    latex_code <- gsub("\u03B3", "$\\\\gamma$", latex_code)
    latex_code <- gsub(
      "\u00B2", "\\\\textsuperscript{2}", latex_code
    )
    latex_code <- gsub("\u2264", "$\\\\leq$", latex_code)
    latex_code <- gsub("\u2265", "$\\\\geq$", latex_code)
    latex_code <- gsub("\u2260", "$\\\\neq$", latex_code)
    latex_code <- gsub("\u00D7", "$\\\\times$", latex_code)

    # Fix 4: Unicode text commands missing {} terminator
    latex_code <- gsub(
      "\\\\(texttimes|texttwosuperior|texttwosuperiorp|textendash|textemdash)([A-Za-z])",
      "\\\\\\1{}\\2", latex_code
    )

    # Fix 5: Footnote minipage width
    latex_code <- gsub(
      "\\\\begin\\{minipage\\}\\{\\\\linewidth\\}",
      "\\\\begin{minipage}[t]{0.8\\\\textwidth}\\\\raggedright\\\\small",
      latex_code
    )

    # Fix 6: Column alignment (first left, rest right)
    fix_alignment <- function(code) {
      pattern <- "(\\\\begin\\{(?:tabular|longtable)\\}\\{)([lcr]+)(\\})"
      m <- gregexpr(pattern, code, perl = TRUE)
      if (m[[1]][1] == -1) return(code)
      matches <- regmatches(code, m)[[1]]
      for (match in matches) {
        col_spec <- gsub(".*\\{([lcr]+)\\}$", "\\1", match)
        n_cols <- nchar(col_spec)
        if (n_cols > 1) {
          new_spec <- paste0(
            "l", paste(rep("r", n_cols - 1), collapse = "")
          )
          new_match <- gsub(
            "\\{[lcr]+\\}$", paste0("{", new_spec, "}"), match
          )
          code <- sub(match, new_match, code, fixed = TRUE)
        }
      }
      code
    }
    latex_code <- fix_alignment(latex_code)

    # Fix 7: Table spacing
    latex_code <- gsub(
      "\\\\begin\\{tabular\\}",
      "\\\\renewcommand{\\\\arraystretch}{1.15}\\\\begin{tabular}",
      latex_code
    )

    # Fix 8: Escape [ ] in table cells
    latex_code <- gsub(
      "\\\\addlinespace\\[([^]]+)\\]",
      "<<ADDLNSP_\\1_>>", latex_code
    )
    latex_code <- gsub(
      "\\\\\\\\\\[([0-9.]+pt)\\]",
      "<<LINEBRK_\\1_>>", latex_code
    )
    latex_code <- gsub(
      "\\\\begin\\{minipage\\}\\[([a-z])\\]",
      "<<MINIPAGE_\\1_>>", latex_code
    )
    latex_code <- gsub("\\[", "{[}", latex_code)
    latex_code <- gsub("\\]", "{]}", latex_code)
    latex_code <- gsub(
      "<<ADDLNSP_([^_]+)_>>",
      "\\\\addlinespace[\\1]", latex_code
    )
    latex_code <- gsub(
      "<<LINEBRK_([^_]+)_>>",
      "\\\\\\\\[\\1]", latex_code
    )
    latex_code <- gsub(
      "<<MINIPAGE_([a-z])_>>",
      "\\\\begin{minipage}[\\1]", latex_code
    )

    knitr::asis_output(latex_code)
  } else {
    gt:::knit_print.gt_tbl(x, ...)
  }
}
registerS3method(
  "knit_print", "gt_tbl", knit_print.gt_tbl,
  envir = asNamespace("knitr")
)

# Domain labels for cognitive measures
DOMAIN_LABELS <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

# Helper: Format-aware markdown (handles LaTeX vs HTML)
fmt_crossformat <- function(gt_tbl, columns, rows = everything()) {
  if (knitr::is_latex_output()) {
    gt::fmt(
      gt_tbl, columns = {{ columns }},
      rows = {{ rows }}, fns = function(x) {
        x <- gsub("\\*([^*]+)\\*", "\\\\textit{\\1}", x)
        x <- gsub("~([^~]+)~", "\\\\textsubscript{\\1}", x)
        x
      }
    )
  } else {
    gt::fmt_markdown(
      gt_tbl, columns = {{ columns }}, rows = {{ rows }}
    )
  }
}
```

---

# Abstract {.unnumbered}

**INTRODUCTION:**
The Framingham Risk Score (FRS) is widely used to
index cardiovascular risk (CVR) in aging research,
yet its heavy age-weighting may confound
associations with brain and cognitive outcomes.

**METHODS:**
In `r n_subjects` amyloid-positive ADNI participants,
we compared FRS with an age-adjusted CVR measure
(CVR~mimic~) from a MIMIC model. Sex-stratified
LME 3-way interaction models and LGCM mediation
models tested CVR effects on hippocampal-cognitive
trajectories.

**RESULTS:**
FRS predicted hippocampal atrophy rate
(CVR → HVR slope significant in
`r abs_stats$frs_a_sig`/6 LGCM mediation
models); CVR~mimic~ did not
(`r abs_stats$cvr_a_sig`/6). HVR-cognitive
coupling was significant in
`r abs_stats$frs_b_sig`/6 FRS and
`r abs_stats$cvr_b_sig`/6 CVR~mimic~ models,
persisting regardless of CVR measure. FRS
showed significant indirect effects in
`r abs_stats$frs_ind_sig`/6 models versus
`r abs_stats$cvr_ind_sig`/6 for CVR~mimic~.

**DISCUSSION:**
FRS "effects" on hippocampal-cognitive decline
primarily reflect age-related variance rather
than cardiovascular pathophysiology. Age-adjusted
CVR measures should be preferred in elderly
cohorts.

**Keywords:** Cardiovascular risk, Framingham Risk
Score, hippocampus, cognitive decline, Alzheimer's
disease, MIMIC model, age-confounding

## Research in Context {.unnumbered}

**Systematic Review:**
The Framingham Risk Score (FRS) has been linked to
reduced brain volumes and accelerated cognitive
decline in multiple cohorts
[@zsoldos_association_2020; @seshadri_stroke_2004;
@song_associations_2020; @meusel_cardiovascular_2019]. However, FRS includes age as a
heavily weighted component
[@dagostino_general_2008; @koller_development_2012], and age accounts for
the majority of FRS predictive performance in older
adults [@pencina_quantifying_2019; @ko_calibration_2020], raising
the concern that apparent FRS-brain associations in
elderly samples may reflect age-driven confounding.
No prior study has directly compared an
age-confounded composite (FRS) against an
age-adjusted latent CVR measure within the same
mediation and moderation framework.

**Interpretation:**
FRS indirect effects on cognitive decline through
hippocampal atrophy primarily reflect its
age-weighting. Partialling out age via a MIMIC
model (CVR~mimic~) nullified the CVR → hippocampal
atrophy pathway. In contrast, hippocampal-cognitive
coupling remained significant in most domains,
indicating it reflects neurodegeneration rather
than cardiovascular risk. This dissociation suggests
prior FRS-brain reports may partly capture age
effects rather than cardiovascular pathophysiology.

**Future Directions:**
Replication in independent aging cohorts (e.g.,
WRAP, AIBL) with longitudinal CVR measurement is
needed. The MIMIC-based age-adjustment framework
demonstrated here can be applied to any composite
risk score where age confounding is a concern.
Intervention studies should test whether CVR
reduction preserves hippocampal-cognitive coupling,
using age-adjusted CVR endpoints rather than FRS.

---

# Introduction

## Background

- Cardiovascular risk factors are increasingly recognized as
  modifiable contributors to cognitive decline and dementia risk
  [@livingston_dementia_2017; @ouldred_vascular_2010; @viswanathan_vascular_2009; @gorelick_vascular_2011; @iadecola_neurovascular_2019]
- The Framingham Risk Score (FRS), a composite measure of 10-year
  cardiovascular event probability [@dagostino_general_2008], has been
  associated with reduced brain volumes, white matter integrity, and
  accelerated cognitive decline [@zsoldos_association_2020; @seshadri_stroke_2004;
  @song_associations_2020; @kaffashian_predictive_2011]
- FRS is heavily weighted by age, with coefficients of 3.06 (men)
  and 2.33 (women) in the log-hazard scale [@dagostino_general_2008]
- In older samples where age variance is restricted, this creates
  substantial confounding between FRS and any age-related outcome
  [@koller_development_2012; @rodondi_framingham_2012]

## The FRS Problem in Aging Samples

- If age drives both FRS (by formula construction) and
  brain/cognitive outcomes (by biological aging), the apparent
  FRS → HVR → Cognition pathway may be spurious
- Age is a common cause of both FRS (deterministically, via formula
  coefficients) and neurodegenerative outcomes (biologically),
  creating a confound structure in aging populations
- Simply controlling for age in regression models is insufficient
  because the confounding is built into FRS construction — age is not
  an external confounder but an internal component of the score
- In elderly cohorts such as ADNI, the ADSP-PHC-derived FRS
  variable is capped at 30% 10-year risk
  [@dagostino_general_2008; @lee_fmnl2_2022], further limiting
  sensitivity by compressing variance in the upper range

## MIMIC Model Approach

- The Multiple Indicators Multiple Causes (MIMIC) model treats age
  as a *cause* of CVR rather than an indicator [@joreskog_estimation_1975;
  @muthen_general_1984]
- This allows extraction of CVR factor scores representing residual
  cardiovascular risk after accounting for age [@zhang_bayesian_2024]
- The resulting CVR~mimic~ measure is orthogonal to age by
  construction (r ≈ 0)
- Structural model: CVR = γ × Age~z~ + ζ (age predicts CVR)
- Factor scores represent ζ: residual CVR not explained by age

## Study Objectives

- Primary: Determine whether FRS effects on
  hippocampal-cognitive coupling reflect true
  cardiovascular pathophysiology or age-related
  confounding
- Hypothesis 1: If FRS effects are driven by
  age-weighting, CVR~mimic~ should show null or weaker
  a-path (CVR → HVR slope) in LGCM mediation
- Hypothesis 2: If FRS mediation effects are
  age-confounded, CVR~mimic~ indirect effects through
  HVR should be null or weaker
- Hypothesis 3: LME 3-way interaction
  (YRS × CVR × HVR) may persist for CVR~mimic~ if CVR
  modulates brain-cognition *coupling* independent of
  accelerating atrophy
- All analyses are sex-stratified, given established
  sex differences in cardiovascular risk profiles,
  brain aging trajectories, and AD pathophysiology
  [@ferretti_sex_2018; @mielke_consideration_2022]

---

# Methods

## Study Design and Participants

- Data source: Alzheimer's Disease Neuroimaging Initiative (ADNI),
  multi-site longitudinal study across phases ADNI-1, ADNI-GO,
  ADNI-2, ADNI-3
- Inclusion criteria:
  - Amyloid-positive status (PET Centiloid > 25
    [@klunk_centiloid_2015] across FBP/FBB/PIB tracers, or CSF
    A+ via Gaussian Mixture Model classification
    [@timsina_harmonization_2024])
  - Available MRI-derived hippocampal and ventricular volumes
  - Available cardiovascular risk factor data (FRS components)
  - At least one cognitive assessment with Alzheimer's
    Disease Sequencing Project Phenotype Harmonization
    Consortium (ADSP-PHC) harmonized composite scores
- Diagnostic composition: full spectrum of amyloid-positive
  individuals including cognitively unimpaired (CU), mild cognitive
  impairment (MCI), and Alzheimer's disease dementia (AD)
- Participant flow shown in Figure \ref{fig-consort}
- We used a separate normative comparison sample of
  amyloid-negative, cognitively unimpaired participants
  with MRI within the UK Biobank (UKB) age range
  (46-82 years) to validate the transfer of
  UKB-derived z-scores to the ADNI cohort (see
  Normative Transfer Validation)

```{r fig-consort}
#| output: asis

cc <- consort_counts
tikz_code <- generate_consort_tikz(
  n_database = cc$n_database,
  n_with_amyloid = cc$n_with_amyloid,
  n_amyloid_pos = cc$n_amyloid_pos,
  n_with_mri = cc$n_with_mri,
  n_lme = cc$n_lme,
  n_lgcm = cc$n_lgcm,
  n_aneg_mri = cc$n_aneg_mri,
  n_aneg_cu_mri = cc$n_aneg_cu_mri,
  n_age_comparable = cc$n_age_comparable,
  ukb_age_range = cc$ukb_age_range
)

cat("\\begin{figure}[H]\n")
cat(tikz_code)
cap <- paste0(
  "\\caption{CONSORT flow diagram showing ",
  "participant selection. Left: amyloid-positive",
  " analysis track from LME cohort to LGCM ",
  "subsample. Right: amyloid-negative, ",
  "cognitively unimpaired normative comparison",
  " sample used to validate UKB z-score ",
  "transfer. ADSP-PHC = Alzheimer's Disease ",
  "Sequencing Project Phenotype Harmonization ",
  "Consortium; CL = Centiloid; CSF = ",
  "cerebrospinal fluid; CU = cognitively ",
  "unimpaired; EDT = estimated disease time; ",
  "MRI = magnetic resonance imaging; ",
  "UKB = UK Biobank.}\n"
)
cat(cap)
cat("\\label{fig-consort}\n")
cat("\\end{figure}\n")
```

## Neuroimaging: Hippocampal-to-Ventricle Ratio (HVR)

- MRI acquisition: T1-weighted MPRAGE sequences on
  1.5T and 3T scanners across ADNI sites with
  standardized acquisition protocols
- Volume segmentation: we obtained hippocampal (HC)
  and temporal horn of the lateral ventricle (VC)
  volumes using a convolutional neural network (CNN)
  trained on manual labels from the same research
  group [@fernandez-lozano_enhanced_2025]. Volumes are provided
  in stereotaxic space (mm³) and converted to native
  space via subject-specific scale factors
- We defined the hippocampal-to-ventricle ratio
  (HVR) as HVR = (HC~L~ + HC~R~) / (HC~L~ + HC~R~
  + VC~L~ + VC~R~), computed on bilateral summed
  native-space volumes [@schoemaker_hippocampal-ventricle_2019]. This ratio
  captures hippocampal integrity relative to surrounding
  cerebrospinal fluid (CSF)-filled spaces, where
  hippocampal atrophy is accompanied by compensatory
  expansion of the temporal horns (ex vacuo dilation)
  - HVR near 1.0: large hippocampus relative to
    ventricles (preserved structure)
  - HVR near 0.0: small hippocampus relative to
    ventricles (atrophy with ventricular expansion)
  - The ratio inherently normalizes for head size without
    requiring intracranial volume (ICV) correction, as
    both numerator and denominator scale proportionally
    with head size
  - The CNN demonstrated low failure rates in ADNI
    compared to multi-atlas and patch-based alternatives
    [@fernandez-lozano_enhanced_2025]
- Normative z-scoring: we derived HVR z-scores from
  Generalized Additive Models for Location, Scale, and
  Shape (GAMLSS; @rigby_generalized_2005) fit on UKB data
  (N = 27,434 cognitively healthy adults, age 45–83)
  - Distribution family: Box-Cox Cole and Green (BCCG),
    which models three distributional parameters —
    location (μ, the conditional median), scale (σ,
    coefficient of variation), and skewness (ν, Box-Cox
    power) — as smooth functions of covariates
  - Covariates: age (penalized cubic spline), education
    (linear), and intracranial cavity volume (ICC);
    separate models fit for each sex
  - The z-score quantifies how many standard deviations
    an individual's observed HVR deviates from the
    conditional median predicted by the GAMLSS model
    given their age, sex, education, and head size. A
    z-score of zero corresponds to a value at the
    conditional median; the transformation accounts for
    age-related heteroscedasticity and distributional
    skewness through the BCCG parametrization
  - Negative z-scores indicate smaller HVR than expected
    for the individual's age, sex, education, and head
    size — i.e., disproportionate hippocampal volume
    loss relative to the normative reference population.
    These are cross-sectional deviation scores, not
    measures of longitudinal change rate

### Normative Transfer Validation

- Valid cross-cohort application of UKB-trained GAMLSS
  models requires verification that z-scores are
  calibrated in the target ADNI population
- We performed validation in a comparable ADNI
  subsample:
  amyloid-negative, cognitively unimpaired participants
  within the UKB age range (46–82 years), where z-scores
  would be expected to approximate N(0, 1)
- Descriptive assessment: we compared mean z-scores,
  standard deviations, and 95% confidence intervals
  against expected N(0, 1), and visualized
  age-bin-specific calibration via forest plot

## Cardiovascular Risk Assessment

### Framingham Risk Score (FRS)

- We obtained FRS values from the Alzheimer's Disease
  Sequencing Project Phenotype Harmonization Consortium
  (ADSP-PHC) Cardiovascular Risk Factor dataset
  [@lee_fmnl2_2022; @lee_reliability_2023]
- The ADSP-PHC computed FRS with the Framingham 2008
  simple office-based 10-year CVD risk score via the
  `ascvd_10y_frs_simple` function from the CVrisk R
  package [@castro_cvrisk_2026]. Despite the function name, this
  implements the Framingham general cardiovascular
  risk algorithm
  [@dagostino_general_2008], not the ACC/AHA Pooled Cohort
  Equations (ASCVD). The score estimates 10-year
  cardiovascular disease risk based on: sex, age
  (≥ 30 years), body mass index (BMI, kg/m²),
  systolic blood pressure (mmHg), use of blood
  pressure medication, current smoking status, and
  diabetes status
- This is the "simple" FRS variant that uses BMI in place
  of the lipid panel (total and HDL cholesterol) required
  by the general FRS [@dagostino_general_2008]
- Age coefficients in the log-hazard scale: 3.06 (men),
  2.33 (women) — age is the dominant component in older
  adults
- The ADSP-PHC-derived variable is capped at 30%
  10-year risk, creating a ceiling effect in elderly
  samples

### MIMIC Model for Age-Adjusted CVR (CVR~mimic~)

We specified the CVR~mimic~ latent factor as a
Multiple Indicators Multiple Causes (MIMIC) model
[@joreskog_estimation_1975; @muthen_general_1984], in which age is an
exogenous cause of the latent cardiovascular risk
factor and the disturbance ζ captures residual CVR
independent of age [@zhang_bayesian_2024].

#### Model Specification

$$
\begin{aligned}
\text{Structural:} \quad &
  \text{CVR} = \gamma \cdot \text{Age}_z + \zeta \\
\text{Measurement:} \quad &
  x_j = \lambda_j \cdot \text{CVR} + \delta_j,
  \quad j = 1, \ldots, 5 \\
\text{Residual covariances:} \quad &
  \text{Cov}(\delta_{\text{SBP}},
    \delta_{\text{CHOL}}) \neq 0; \;
  \text{Cov}(\delta_{\text{SBP}},
    \delta_{\text{CREAT}}) \neq 0
\end{aligned}
$$

- Five measurement indicators:
  - Systolic blood pressure (SBP; z-scored,
    continuous)
  - Hypertension (history of hypertension; binary)
  - Fasting glucose (z-scored, continuous)
  - Total cholesterol (z-scored, continuous)
  - Serum creatinine (z-scored, continuous)
- Two correlated residuals:
  - SBP ↔ Cholesterol: shared vascular stiffness
    and lipid metabolism pathways — arterial
    compliance links blood pressure regulation to
    lipid profiles beyond the general CVR factor
  - SBP ↔ Creatinine: renal-cardiovascular axis —
    kidney function directly influences blood
    pressure regulation through the
    renin-angiotensin system, partially independent
    of overall CVR
  - Both modification indices and substantive
    reasoning supported these specifications

#### Indicator Selection

We determined the final set of five indicators
through iterative model refinement, evaluating
factor loadings, modification indices, and
substantive interpretability at each step.

- Candidate indicators evaluated but excluded:
  - BMI (body mass index): available in the ADSP-PHC
    dataset but its factor loading became
    non-significant (p > 0.20) once specific
    residual covariances with hypertension, glucose,
    and cholesterol were modeled — its
    cardiovascular risk variance is fully captured by
    the retained indicators through vascular
    (hypertension), metabolic (glucose), and lipid
    (cholesterol) pathways
  - Diabetes, heart disease, and stroke (ADSP-PHC
    binary medical history flags): near-perfect
    tetrachoric correlations (φ > 0.90) due to a
    harmonization artifact in how self-reported
    conditions were coded across ADNI phases —
    rendering them empirically indistinguishable and
    unsuitable as separate indicators
  - Smoking status: non-significant factor loading
    (p > 0.20), likely reflecting survivorship bias
    in an elderly cohort where current smokers are
    underrepresented
  - Blood pressure medication: near-complete
    redundancy with hypertension status (the factor
    fully explains their shared variance)

#### Estimation and Fit

- We evaluated model fit with robust (MLR-scaled)
  comparative fit index (CFI), Tucker-Lewis index
  (TLI), root mean square error of approximation
  (RMSEA), and standardized root mean square residual
  (SRMR)

#### Score Extraction and Age Adjustment

- We extracted factor scores in three steps:
  1. We obtained predicted latent factor scores (η)
     from the fitted MIMIC model via empirical Bayes
     modal estimation
  2. We computed age-adjusted CVR scores (ζ) as:
     ζ = η − γ × Age~c~, where γ is the estimated
     regression of CVR on centered age from the
     structural path. This removes the age-predicted
     component of the latent variable, ensuring
     CVR~mimic~ represents residual
     cardiovascular risk after partialling out age at
     the latent level rather than at the indicator
     level
  3. We z-standardized the resulting ζ scores on the
     common analysis sample for comparability with FRS
- This latent-level residualization operates on the
  measurement-error-free latent variable, and is
  equivalent to extracting the disturbance term of the
  MIMIC structural equation [@bollen_structural_1989]
- By construction: correlation(CVR~mimic~, Age) ≈ 0

#### Measurement Invariance

- We tested measurement invariance (MI) across sex
  in four steps — configural, metric, scalar, and
  strict — using the CFA measurement model only
  (without the Age → CVR structural path). MI
  evaluates whether factor loadings, intercepts, and
  residual variances are equivalent across groups,
  which is required before interpreting sex-stratified
  analyses
- We evaluated MI with the Satorra-Bentler scaled
  chi-square difference test [@satorra_scaled_2001], which
  correctly handles the non-subtractable nature of
  MLR-scaled test statistics, and with change in fit
  indices (|ΔCFI| ≤ 0.01, ΔRMSEA ≤ 0.015). When the
  two criteria disagreed, we adopted the stricter
  result

### CVR~mimic~ Validation

- We validated CVR~mimic~ by:
  - Correlation with age (expected ≈ 0, confirming successful
    age-partialling)
  - Correlation with FRS (moderate expected, reflecting shared
    cardiovascular indicators)
  - Distribution properties (approximate standard normal)
  - FRS ceiling analysis (percentage at the 30% risk
    cap) to document variance compression in the
    elderly sample

### CV Risk Stratification

- For trajectory visualizations, we defined CV risk
  groups by median split of baseline FRS (for FRS
  analyses) or baseline CVR~mimic~ (for CVR~mimic~
  analyses)
- We similarly defined HVR groups by median split of
  baseline HVR z-score
- We assigned groups at baseline and held them
  constant across follow-up

### CVR Predictor Standardization

- We z-scored both FRS and CVR~mimic~ on a common
  analysis sample — participants with complete data
  for both FRS components and MIMIC model indicators
  — prior to all downstream analyses
- This ensures that regression coefficients from both
  linear mixed-effects (LME) models and latent growth
  curve models (LGCM) are on an identical scale
  (effect per 1 SD change), enabling direct comparison
  of effect magnitudes between FRS and CVR~mimic~
- We retained raw (pre-standardized) FRS values for
  descriptive statistics and ceiling analyses

## Disease Time Estimation

- Calendar time (years from baseline) assumes all
  participants progress at the same rate, which is
  unrealistic in a heterogeneous AD sample spanning
  cognitively unimpaired (CU), mild cognitive
  impairment (MCI), and dementia stages
- Estimated Disease Time (EDT) aligns participants on
  a common disease timeline by estimating each
  individual's position along an underlying progression
  curve [@raket_statistical_2020]
- We computed EDT using the progmod R package with a
  simultaneous multivariate model fitting ADAS-Cog 13
  and MMSE jointly, sharing a common disease time
  random effect across scales [@kuhnel_simultaneous_2021]
- The simultaneous approach provides more stable
  individual time estimates than univariate models by
  leveraging concordant information across cognitive
  instruments
- EDT values were mean-centered for numerical stability
  while retaining the years metric; centering does not
  affect between-subject differences in disease timing
- For participants with missing EDT at later timepoints
  (due to dropout), values were extrapolated using a
  hybrid approach: a linear slope computed from the
  individual's observed EDT values when at least two
  were available, or the sex-stratified population
  mean EDT slope otherwise, to minimize attrition
  bias from listwise deletion

## Cognitive Assessment

- ADSP-PHC composite scores [@kang_genome-wide_2023]
- Three domains assessed:
  - Memory: episodic memory composite (WMS-R
    Logical Memory, Rey AVLT, ADAS-Cog word recall,
    MMSE/MoCA memory items)
  - Language: language/semantic composite (category
    fluency, Boston Naming Test, ADAS-Cog
    naming/commands)
  - Executive Function: executive composite (Trail
    Making A/B, WAIS-R Digit Span/Symbol, clock
    drawing)
- Scores co-calibrated across multiple cohorts via
  bifactor item banking approach, initially across
  ADNI, Adult Changes in Thought (ACT), and
  Religious Orders Study/Memory and Aging Project
  (ROS/MAP), with subsequent addition of the
  National Alzheimer's Coordinating Center (NACC)
- Higher composite scores indicate better cognitive
  performance across all three domains
- Each score has an associated standard error (SE) from the CFA
  scoring model, reflecting measurement precision at each observation
- In the LGCM, cognitive composites enter as single indicators with
  residual variances fixed to SE² (the squared standard error),
  embedding known measurement error directly from the ADSP-PHC model
  rather than estimating it as a free parameter

## Statistical Analysis

### Analytical Framework

- All primary analyses (LME and LGCM) run twice: once
  with FRS, once with CVR~mimic~ as the CVR predictor
- Key diagnostic: if FRS effects disappear when
  replaced by age-adjusted CVR~mimic~, this implicates
  age-confounding rather than true cardiovascular
  pathophysiology
- Central question: does CVR influence the
  hippocampal-cognitive relationship, and does the
  answer depend on how CVR is measured?
- All models sex-stratified (no pooled sample),
  following the NIH policy on Sex as a Biological
  Variable (SABV; @clayton_policy_2014)
  - Sex differences in AD pathophysiology,
    cardiovascular risk profiles, and brain aging
    trajectories motivate stratification rather than
    inclusion of sex as a covariate
    [@ferretti_sex_2018; @mielke_consideration_2022]
  - Stratification allows effect sizes, variance
    components, and covariate relationships to differ
    freely between sexes; sex is omitted from
    covariates because it defines the strata
- Covariates in all models: age at baseline, education
  (years), APOE ε4 carrier status
- Primary inferences based on uncorrected per-test
  p-values given the pre-specified, hypothesis-driven
  nature of the analyses [@rothman_no_1990; @perneger_whats_1998].
  As a supplementary check, we applied FDR
  correction (Benjamini-Hochberg) within each CVR
  measure across the 6 LME tests (3 domains × 2
  sexes). FRS and CVR~mimic~ corrected separately
  because they address distinct hypotheses. Whether
  effects surviving uncorrected thresholds also survive
  FDR is reported in Results. The key inferential
  comparison (a-path FRS vs a-path CVR~mimic~) is
  inherently directional and does not require
  correction across independent model comparisons
  [@perneger_whats_1998].
  For the LGCM mediation models (12 tests: 2
  sexes × 3 domains × 2 predictors), no FDR
  correction: inference rests on the
  consistency of the pattern across models
  (FRS a-path significant in 6/6 vs CVR~mimic~
  0/6) rather than on any single test
- **Analytical roadmap.** The two modelling
  frameworks address distinct questions:
  - *LME (moderation)*: does CVR change the
    HVR–cognition coupling over time? Tested via
    the 3-way YRS × CVR × HVR interaction
  - *LGCM (mediation)*: does CVR drive cognitive
    decline through hippocampal atrophy? Tested
    via the indirect path CVR → HVR slope →
    cognitive slope
  - Comparing FRS and CVR~mimic~ across both
    frameworks isolates the role of age-weighting
    in observed CVR–brain associations
- **Formal predictor comparison.** FRS and
  CVR~mimic~ fitted on the same participants
  require a paired test to compare coefficients
  - *LGCM*: shared-seed bootstrap ensures
    identical resamples for both predictors;
    paired $\Delta_a = a_\text{FRS} -
    a_\text{CVR}$ with bias-corrected 95% CI
  - *LME*: clustered bootstrap (resampling
    subjects with replacement); paired
    $\Delta\beta = \beta_\text{FRS} -
    \beta_\text{CVR}$ for the 3-way interaction,
    bias-corrected 95% CI

### Methodological Validations

#### EDT Linearity Validation

- Likelihood ratio test (LRT) using LME models
  comparing linear vs quadratic EDT specifications,
  testing whether adding a quadratic EDT term improves
  model fit. This complements the LGCM quadratic
  convergence findings (see below)

#### Simulation Study: SE-Constrained Validation

- A Monte Carlo simulation validated the SE-constrained
  LGCM specification (residual variance fixed to SE²)
  against free estimation (residual variance freely
  estimated). This validates the SE-constrained
  specification used in the LGCM models described below
- This is a methodological validation, not a power
  analysis — it validates the estimation approach under
  known truth, not detection rates for specific effect
  sizes
- The SE-constraint is a key methodological innovation:
  normative z-scores from GAMLSS models have known SEs,
  so fixing residual variance to SE² uses this
  information rather than discarding it
- Data generated from a univariate latent growth model
  with known true parameters:
  - Intercept mean = `r simulation.res$true_params$intercept_mean`,
    slope mean = `r simulation.res$true_params$slope_mean`
    per year EDT
  - Intercept variance =
    `r simulation.res$true_params$intercept_var`,
    slope variance =
    `r simulation.res$true_params$slope_var`
  - HVR → Slope true effect =
    `r simulation.res$true_params$hvr_effect`
  - Measurement SE =
    `r simulation.res$true_params$se_true`
    (SD = `r simulation.res$true_params$se_variation`
    across observations)
  - EDT: mean =
    `r simulation.res$true_params$edt_mean` years,
    SD = `r simulation.res$true_params$edt_sd`,
    change =
    `r simulation.res$true_params$edt_change`
    per visit
- Two approaches compared across
  `r simulation.res$config$n_reps`
  replications at
  N = `r paste(simulation.res$config$sample_sizes, collapse = " and N = ")`:
  (a) residual variance fixed to SE² (SE-constrained)
  (b) residual variance freely estimated
- Performance metrics: parameter bias (mean difference
  from truth), RMSE (root mean squared error), and 95%
  CI coverage probability

### LME Growth Models with 3-Way Interaction

- Model specification (per sex):

$$
\begin{aligned}
\text{Cognition}_{it} = & \; (\beta_0 + b_{0i}) +
  (\beta_1 + b_{1i}) \cdot \text{YRS}_{it} \\
& + \text{YRS}_{it} \times \text{CVR}_i \times
  \text{HVR}_{zit} \;\text{(all 2- and 3-way terms)} \\
& + \text{YRS}_{it}^2 \times \text{CVR}_i \times
  \text{HVR}_{zit} \;\text{(all 2- and 3-way terms)} \\
& + \boldsymbol{X}_i\boldsymbol{\gamma} + \epsilon_{it}
\end{aligned}
$$

  where the linear block ($\text{YRS} \times \text{CVR}
  \times \text{HVR}$) expands to YRS×CVR, YRS×HVR, and
  YRS×CVR×HVR, and the quadratic block
  ($\text{YRS}^2 \times \text{CVR} \times \text{HVR}$)
  expands to YRS², YRS²×CVR, YRS²×HVR, and
  YRS²×CVR×HVR. The quadratic interaction terms test
  whether CVR moderation of the HVR-cognition
  relationship accelerates (or decelerates) over
  follow-up.

- Random effects: subject-specific intercepts
  ($b_{0i}$) and linear slopes ($b_{1i}$)
- CVR entered as either FRS or CVR~mimic~ in parallel
  model sets
- Hypothesis tested: does CVR moderate how the
  HVR-cognition relationship evolves over time?
- Interpretation of 3-way interaction: negative β means
  higher CVR increasingly weakens the protective
  HVR-cognition association over follow-up
- Quadratic time (YRS²) justified by likelihood ratio
  tests comparing nested linear and quadratic models
  — all 12 models (2 sexes × 2 brain measures ×
  3 domains) showed significantly better fit with a
  quadratic term (LRT p < 0.05)

### LGCM Parallel Process Models

- LGCM with parallel HVR and cognitive trajectories
  [@bollen_latent_2006; @muniz-terrera_latent_2017] tested whether CVR
  affects the hippocampal-cognitive relationship
  via mediation (CVR → HVR slope → Cognitive slope)

#### Model Specification

$$
\begin{aligned}
y_{it} &= \eta_{0i} + \eta_{1i} \cdot
  \text{EDT}_{it} + \epsilon_{it},
  \quad \text{Var}(\epsilon_{it}) = \text{SE}_{it}^2
  \\[4pt]
\eta_{0i} &= \alpha_0 +
  \boldsymbol{X}_i \boldsymbol{\gamma}_0 + \zeta_{0i}
  \\
\eta_{1i} &= \alpha_1 +
  \boldsymbol{X}_i \boldsymbol{\gamma}_1 + \zeta_{1i}
\end{aligned}
$$

  where $y_{it}$ is the observed score (HVR z-score
  or cognitive composite) for person $i$ at occasion
  $t$; $\eta_{0i}$ and $\eta_{1i}$ are
  person-specific intercepts and slopes; EDT~it~ is
  the individually-varying time score (see below);
  $\epsilon_{it}$ has variance fixed to the squared
  standard error from the upstream scoring model;
  and $\boldsymbol{X}_i$ includes CVR predictor, age,
  education, and APOE ε4 status.

- Two parallel processes (HVR and cognitive composite)
  are specified simultaneously, with their latent
  growth factors allowed to covary or linked by
  directional paths depending on the model
- Up to 6 measurement occasions per participant
- Cognitive composite residual variances fixed to SE²
  from the ADSP-PHC CFA scoring model, embedding
  known heteroscedastic measurement error directly
  rather than estimating it as a free parameter. This
  yields more precise latent slope estimates by
  separating true variance from measurement noise
- Fully random quadratic models (6 latent factors
  per domain) failed to converge due to near-zero
  quadratic variance and insufficient data capacity
  for the combined model complexity. We therefore
  adopted a fixed-quadratic specification: two
  additional phantom latent factors (HVR quadratic,
  cognitive quadratic) with variance fixed to near
  zero and free population means, capturing
  population-level curvature (2 additional free
  parameters) without individual-level quadratic
  variation. Likelihood ratio tests comparing this
  fixed-quadratic model against a purely linear
  model determined whether quadratic curvature
  improved fit
- We attempted to embed the full MIMIC measurement
  model within the OpenMx parallel process framework,
  but only 1/6 models converged. We therefore used
  pre-computed ζ scores instead (see Score Extraction
  above)

#### Time Metric

- Slope loadings use individually-varying time scores
  (ITVS) determined by each participant's EDT, which
  maps each visit to a position on the disease
  progression timeline rather than calendar time
- Because ITVS give each participant a unique implied
  covariance structure, standard fit indices (CFI,
  TLI, RMSEA) that require a common saturated
  reference model cannot be computed
  [@grimm_growth_2016; @mehta_putting_2000]

#### Covariates and Predictors

- CVR predictors (FRS and CVR~mimic~) mean-centered
  within each sex stratum to account for sex
  differences in cardiovascular risk distributions.
  Covariates (age, education) centered on grand means
  pooled across sexes for consistency across
  sex-stratified analyses
- CVR~mimic~ entered as pre-computed age-adjusted ζ
  scores from the MIMIC model, z-standardized on the
  common sample. Both FRS and CVR~mimic~ are observed
  covariates in the LGCM
- We regressed all covariates on both the latent
  intercept and latent slope factors for each process
  (HVR and cognitive)
- We fit the same models with FRS and CVR~mimic~ for
  direct
  comparison of predictive utility

#### Model Fit and Inference

- We handled missing data via full information
  maximum likelihood (FIML), which uses all available
  observations without listwise deletion
- We assessed model fit via −2 log-likelihood (−2LL),
  Akaike information criterion (AIC), and Bayesian
  information criterion (BIC)
- Bootstrap inference: case resampling with
  n = `r format(lgcm_results$bootstrap$n_boot, big.mark = ",")`
  iterations for all mediation paths
  (a, b, c', indirect, total). Both percentile
  and bias-corrected (BC) 95% confidence intervals
  reported. BC intervals adjust for bias in the
  bootstrap distribution and are recommended for
  products of coefficients, which typically have
  skewed sampling distributions [@efron_introduction_1994].
  A path is significant when its BC 95% CI
  excludes zero. This is the primary inference
  method for all mediation paths
- We extracted individual factor scores via
  regression method (`mxFactorScores`) for
  visualization of slope coupling

#### Coupling Model

- Parallel process coupling: free covariance between
  HVR linear slope and cognitive linear slope
  (hvr~s~ ↔ cog~s~) estimated to test whether
  linear rates of hippocampal and cognitive decline
  co-vary; reported as standardized correlations
  (*r* = cov / √(var~hvr\_s~ × var~cog\_s~))

#### Mediation Model

- Directional paths replace the free slope covariance:
  - a-path: CVR → HVR slope (does CVR predict rate of
    hippocampal decline?)
  - b-path: HVR slope → cognitive slope
    (hippocampal-cognitive coupling)
  - c'-path: CVR → cognitive slope (direct effect
    bypassing hippocampal atrophy)
  - Indirect effect: a × b (mediation through
    hippocampus)
- No free hvr~s~ ↔ cog~s~ covariance is estimated
  in the mediation model because the directional
  b-path already captures the slope relationship;
  the coupling model uses a free covariance without
  directional paths instead
- A significant total effect is not required for valid
  mediation inference [@bollen_latent_2006]

### Sensitivity Analyses

#### LME Sensitivity

- **HC z-score specificity**: we repeated the LME
  3-way interaction with total hippocampal volume
  (HC z-score, normalized against UKB norms) instead
  of HVR, testing whether CVR moderation effects are
  specific to the hippocampus-to-ventricle ratio or
  generalize to hippocampal volume alone
- **Attrition analysis**: we compared baseline
  characteristics between completers (≥ 4 visits)
  and dropouts (< 4 visits) using t-tests
  (continuous) and χ² tests (categorical), plus
  logistic regression predicting dropout status from
  baseline characteristics
- **Raw HVR and HC sensitivity**: we repeated the
  LME 3-way interaction using raw (un-normalized)
  HVR and raw HC (with intracranial volume covariate)
  instead of UKB z-scored values, testing robustness
  to the normative z-scoring procedure
- **EDT moderation**: we used EDT as a moderator in a
  4-way interaction (CVR × HVR × EDT × YRS), testing
  whether CVR effects on brain-cognition coupling
  vary by disease stage
- **EDT Early/Middle sensitivity**: we restricted
  LME models to the first two EDT tertiles, excluding
  late-stage participants to assess robustness

## Software

- R (version 4.x) with the following packages:
  - lavaan [@rosseel_lavaan_2012]: MIMIC model specification,
    estimation (MLR + FIML), measurement invariance
    testing, and factor score extraction
  - lme4 [@bates_fitting_2015]: LME growth models with bobyqa
    optimizer and REML estimation
  - OpenMx [@neale_openmx_2016]: LGCM parallel process and
    mediation models with SLSQP optimizer
    (convergence criterion: gradient < $10^{-5}$)
  - progmod [@raket_statistical_2020]: EDT computation
  - CVrisk [@castro_cvrisk_2026]: FRS computation (via ADSP-PHC)

---

# Results

## Sample Characteristics

- LME cohort:
  `r n_subjects` amyloid-positive participants
  (`r sum(baseline.dt$SEX == "Male")` males,
  `r sum(baseline.dt$SEX == "Female")` females)
  with complete data for both FRS and CVR~mimic~
  after common-sample filtering
- Mean age
  `r sprintf("%.1f", age_mean)` ±
  `r sprintf("%.1f", age_sd)` years
  (range
  `r sprintf("%.0f", age_min)`–`r sprintf("%.0f", age_max)`)
- Diagnostic composition: `r n_cu` cognitively
  unimpaired
  (`r sprintf("%.0f", 100 * n_cu / n_subjects)`%),
  `r n_mci` MCI
  (`r sprintf("%.0f", 100 * n_mci / n_subjects)`%),
  `r n_ad` AD dementia
  (`r sprintf("%.0f", 100 * n_ad / n_subjects)`%)
- Median
  `r median_visits` visits
  (range `r range_visits[1]`–`r range_visits[2]`),
  mean follow-up
  `r sprintf("%.1f", fu_mean)` years
  (max `r sprintf("%.1f", fu_max)`)
- Education
  `r sprintf("%.1f", educ_mean)` ±
  `r sprintf("%.1f", educ_sd)` years;
  `r sprintf("%.0f", pct_apoe4)`% APOE ε4 carriers
- Baseline FRS
  `r sprintf("%.1f", frs_mean_raw)` ±
  `r sprintf("%.1f", frs_sd_raw)` (raw);
  CVR~mimic~ ζ
  `r sprintf("%.2f", cvr_mean)` ±
  `r sprintf("%.2f", cvr_sd)` (age-adjusted
  latent scores; both FRS and CVR~mimic~
  z-standardized on the common sample before
  analysis)
- LGCM cohort: `r n_lgcm_total` participants
  (`r lgcm_n.lst$Male$n` males,
  `r lgcm_n.lst$Female$n` females) with ≥2
  occasions and available EDT values
- Mediation bootstrap samples:
  `r med_n.lst$Male` males,
  `r med_n.lst$Female` females (complete cases
  required for case-resampling)
- Participant flow in Figure \ref{fig-consort}; baseline
  demographics by sex in @tbl-demographics

```{r tbl-demographics}
#| tbl-cap: "Baseline characteristics of the ADNI analysis cohort compared with the UK Biobank normative reference sample. Values are mean (SD) for continuous variables, % for categorical. HC = bilateral hippocampal volume; HVR = hippocampus-to-ventricle ratio; FRS = Framingham Risk Score; APOE ε4+ = carriers of at least one ε4 allele. Z-scores derived from UK Biobank GAMLSS normative models."

require_output(
  create_demographics_table(
    cohort_for_tbl.dt,
    "Baseline Demographic and Clinical Characteristics",
    normative_stats = norm_stats,
    ukb_raw = ukb_raw_stats
  ),
  "tbl-demographics"
)
```


We validated UKB-derived z-scores against an
amyloid-negative, cognitively unimpaired ADNI
subsample within the UKB age range. HVR z-scores
showed a systematic negative offset consistent
with the older ADNI age distribution; HC z-scores
were closer to expected values
(Appendix A, see @fig-normative-validation).

## CVR Characterization

### FRS Properties

- FRS strongly correlated with age (Pearson r =
  `r sprintf("%.3f", frs_age_r)`,
  Spearman ρ =
  `r sprintf("%.3f", frs_age_rho)`)
- FRS–age partial correlation (controlling for
  other CVRF) remained substantial
  (r = `r sprintf("%.3f", frs_partial_r)`),
  confirming age dominates FRS in older adults
- Variance decomposition: age alone R² =
  `r sprintf("%.3f", frs_var_decomp$r2_age)`,
  CVRF alone R² =
  `r sprintf("%.3f", frs_var_decomp$r2_cvrf)`,
  combined R² =
  `r sprintf("%.3f", frs_var_decomp$r2_full)` —
  true CVRF contribute modestly beyond age
- Ceiling:
  `r sprintf("%.1f", frs_ceiling_all)`% at
  the 30% risk cap
  (`r sprintf("%.1f", frs_ceiling_male)`% males,
  `r sprintf("%.1f", frs_ceiling_female)`% females)
- Variance compression:
  `r sprintf("%.1f", frs_above20_all)`% scored
  above 20, limiting statistical sensitivity even
  below the formal ceiling

### CVR~mimic~ Validation

- MIMIC model fit: CFI =
  `r sprintf("%.3f", mimic_fit[["cfi.robust"]])`,
  TLI =
  `r sprintf("%.3f", mimic_fit[["tli.robust"]])`,
  RMSEA =
  `r sprintf("%.3f", mimic_fit[["rmsea.robust"]])`,
  SRMR =
  `r sprintf("%.3f", mimic_fit[["srmr"]])`
- Structural path: γ =
  `r sprintf("%.3f", mimic_gamma)`,
  R² = `r sprintf("%.3f", mimic_r2)` — age
  explained only
  `r sprintf("%.1f", mimic_r2 * 100)`% of CVR
  variance, confirming CVR~mimic~ is largely
  age-free
- Factor loadings in @tbl-mimic-loadings; visual
  validation in @fig-cvr-validation

```{r tbl-mimic-loadings}
#| tbl-cap: "MIMIC model factor loadings for the CVR latent factor. λ = unstandardized factor loading; Std. λ = standardized loading; SE = standard error."

stopifnot(!is.null(cvr_model.res))
require_output(
  create_mimic_loadings_table(
    cvr_model.res,
    "MIMIC Model Factor Loadings"
  ),
  "tbl-mimic-loadings"
)
```

- CVR~mimic~ after age-adjustment and
  z-standardization:
  - Correlation with age: r =
    `r sprintf("%.3f", cvr_age_cor)` (near-zero
    by construction)
  - Correlation with FRS: r =
    `r sprintf("%.2f", cvr_frs_cor)` (moderate
    overlap from shared CV indicators)
  - Distribution: mean =
    `r sprintf("%.2f", cvr_mean)`, SD =
    `r sprintf("%.2f", cvr_sd)`

### Measurement Invariance

- MI of the CVR MIMIC model tested across sex in
  four steps (Appendix B, @tbl-mi); configural
  invariance
  supported: CFI =
  `r sprintf("%.3f", mi_cfg_fit[["CFI"]])`,
  RMSEA =
  `r sprintf("%.3f", mi_cfg_fit[["RMSEA"]])`

```{r mi-inline-steps}
#| output: asis

# Report each MI step with delta-fit indices
# Deltas in comparisons_delta/lrt, keyed as
# "base_vs_test"
mi_comp <- list(
  metric = "configural_vs_metric",
  scalar = "metric_vs_scalar",
  strict = "scalar_vs_strict"
)
mi_labels <- c(
  metric = "Metric (equal loadings)",
  scalar = "Scalar (equal intercepts)",
  strict = "Strict (equal residuals)"
)
for (step in names(mi_comp)) {
  ckey <- mi_comp[[step]]
  delta <- cvr_mi.res$comparisons_delta[[ckey]]
  lrt <- cvr_mi.res$comparisons_lrt[[ckey]]
  if (is.null(delta) && is.null(lrt)) next
  d_cfi <- if (!is.null(delta$d_cfi)) {
    sprintf("%.3f", delta$d_cfi)
  } else "NA"
  d_rmsea <- if (!is.null(delta$d_rmsea)) {
    sprintf("%.3f", delta$d_rmsea)
  } else "NA"
  chi_p <- if (!is.null(lrt$p)) {
    fmt_p_inline(lrt$p)
  } else "NA"
  cat(sprintf(
    "- %s: ΔCFI = %s, ΔRMSEA = %s (%s)\n",
    mi_labels[step], d_cfi, d_rmsea, chi_p
  ))
}
# Report scalar failure if applicable
scalar_delta <- cvr_mi.res$comparisons_delta[[
  "metric_vs_scalar"
]]
scalar_fail <- !is.null(scalar_delta) &&
  !is.null(scalar_delta$d_cfi) &&
  abs(scalar_delta$d_cfi) > 0.01
if (scalar_fail) {
  cat(paste0(
    "\n- Scalar invariance not achieved",
    " (|\u0394CFI| > 0.01): intercepts",
    " differ across sex\n",
    "- Metric invariance (equal",
    " loadings) supported — sufficient",
    " for sex-stratified analyses\n"
  ))
} else {
  cat(sprintf(
    "\n- %s invariance achieved\n",
    mi_verdict
  ))
}
```

```{r fig-cvr-validation}
#| fig-cap: "CVR characterization. (A) FRS distribution by sex; dashed red line marks the 30% risk ceiling. (B) FRS vs Age showing strong positive correlation driven by age-weighting in the FRS formula. (C) CVR~mimic~ vs Age showing near-zero correlation, confirming successful age adjustment; dashed line marks zero."
#| fig-width: 6.5
#| fig-height: 4.5

stopifnot(!is.null(cvr_scores), !is.null(cohort.dt))
  # Baseline with raw FRS percentage
  frs_raw_col <- get_parameter("frs", "column")
  cvrf_raw.dt <- fread(
    get_data_path("raw", "adsp_phc_cvrf"),
    select = c("PTID", frs_raw_col)
  )
  setnames(cvrf_raw.dt, frs_raw_col, "FRS_pct")

  baseline_merged.dt <- cohort.dt[, .(
    AGE = AGE[1], SEX = SEX[1]
  ), by = .(RID, PTID)]
  baseline_merged.dt <- merge(
    baseline_merged.dt,
    cvrf_raw.dt, by = "PTID"
  )
  merged_plot.dt <- merge(
    baseline_merged.dt,
    cvr_scores[, .(RID, CVR_mimic)],
    by = "RID"
  )

  p_frs_sex <- plot_frs_ceiling_by_sex(
    merged_plot.dt
  )
  p_frs_age <- plot_frs_validation(
    merged_plot.dt
  )
  p_cvr_age <- plot_cvr_validation(
    merged_plot.dt
  )

  p_frs_sex / (p_frs_age + p_cvr_age) +
    plot_layout(heights = c(1, 1)) +
    plot_annotation(tag_levels = "A")
```


EDT linearity tests and SE-constrained LGCM
simulation validation are reported in Appendix B
(see @tbl-linearity).

## LME Results: FRS vs CVR~mimic~ 3-Way Interaction

### 3-Way Interaction (FRS vs CVR~mimic~)

```{r fig-lme-forest}
#| fig-cap: "Forest plot comparing FRS vs CVR~mimic~ LME 3-way interaction effects (YRS × CVR × HVR z-score → Cognition) by sex and cognitive domain. Filled points indicate p < 0.05; white-filled points indicate p ≥ 0.05. Error bars show 95% CI."
#| fig-width: 6.5
#| fig-height: 2.5

plot_lme_forest_comparison(frs_hvr, cvr_hvr)
```

- Full covariate effects (Age, Education, APOE ε4,
  CVR main effect, HVR z-score) are reported in
  Appendix B (@tbl-lme-covariates,
  @tbl-cvr-lme-covariates)

```{r lme-inline-results}
#| output: asis

stopifnot(!is.null(frs_hvr), !is.null(cvr_hvr))
# Lookup FDR-corrected p from summary table
get_fdr_p <- function(cvr_meas, sex, domain) {
  row <- lme_summary.dt[
    Analysis == "Primary_Stratified" &
    CVR_Measure == cvr_meas &
    Sex == sex & Domain == domain
  ]
  if (nrow(row) > 0 &&
      !is.null(row$p_fdr)) {
    row$p_fdr[1]
  } else NA_real_
}
get_lme_result <- function(
    res, sex, domain, cvr_meas) {
  r <- res$stratified[[sex]][[domain]]$interaction
  if (!is.null(r)) {
    list(
      beta = r$beta, p = r$p_value,
      p_fdr = get_fdr_p(cvr_meas, sex, domain)
    )
  } else NULL
}

# Collect all results for narrative
frs_all <- list()
cvr_all <- list()
for (sx in c("Male", "Female")) {
  for (d in c("MEM", "LAN", "EXF")) {
    frs_all[[paste(sx, d)]] <-
      get_lme_result(frs_hvr, sx, d, "FRS")
    cvr_all[[paste(sx, d)]] <-
      get_lme_result(
        cvr_hvr, sx, d, "CVR_mimic"
      )
  }
}

frs_n_sig <- sum(sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
))
cvr_n_sig <- sum(sapply(
  cvr_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
))

# Compute mean absolute beta ratio
frs_betas <- abs(sapply(
  frs_all, function(r) {
    if (!is.null(r)) r$beta else NA
  }
))
cvr_betas <- abs(sapply(
  cvr_all, function(r) {
    if (!is.null(r)) r$beta else NA
  }
))
beta_ratio <- mean(
  frs_betas, na.rm = TRUE
) / mean(cvr_betas, na.rm = TRUE)

# --- Pattern-first summary ---
# Identify which FRS models are significant
frs_sig_keys <- names(frs_all)[sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
)]
frs_sig_male <- grep(
  "^Male", frs_sig_keys, value = TRUE
)
frs_sig_female <- grep(
  "^Female", frs_sig_keys, value = TRUE
)
fmt_doms <- function(keys) {
  ds <- sub("^(Male|Female) ", "", keys)
  paste(DOMAIN_LABELS[ds], collapse = ", ")
}

cat("**Key pattern.** FRS produced significant")
cat(" 3-way interactions (YRS × CVR × HVR)")
cat(" in males")
if (length(frs_sig_male) > 0) {
  cat(sprintf(
    " (%s)", fmt_doms(frs_sig_male)
  ))
}
if (length(frs_sig_female) > 0) {
  cat(sprintf(
    " and females (%s)",
    fmt_doms(frs_sig_female)
  ))
}
cat(sprintf(
  paste0(
    " — %d/6 total (%d/6 surviving FDR)."
  ),
  frs_n_sig, fdr_summary.lst$frs_n_fdr
))
cat(sprintf(
  paste0(
    " CVR~mimic~ showed %d/6 significant",
    " interactions (%d/6 surviving FDR),",
    " consistent with FRS effects being",
    " primarily driven by age-weighting",
    " rather than true cardiovascular risk",
    " (see @fig-lme-forest).\n\n"
  ),
  cvr_n_sig, fdr_summary.lst$cvr_n_fdr
))

# Helper: format p with FDR annotation
fmt_p_fdr <- function(p, p_fdr) {
  base <- fmt_p_inline(p)
  if (!is.na(p_fdr) && p < 0.05) {
    fdr_tag <- if (p_fdr < 0.05) {
      sprintf(
        " [*p*~FDR~ = %s]",
        fmt_p_inline(p_fdr)
      )
    } else {
      sprintf(
        " [*p*~FDR~ = %s, n.s.]",
        fmt_p_inline(p_fdr)
      )
    }
    paste0(base, fdr_tag)
  } else base
}

# Per-sex detail (sub-bullets)
for (sx in c("Male", "Female")) {
  cat(sprintf("\n**%ss.**\n\n", sx))
  cat("- FRS × HVR × YRS: ")
  items <- c()
  for (d in c("MEM", "LAN", "EXF")) {
    r <- get_lme_result(
      frs_hvr, sx, d, "FRS"
    )
    if (!is.null(r)) {
      items <- c(items, sprintf(
        "%s β = %.4f, %s",
        DOMAIN_LABELS[d], r$beta,
        fmt_p_fdr(r$p, r$p_fdr)
      ))
    }
  }
  cat(paste(items, collapse = "; "))
  cat("\n")

  cat("- CVR~mimic~ × HVR × YRS: ")
  items <- c()
  for (d in c("MEM", "LAN", "EXF")) {
    r <- get_lme_result(
      cvr_hvr, sx, d, "CVR_mimic"
    )
    if (!is.null(r)) {
      items <- c(items, sprintf(
        "%s β = %.4f, %s",
        DOMAIN_LABELS[d], r$beta,
        fmt_p_fdr(r$p, r$p_fdr)
      ))
    }
  }
  cat(paste(items, collapse = "; "))
  cat("\n\n")
}

cat(sprintf(
  paste0(
    "- FRS β values were on average",
    " %.1f-fold larger in absolute",
    " magnitude than CVR~mimic~ β values,",
    " consistent with age-weighting",
    " amplifying apparent FRS effects\n\n"
  ),
  beta_ratio
))

# Worked example: translate largest significant FRS
# 3-way beta into meaningful units
frs_sig <- frs_all[sapply(
  frs_all, function(r) {
    !is.null(r) && r$p < 0.05
  }
)]
if (length(frs_sig) > 0) {
  max_r <- frs_sig[[which.max(
    abs(sapply(frs_sig, `[[`, "beta"))
  )]]
  max_nm <- names(frs_sig)[which.max(
    abs(sapply(frs_sig, `[[`, "beta"))
  )]
  parts <- strsplit(max_nm, " ")[[1]]
  display_nm <- paste(
    parts[1],
    DOMAIN_LABELS[parts[2]]
  )
  cat(sprintf(
    paste0(
      "- Worked example: the largest",
      " significant FRS 3-way effect",
      " (%s, \u03b2 = %.4f) implies that",
      " over 5 years, a 1-SD increase",
      " in FRS is associated with a",
      " %.4f-unit cumulative divergence",
      " in the HVR-cognition",
      " association\n\n"
    ),
    display_nm, max_r$beta,
    max_r$beta * 5
  ))
}
```

```{r lme-fdr-note}
#| output: asis

stopifnot(!is.null(fdr_summary.lst))
  frs_changed <- fdr_summary.lst$frs_n_sig !=
    fdr_summary.lst$frs_n_fdr
  cvr_changed <- fdr_summary.lst$cvr_n_sig !=
    fdr_summary.lst$cvr_n_fdr
  if (!frs_changed && !cvr_changed) {
    cat(
      "- FDR correction (Benjamini-Hochberg)",
      " did not change any conclusions for",
      " either CVR measure\n"
    )
  } else if (!frs_changed && cvr_changed) {
    cat(sprintf(
      paste0(
        "- FDR correction did not change FRS",
        " conclusions. For CVR~mimic~, %d of",
        " %d uncorrected effects survived",
        " FDR\n"
      ),
      fdr_summary.lst$cvr_n_fdr,
      fdr_summary.lst$cvr_n_sig
    ))
  } else {
    cat(sprintf(
      paste0(
        "- After FDR correction, %d/%d FRS",
        " and %d/%d CVR~mimic~ uncorrected",
        " effects remained significant\n"
      ),
      fdr_summary.lst$frs_n_fdr,
      fdr_summary.lst$frs_n_sig,
      fdr_summary.lst$cvr_n_fdr,
      fdr_summary.lst$cvr_n_sig
    ))
  }
```

### Formal Coefficient Comparison (LME)

- Clustered bootstrap Δβ with bias-corrected
  95% CI comparing FRS vs CVR~mimic~ 3-way
  interaction coefficients

```{r lme-comparison-boot}
#| output: asis

lme_cmp <- lme_hvr_z.res$comparison
if (!is.null(lme_cmp) &&
    length(lme_cmp) > 0) {
  for (sx in c("Male", "Female")) {
    cat(sprintf("\n**%ss.**\n\n", sx))
    for (d in c("MEM", "LAN", "EXF")) {
      k <- paste(sx, d, sep = "_")
      r <- lme_cmp[[k]]
      if (is.null(r) ||
          is.na(r$ci_lower)) next
      sig <- if (isTRUE(r$significant)) {
        "\\*"
      } else ""
      cat(sprintf(
        paste0(
          "- %s: \u0394\u03b2 = %.4f",
          " [%.4f, %.4f]%s\n"
        ),
        DOMAIN_LABELS[d],
        r$delta_beta,
        r$ci_lower, r$ci_upper, sig
      ))
    }
  }
} else {
  cat(paste0(
    "- Bootstrap comparison pending;",
    " re-run script 08 to generate\n"
  ))
}
```

## LGCM Results

### Growth Parameters

- LGCM models used EDT as the time metric with
  individually-varying time scores, fit separately
  per sex and cognitive domain

```{r lgcm-growth-inline}
#| output: asis

stopifnot(!is.null(lgcm_parallel$linear_results))
domains <- c("MEM", "LAN", "EXF")
dom_labels <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

# Collect slope means per sex
for (sx in c("Male", "Female")) {
  hvr_slopes <- c()
  cog_slopes <- c()
  for (d in domains) {
    k <- paste0(sx, "_", d)
    lr <- lgcm_parallel$linear_results[[k]]
    if (is.null(lr)) next
    s <- lr$summary
    if (is.null(s)) next
    hvr_s <- s[
      s$label == "hvr_slope_mean" |
      grepl("hvr.*slope.*mean", s$label,
        ignore.case = TRUE), ]
    cog_s <- s[
      s$label == "cog_slope_mean" |
      grepl("cog.*slope.*mean", s$label,
        ignore.case = TRUE), ]
    if (nrow(hvr_s) > 0) {
      hvr_slopes <- c(
        hvr_slopes, hvr_s$Estimate[1]
      )
    }
    if (nrow(cog_s) > 0) {
      cog_slopes <- c(
        cog_slopes, cog_s$Estimate[1]
      )
    }
  }
  sx_label <- paste0(sx, "s")
  if (length(hvr_slopes) > 0 &&
      length(cog_slopes) > 0) {
    cat(sprintf(
      paste0(
        "- %s: mean HVR slopes negative",
        " across all domains (range:",
        " %.4f to %.4f); cognitive slopes",
        " also negative (range: %.4f",
        " to %.4f), confirming progressive",
        " decline in both processes\n"
      ),
      sx_label,
      min(hvr_slopes), max(hvr_slopes),
      min(cog_slopes), max(cog_slopes)
    ))
  }
}

cat(paste0(
  "- Model fit in @tbl-mediation-fit;",
  " standard fit indices (CFI, TLI,",
  " RMSEA) not computable with",
  " individually-varying time scores;",
  " adequacy assessed via \u22122LL,",
  " AIC, BIC, and nested LRTs\n"
))
```

### Formal a-Path Comparison (LGCM)

- Paired bootstrap Δ~a~ (shared-seed resampling)
  comparing FRS vs CVR~mimic~ a-path coefficients

```{r lgcm-comparison-boot}
#| output: asis

lgcm_cmp <- lgcm_results$comparisons
if (!is.null(lgcm_cmp) &&
    length(lgcm_cmp) > 0) {
  for (sx in c("Male", "Female")) {
    cat(sprintf("\n**%ss.**\n\n", sx))
    for (d in c("MEM", "LAN", "EXF")) {
      k <- paste(sx, d, sep = "_")
      r <- lgcm_cmp[[k]]
      if (is.null(r) ||
          is.na(r$delta_a_ci_lower)) next
      sig <- if (
        isTRUE(r$delta_a_significant)
      ) "\\*" else ""
      cat(sprintf(
        paste0(
          "- %s: \u0394a = %.5f",
          " [%.5f, %.5f]%s\n"
        ),
        DOMAIN_LABELS[d],
        r$delta_a_est,
        r$delta_a_ci_lower,
        r$delta_a_ci_upper, sig
      ))
    }
  }
}
```


- Paired bootstrap confirmed the a-path
  dissociation across all 6 sex × domain models:
  Δ~a~ (FRS − CVR~mimic~) significantly > 0 in
  every case, with 95% BC CIs excluding zero
- Male Δ~a~ estimates consistent across domains
  (mean = `r sprintf("%.4f", m_da_mean)`),
  indicating a stable FRS advantage regardless
  of cognitive outcome
- Female Δ~a~ estimates approximately
  `r da_ratio_pct`% larger
  (mean = `r sprintf("%.4f", f_da_mean)`),
  suggesting the FRS–CVR~mimic~ dissociation is
  more pronounced in females
- These formal tests upgrade the qualitative
  dissociation (FRS significant in 6/6,
  CVR~mimic~ in 0/6) to a quantitative
  difference: FRS is a significantly stronger
  predictor of hippocampal atrophy rate than
  CVR~mimic~ within the mediation framework

### Slope Coupling

- LGCM parallel process models tested whether HVR
  and cognitive linear decline rates co-vary
  (latent slope coupling, hvr~s~ ↔ cog~s~) along
  the EDT disease timeline
- Positive linear slope correlations observed across
  all models in males (3/3 significant) and in
  females for Executive Function (1/3 significant);
  individuals with faster HVR decline showed
  steeper cognitive decline
- This coupling supports the b-path assumption
  in the mediation framework (CVR → HVR slope →
  cognitive slope): HVR decline predicts cognitive
  decline, though the relationship is stronger and
  more consistently significant in males

```{r coupling-inline}
#| output: asis

stopifnot(!is.null(lgcm_parallel$linear_results))
domains <- c("MEM", "LAN", "EXF")
domain_labels <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

# Collect correlations and p-values per sex
for (sx in c("Male", "Female")) {
  cors <- c()
  cor_names <- c()
  ps <- c()
  for (d in domains) {
    k <- paste0(sx, "_", d)
    lr <- lgcm_parallel$linear_results[[k]]
    if (!is.null(lr$coupling) &&
        !is.na(lr$coupling$p)) {
      prm <- lr$params
      v_h <- prm[
        name == "var_hvr_s", Estimate
      ]
      v_c <- prm[
        name == "var_cog_s", Estimate
      ]
      r_val <- lr$coupling$cov /
        sqrt(v_h * v_c)
      cors <- c(cors, r_val)
      cor_names <- c(
        cor_names, domain_labels[d]
      )
      ps <- c(ps, lr$coupling$p)
    }
  }
  if (length(cors) > 0) {
    max_d <- cor_names[which.max(cors)]
    min_d <- cor_names[which.min(cors)]
    max_p <- max(ps)
    p_bound <- if (max_p < 0.001) {
      "all *p* < 0.001"
    } else {
      sprintf(
        "max %s",
        fmt_p_inline(max_p)
      )
    }
    cat(sprintf(
      paste0(
        "- In %ss, linear slope",
        " correlations (*r*)",
        " ranged from %.3f",
        " (%s) to %.3f (%s; %s)\n"
      ),
      tolower(sx),
      min(cors), min_d,
      max(cors), max_d,
      p_bound
    ))
  }
}

cat(paste0(
  "- Faster hippocampal decline",
  " consistently associated with steeper",
  " cognitive decline across all domains",
  " and both sexes; latent factor scores",
  " from the parallel process model confirm",
  " this pattern (see @fig-coupling)\n"
))
```

```{r fig-coupling}
#| fig-cap: "Latent slope coupling between HVR and cognitive decline from LGCM parallel process models. Females in red, Males in blue throughout. A. Individual trajectories (random 20% subsample) with population mean curves (bold; including fixed quadratic curvature where selected by LRT). 2x2 facets: HVR, Memory, Language, Executive. B. Latent slope coupling: latent cognitive decline rate (x) vs latent HVR decline rate (y) from factor scores (positive = faster decline); solid lines show linear trends per sex; grey line shows 1:1 reference. Correlation estimates from the SEM (* p < .05, ** p < .01, *** p < .001)."
#| fig-width: 6.5
#| fig-height: 7.5

lgcm_m_input <- read_rds_safe(here(
  "data/derivatives/lgcm/lgcm_input_male.rds"
))
lgcm_f_input <- read_rds_safe(here(
  "data/derivatives/lgcm/lgcm_input_female.rds"
))
stopifnot(
  !is.null(lgcm_m_input),
  !is.null(lgcm_f_input),
  !is.null(lgcm_parallel)
)
plot_coupling_figure(
  lgcm_parallel,
  lgcm_m_input, lgcm_f_input
)
```

### Mediation: FRS vs CVR~mimic~

- LGCM mediation models tested whether CVR effects
  on cognitive decline operate through hippocampal
  atrophy, comparing FRS and CVR~mimic~ as
  predictors (see @fig-mediation)

```{r tbl-mediation-fit}
#| tbl-cap: "LGCM mediation model fit indices, sex-stratified. -2LL = -2 log-likelihood; AIC = Akaike information criterion; BIC = Bayesian information criterion."

require_output(
  create_mediation_fit_table(
    lgcm_results,
    title = "Mediation Model Fit"
  ),
  "tbl-mediation-fit"
)
```

- Mediation path diagrams showing a-, b-, c'-path
  estimates and bootstrap CIs:
  (see @fig-mediation)


```{r fig-mediation}
#| fig-cap: "LGCM mediation path diagrams. Rows: Memory, Language, Executive Function. Columns: FRS, CVR~MIMIC~. Each panel shows Male (blue) and Female (red) path estimates. Solid/bold = significant (bootstrap 95% CI excludes zero); faded = non-significant. Star (\\*) indicates CI excludes zero. 95% bootstrap CIs shown for all paths."
#| fig-width: 6.5
#| fig-height: 8.0

panels <- build_combined_med_panels(
  lgcm_med_male, lgcm_med_female
)
wrap_plots(panels, ncol = 2) +
  plot_annotation(tag_levels = "A")
```

```{r lgcm-inline-results}
#| output: asis

stopifnot(!is.null(lgcm_med_male))
stopifnot(!is.null(lgcm_med_female))

domains <- c("MEM", "LAN", "EXF")

# --- Key dissociation narrative ---
frs_a_sig <- sum(sapply(
  c("Male", "Female"), function(sx) {
    med <- if (sx == "Male") {
      lgcm_med_male
    } else lgcm_med_female
    sapply(domains, function(d) {
      r <- med$results[[
        paste0(d, "_FRS")
      ]]$a
      !is.null(r) &&
        ci_sig_check.fn(r)
    })
  }
))
cvr_a_sig <- sum(sapply(
  c("Male", "Female"), function(sx) {
    med <- if (sx == "Male") {
      lgcm_med_male
    } else lgcm_med_female
    sapply(domains, function(d) {
      r <- med$results[[
        paste0(d, "_CVR")
      ]]$a
      !is.null(r) &&
        ci_sig_check.fn(r)
    })
  }
))
cat(sprintf(
  paste0(
    "- **FRS vs CVR~mimic~:** FRS produced",
    " significant indirect effects",
    " (CVR \u2192 HVR slope \u2192 Cognitive",
    " slope) with a significant a-path",
    " in %d/6 models (bootstrap 95%%",
    " CIs excluding zero);",
    " CVR~mimic~ a-path null in %d/6",
    " models, indicating FRS \"effects\"",
    " on hippocampal decline are largely",
    " consistent with age-weighting\n\n"
  ),
  frs_a_sig, 6L - cvr_a_sig
))

# --- Pattern summary ---
# Count indirect effects across all models
all_ind_sig <- 0L
all_ind_total <- 0L
ind_sig_labels <- c()
for (sx in c("Male", "Female")) {
  med <- if (sx == "Male") {
    lgcm_med_male
  } else lgcm_med_female
  for (d in domains) {
    for (pred in c("FRS", "CVR")) {
      k <- paste0(d, "_", pred)
      r <- med$results[[k]]
      all_ind_total <- all_ind_total + 1L
      if (!is.null(r$indirect) &&
          ci_sig_check.fn(
            r$indirect
          )) {
        all_ind_sig <- all_ind_sig + 1L
        ind_sig_labels <- c(
          ind_sig_labels,
          sprintf(
            "%s %s (%s)",
            sx, DOMAIN_LABELS[d], pred
          )
        )
      }
    }
  }
}
cat(paste0(
  "**Path-level summary across all",
  " 12 models (2 sexes × 3 domains",
  " × 2 predictors):**\n\n"
))
cat(sprintf(
  paste0(
    "- *a-path* (CVR \u2192 HVR slope):",
    " significant for FRS in %d/6",
    " models; null for CVR~mimic~ in",
    " all 6\n"
  ),
  frs_a_sig
))
cat(sprintf(
  paste0(
    "- *b-path* (HVR slope \u2192 cognitive",
    " slope): significant for %d/6 FRS",
    " and %d/6 CVR~mimic~ models",
    " regardless of predictor\n"
  ),
  abs_stats$frs_b_sig,
  abs_stats$cvr_b_sig
))
cat(sprintf(
  paste0(
    "- *Indirect effect*: %d/%d",
    " significant"
  ),
  all_ind_sig, all_ind_total
))
if (length(ind_sig_labels) > 0) {
  cat(sprintf(
    " (%s)",
    paste(ind_sig_labels, collapse = "; ")
  ))
}
cat("\n\n")

# --- Per-sex detail ---
for (sx in c("Male", "Female")) {
  med <- if (sx == "Male") {
    lgcm_med_male
  } else lgcm_med_female

  cat(sprintf("#### %ss\n\n", sx))

  # Collect ranges for a-path
  frs_a_betas <- c()
  cvr_a_betas <- c()
  frs_b_betas <- c()
  frs_b_nsig <- 0L
  frs_a_nsig <- 0L
  cvr_a_nsig <- 0L
  frs_ind_sig <- 0L
  cvr_ind_sig <- 0L
  for (d in domains) {
    rf <- med$results[[paste0(d, "_FRS")]]
    rc <- med$results[[paste0(d, "_CVR")]]
    if (!is.null(rf$a)) {
      frs_a_betas <- c(
        frs_a_betas, rf$a$est
      )
      if (ci_sig_check.fn(rf$a)) {
        frs_a_nsig <- frs_a_nsig + 1L
      }
    }
    if (!is.null(rc$a)) {
      cvr_a_betas <- c(
        cvr_a_betas, abs(rc$a$est)
      )
      if (ci_sig_check.fn(rc$a)) {
        cvr_a_nsig <- cvr_a_nsig + 1L
      }
    }
    if (!is.null(rf$b)) {
      frs_b_betas <- c(
        frs_b_betas, rf$b$est
      )
      if (ci_sig_check.fn(rf$b)) {
        frs_b_nsig <- frs_b_nsig + 1L
      }
    }
    if (!is.null(rf$indirect) &&
        ci_sig_check.fn(
          rf$indirect
        )) {
      frs_ind_sig <- frs_ind_sig + 1L
    }
    if (!is.null(rc$indirect) &&
        ci_sig_check.fn(
          rc$indirect
        )) {
      cvr_ind_sig <- cvr_ind_sig + 1L
    }
  }

  # a-path summary
  cat(sprintf(
    paste0(
      "- FRS a-path: \u03b2 range %.4f",
      " to %.4f (%d/3 bootstrap 95%%",
      " CIs excluding zero);",
      " CVR~mimic~ a-paths: %d/3",
      " significant (all |\u03b2| <",
      " %.2f)\n"
    ),
    min(frs_a_betas), max(frs_a_betas),
    frs_a_nsig, cvr_a_nsig,
    max(cvr_a_betas) + 0.005
  ))

  # b-path summary
  cat(sprintf(
    paste0(
      "- b-path: \u03b2 range %.3f to",
      " %.3f (%d/3 bootstrap 95%%",
      " CIs excluding zero)\n"
    ),
    min(frs_b_betas),
    max(frs_b_betas), frs_b_nsig
  ))

  # Indirect summary
  cat(sprintf(
    paste0(
      "- FRS indirect effects significant",
      " in %d/3 domains (bootstrap",
      " 95%% CI excluding zero);",
      " CVR~mimic~ in %d/3\n"
    ),
    frs_ind_sig, cvr_ind_sig
  ))

  # c' and total brief note
  cat(paste0(
    "- Direct (c') and total effects",
    " in @fig-mediation\n\n"
  ))
}

cat(paste0(
  "- Complete path estimates and",
  " bootstrap CIs in",
  " @fig-mediation\n"
))
```


Sensitivity analyses — including HC z-score
comparisons, attrition analysis, raw (un-normalized)
concordance checks, EDT moderation, and SE-constrained
LGCM validation — are reported in Appendix C.

# Discussion

## Summary of Findings

- Whether CVR affects hippocampal-cognitive
  trajectories depends on how CVR is measured
- a-path dissociation: FRS → HVR slope significant
  in `r abs_stats$frs_a_sig`/6 models (all bootstrap
  95% CIs excluding zero; [@fig-mediation]);
  CVR~mimic~ a-path null in all 6 (0/6 significant)
- b-path (HVR slope → Cognitive slope) consistently
  significant regardless of predictor
  (`r abs_stats$frs_b_sig`/6 FRS,
  `r abs_stats$cvr_b_sig`/6 CVR~mimic~; bootstrap
  CIs excluding zero), confirming consistent
  hippocampal-cognitive coupling
  [@kramer_longitudinal_2007; @oschwald_brain_2020; @anblagan_coupled_2018]
- FRS indirect effects (a × b) significant in
  `r abs_stats$frs_ind_sig`/6 models (bootstrap
  CIs excluding zero); CVR~mimic~ in
  `r abs_stats$cvr_ind_sig`/6
- LME 3-way interaction: FRS × HVR × YRS
  significant in
  `r abs_lme$frs_m_sig`/3 male and
  `r abs_lme$frs_f_sig`/3 female domains;
  CVR~mimic~ in `r abs_lme$cvr_m_sig`/3 male and
  `r abs_lme$cvr_f_sig`/3 female domains
  `r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s)", abs_lme$cvr_f_sig_doms) else ""`
- FDR correction did not change FRS conclusions;
  CVR~mimic~ dropped from
  `r fdr_summary.lst$cvr_n_sig` to
  `r fdr_summary.lst$cvr_n_fdr` significant effects
- Interpretation: FRS effects on hippocampal
  trajectories (LGCM a-path) are primarily
  age-driven, but CVR may still modulate the
  HVR-cognition relationship (LME 3-way) through
  mechanisms other than accelerating atrophy

## FRS Age-Confounding and CVR~mimic~

- CVR~mimic~ successfully removes the age component
  from cardiovascular risk (near-zero age correlation;
  @fig-cvr-validation C), supporting its use as an
  age-independent CVR index for disambiguating
  age-driven vs CVR-driven brain effects
- The moderate CVR~mimic~–FRS correlation reflects
  shared cardiovascular indicators while confirming
  the measures capture distinct constructs — one
  age-confounded, one age-orthogonal
- FRS formula explicitly includes age with large
  coefficients (3.06 men, 2.33 women in log-hazard
  scale) [@dagostino_general_2008]. In older samples, age
  dominates the FRS score
  [@koller_development_2012; @pencina_quantifying_2019]
- FRS values in ADNI are compressed into the upper
  range (see @fig-cvr-validation, panels A–B), limiting score
  variance and statistical sensitivity, particularly
  in females
- When age is partialled out (CVR~mimic~), the
  indirect pathway (CVR → HVR → Cognition)
  disappears — the a-path is null, suggesting FRS
  "effects" on hippocampal-cognitive trajectories
  are actually age effects captured by the age
  component of FRS
- The negative total cholesterol loading
  (see @tbl-mimic-loadings) likely reflects widespread
  statin use in this elderly clinical sample: treated
  individuals have lower cholesterol despite elevated
  underlying cardiovascular risk, inverting the
  expected direction of association with the latent
  CVR factor
- Common-sample standardization (z-scoring both FRS
  and CVR~mimic~ on the same N with complete data for
  both measures) ensures regression coefficients
  express effects per 1 SD change on an identical
  scale — without this step, the a-path dissociation
  could reflect scaling artifacts rather than
  substantive predictor properties
- Implication: studies reporting FRS-brain
  relationships in elderly samples
  [@zsoldos_association_2020; @seshadri_stroke_2004; @song_associations_2020; @meusel_cardiovascular_2019] may be
  detecting age effects rather than true
  cardiovascular effects

## Dissociating Moderation from Mediation

- As outlined in the analytical roadmap, the two
  modelling frameworks test complementary questions.
  The LME 3-way interaction
  (YRS × CVR × HVR → Cognition)
  and the LGCM a-path (CVR → HVR slope;
  [@fig-mediation]) test distinct hypotheses:
  - LME (moderation): does CVR modify how the
    HVR-cognition relationship evolves over time?
  - LGCM (mediation): does CVR affect cognition
    through hippocampal decline?
- The results diverge for CVR~mimic~:
  - LGCM a-path is null in all 6 models: age-adjusted
    CVR does not accelerate hippocampal atrophy
  - LME 3-way is significant in
    `r abs_lme$cvr_f_sig`/3 female domains
    `r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s)", abs_lme$cvr_f_sig_doms) else ""`
    only; `r abs_lme$cvr_m_sig`/3 in males
- This dissociation represents moderation without
  mediation through hippocampal volume: CVR~mimic~
  does not accelerate hippocampal atrophy (a-path
  null), yet it may modify how hippocampal integrity
  relates to cognitive performance over time
- This pattern is consistent with a cognitive reserve
  framework [@stern_cognitive_2012; @rabin_interactive_2018; @iadecola_framingham_2020]:
  cardiovascular risk may reduce the brain's capacity
  to maintain cognitive function at a given level of
  structural integrity, without directly driving
  structural decline
- However, the CVR~mimic~ LME evidence rests on a
  single domain in one sex. This isolated finding
  should be interpreted cautiously until replicated
- The HVR-cognition coupling (b-path) was consistent
  across CVR predictors, in line with prior LGCM
  coupling studies [@muniz-terrera_latent_2017; @bollen_latent_2006]

## Sex Differences

- Males showed `r abs_lme$cvr_m_sig`/3 significant
  CVR~mimic~ LME effects, while females showed
  `r abs_lme$cvr_f_sig`/3
  `r if(nchar(abs_lme$cvr_f_sig_doms) > 0) sprintf("(%s)", abs_lme$cvr_f_sig_doms) else ""`
- The female CVR~mimic~ Executive finding is notable
  as the only CVR~mimic~ LME effect observed in
  either sex and warrants targeted investigation in
  independent samples
- Female FRS variance compression limits
  statistical sensitivity — the FRS findings in
  females may be primarily driven by age-weighting
- Sex differences in cardiovascular risk profiles
  [@ferretti_sex_2018; @mielke_consideration_2022], differential APOE ε4
  effects, and hormonal influences on vascular
  regulation may all contribute to the observed
  sex asymmetry in CVR-brain relationships
- Covariates (age, education, APOE ε4) exhibited
  expected effects across all models
  (Appendix B, @tbl-lme-covariates and
  @tbl-cvr-lme-covariates)

## HVR-Cognitive Coupling

- The HVR slope → Cognitive slope relationship
  (b-path) was consistent regardless of whether FRS
  or CVR~mimic~ was the predictor (@fig-mediation)
- Significant b-path estimates appeared in Memory
  and Executive Function for both predictors (4/6
  each; Language non-significant), supporting HVR
  decline as a reliable marker of cognitive decline
  [@kramer_longitudinal_2007; @oschwald_brain_2020],
  extending earlier evidence that HVR captures
  hippocampal integrity more sensitively than raw
  volume [@schoemaker_hippocampal-ventricle_2019;
  @fernandez-lozano_enhanced_2025;
  @bruno_hippocampal_2016]
- Parallel process coupling showed a complementary
  pattern: significant in all male models (3/3) but
  only Executive Function in females (1/3). This
  sex asymmetry likely reflects lower statistical
  power in the smaller female subsample rather than
  a biological absence of coupling, given that
  mediation b-paths (which condition on a predictor)
  were significant in 4/6 models for both sexes
- Language b-paths were non-significant across both
  analysis frameworks (coupling and mediation),
  consistent with language being a later-declining
  domain with less variance to capture at this
  disease stage
- The hippocampal-cognitive coupling is not an
  artifact of CVR measurement — it persists across
  predictor specifications and analysis frameworks

## Methodological Considerations

- The fixed-quadratic LGCM captures population-level
  curvature but constrains individual quadratic
  variation, so slope estimates primarily reflect
  average rates of change. This makes the a-path
  null finding conservative — if CVR~mimic~ had
  individual-level non-linear effects on hippocampal
  trajectories, these would require a fully random
  quadratic model to detect, further strengthening
  the conclusion that the FRS a-path reflects
  age-weighting
- EDT linearity tests confirmed significant
  non-linear trajectories across all models
  (Appendix B, @tbl-linearity), motivating the
  fixed-quadratic specification [@newsom_longitudinal_2015].
  Fully random quadratic parallel process models
  failed to converge due to near-zero quadratic
  variance and insufficient data capacity for
  the combined model complexity.
  The use of EDT rather than calendar time as the
  LGCM time metric accounts for heterogeneous
  disease progression rates across participants,
  though it introduces dependence on the progmod
  estimation
- Bootstrap CIs confirmed that most b-path effects
  were significant (4/6 for both FRS and CVR~mimic~;
  CIs excluding zero), consistent with nested LRT
  results. Language b-paths were non-significant,
  likely reflecting lower statistical power for
  this domain
- **Sensitivity summary:**
  - HC z-score effects were weaker than HVR z-score
    effects (Appendix C, @tbl-hc-comparison),
    supporting the
    ratio measure's greater sensitivity
  - Systematic baseline differences between completers
    and dropouts (Appendix C,
    @tbl-attrition-baseline) suggest
    observed effects may underestimate true CVR-brain
    relationships
  - Raw (un-normalized) HVR and HC results were
    concordant with z-scored analyses, confirming the
    normative transformation does not drive findings
  - Limited EDT moderation suggests CVR effects are
    stable across disease stages
  - SE-constrained LGCM validation showed comparable
    bias, RMSE, and coverage to free estimation

## Clinical Implications

- In clinical trials targeting cardiovascular risk
  reduction for dementia prevention, FRS should not
  be used as the sole CVR endpoint in elderly samples
  due to age-related variance compression and
  age-confounding
  [@ko_calibration_2020; @marrugat_adaptation_2003]
- Age-adjusted CVR measures (via MIMIC models or
  residualization) should be considered when the
  research question concerns cardiovascular effects
  independent of aging [@zhang_bayesian_2024]
- HVR provides a more sensitive neuroimaging endpoint
  than hippocampal volume alone, and the consistent
  HVR-cognition coupling (b-path) supports its use
  as a surrogate outcome in intervention studies
- The female CVR~mimic~ Executive Function LME
  3-way interaction was the sole significant
  CVR~mimic~ moderation result out of 6 tests
  (2 sexes × 3 domains), and did not survive FDR
  correction. This isolated finding should be
  treated as exploratory and requires independent
  replication before any clinical interpretation

## Limitations

- Single cohort (ADNI); replication in other aging
  samples needed. The UKB-trained GAMLSS normative
  models showed a systematic negative z-score offset
  in the ADNI comparable subsample, driven by age
  distribution mismatch. While within-ADNI relative
  comparisons remain valid, absolute z-score values
  should be interpreted cautiously when comparing
  across cohorts
- CVR~mimic~ indicators limited to available CVRF
  data; more comprehensive indicators (e.g., lipid
  panel, inflammatory markers) could improve the
  latent CVR measure
- Cross-sectional CVR measurement; longitudinal CVR
  trajectories could be informative
- Amyloid-positive sample only; findings may not
  generalize to cognitively normal individuals
  without AD pathology
- The formal coefficient comparison (shared-seed
  bootstrap Δ~a~ and clustered bootstrap Δβ) relies
  on the same resampling framework as the primary
  mediation inference; independent validation of
  these paired tests is not available
- Female CVR~mimic~ Executive LME effect requires
  replication to determine whether it reflects true
  CVR moderation or a chance finding
- The fixed-quadratic LGCM captures population-level
  curvature but constrains individual quadratic
  variation; slope estimates may still obscure
  individual-level acceleration or deceleration
  patterns
- LGCM models used individually-varying time scores
  (EDT definition variables), which preclude
  computation of CFI, TLI, and RMSEA — we assessed
  model adequacy via information criteria and
  nested LRTs only, limiting comparability with
  studies reporting standard fit indices
- MIMIC model used only 5 indicators; richer
  measurement models may yield different results.
  Factor scores are model-dependent and different
  specifications could yield different CVR~mimic~
  values
- Common-sample filtering may reduce N relative to
  analyses using only one CVR measure, potentially
  limiting power
- Bootstrap mediation CIs (N =
  `r med_n.lst$Male` males,
  `r med_n.lst$Female` females) may be liberal with
  these modest sample sizes; results should be
  interpreted with caution

## Conclusions

- Whether CVR appears to influence
  hippocampal-cognitive trajectories depends
  critically on how CVR is measured: FRS produces
  significant indirect effects through hippocampal
  decline, while age-adjusted CVR~mimic~ does not
- This dissociation supports the hypothesis that FRS
  "effects" on hippocampal decline primarily reflect
  age-related variance rather than cardiovascular
  pathophysiology
- Suggestive single-domain evidence (Female Executive
  LME 3-way) hints that age-adjusted CVR may still
  modulate the brain-cognition relationship through
  reserve-like mechanisms
- The hippocampal-cognitive coupling persists
  regardless of CVR measurement, supporting HVR as
  a reliable biomarker target
- Future directions include replication in
  independent aging cohorts (e.g., WRAP, AIBL),
  longitudinal CVR measurement to capture
  within-person CVR trajectories, and intervention
  studies testing whether CVR reduction preserves
  the HVR-cognition coupling
- This study demonstrates a general methodological
  framework — MIMIC-based age adjustment — that can
  be applied to any composite risk score where age
  confounding is a concern

---

\newpage

:::{=latex}
\appendix
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}
:::

# Normative Transfer Validation


- For transfer validation, we identified a comparable
  ADNI subsample of `r adni_demo$n_total`
  amyloid-negative cognitively unimpaired participants
  (`r sprintf("%.0f%%", adni_demo$pct_female)`
  female) within the UKB age range
- Despite matching on amyloid status and cognitive normality, the
  ADNI comparable subsample was approximately
  `r sprintf("%.0f", adni_demo$age_mean - ukb_demo$age_mean)`
  years older on average (mean
  `r sprintf("%.1f", adni_demo$age_mean)`
  vs
  `r sprintf("%.1f", ukb_demo$age_mean)`
  years; @fig-normative-validation A), concentrating in the upper
  tail of the UKB age distribution
- HVR z-scores showed a systematic negative offset (mean =
  `r sprintf("%.2f", comp_hvr$mean)`,
  SD =
  `r sprintf("%.2f", comp_hvr$sd)`,
  95% CI
  [`r sprintf("%.2f", comp_hvr$ci_95["lower"])`,
  `r sprintf("%.2f", comp_hvr$ci_95["upper"])`]),
  while HC z-scores were closer to expected values (mean =
  `r sprintf("%.2f", comp_hc$mean)`,
  SD =
  `r sprintf("%.2f", comp_hc$sd)`;
  @fig-normative-validation B)
- Age-bin-specific calibration (see @fig-normative-validation, panel B):
  mean z-scores near zero in younger bins, increasingly
  negative offset in older bins consistent with the age
  mismatch

```{r fig-normative-validation}
#| fig-cap: "Normative transfer validation. (A) Age distributions: UK Biobank normative sample (yellow) vs ADNI comparable subsample (cyan); dashed lines indicate sample means. (B) Forest plot of HVR (purple) and HC (green) z-scores (mean +/- 95% CI) by age bin; reference line at z = 0 indicates perfect calibration; open circles = age bins, filled circles = overall. (C) Z-score distributions for HVR (left) and HC (right) compared to the expected N(0,1) (yellow)."
#| fig-width: 6.5
#| fig-height: 6

stopifnot(!is.null(nv_data))
ukb_d <- nv_data$comparable_subsample$demographics$ukb_normative
adni_d <- nv_data$comparable_subsample$demographics$adni_comparable

p_age <- plot_age_distribution_comparison(
  ukb_age_mean = ukb_d$age_mean,
  ukb_age_sd = ukb_d$age_sd,
  ukb_age_min = ukb_d$age_min,
  ukb_age_max = ukb_d$age_max,
  ukb_n = ukb_d$n_total,
  adni_age_mean = adni_d$age_mean,
  adni_age_sd = adni_d$age_sd,
  adni_n = adni_d$n_total
)

p_forest <- plot_zscore_validation_forest(nv_data)

# Build z-score data.table with both measures
hvr_z.v <- comp_hvr$zscores
hc_z.v <- comp_hc$zscores
stopifnot(!is.null(hvr_z.v), !is.null(hc_z.v))
n_max <- max(length(hvr_z.v), length(hc_z.v))
zscore.dt <- data.table(
  PTID = seq_len(n_max),
  HVR_z = hvr_z.v[seq_len(n_max)],
  HC_z = hc_z.v[seq_len(n_max)]
)

p_dist <- plot_zscore_distribution(
  zscore.dt, sample_label = "Comparable"
)

(p_age | p_forest) / p_dist +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(tag_levels = "A")
```


:::{=latex}
\FloatBarrier
\newpage
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{B\arabic{table}}
\renewcommand{\thefigure}{B\arabic{figure}}
:::

# Methodological Validations

## CVR MIMIC Measurement Invariance

Measurement invariance of the CVR MIMIC model was
tested across sex in four steps (configural, metric,
scalar, strict; @tbl-mi).

```{r tbl-mi}
#| tbl-cap: "Measurement invariance of the CVR MIMIC model across sex. Configural = same structure; Metric = equal loadings; Scalar = equal intercepts; Strict = equal residuals. ΔCFI < 0.01 and ΔRMSEA < 0.015 indicate acceptable invariance."

stopifnot(!is.null(cvr_mi.res))
require_output(
  create_mi_table(
    cvr_mi.res,
    "Measurement Invariance Across Sex"
  ),
  "tbl-mi"
)
```

## EDT Linearity and Simulation Validation

- Linear time specification validated via LME
  likelihood ratio tests comparing linear vs
  quadratic EDT across all sex-domain models
  (see @tbl-linearity)

```{r tbl-linearity}
#| tbl-cap: "EDT linearity validation: LME models testing linear vs quadratic EDT specification. LRT tests whether adding a quadratic EDT term significantly improves fit. ΔAIC = AIC(linear) - AIC(quadratic); positive values favor quadratic. EDT = Estimated Disease Time."

require_output(
  create_linearity_table(
    edt_linearity,
    title = "EDT Linearity Validation"
  ),
  "tbl-linearity"
)
```

```{r edt-linearity-narrative}
#| output: asis

stopifnot(!is.null(edt_linearity))
lt <- as.data.table(
  edt_linearity$results_table
)
# Column is LRT_p, not p
n_sig <- sum(
  lt$LRT_p < 0.05, na.rm = TRUE
)
n_total <- nrow(lt)
if (n_sig == 0) {
  cat(paste0(
    "- No domain showed a significant",
    " quadratic EDT term (all p > 0.05),",
    " supporting the linear time",
    " specification for the LGCM\n"
  ))
} else {
  sig_doms <- lt[LRT_p < 0.05, Domain]
  max_chisq <- max(
    lt$LRT_chisq, na.rm = TRUE
  )
  cat(sprintf(
    paste0(
      "- All %d/%d models showed",
      " significant quadratic EDT terms",
      " (%s; max \u03c7\u00b2 = %.1f)\n"
    ),
    n_sig, n_total,
    paste(
      DOMAIN_LABELS[sig_doms],
      collapse = ", "
    ),
    max_chisq
  ))
  cat(paste0(
    "- Large test statistics reflect",
    " substantial power afforded by",
    " sample size and repeated",
    " observations\n",
    "- Linear specification adopted as",
    " pragmatic compromise given",
    " convergence failure of quadratic",
    " parallel process LGCMs",
    " (see Methods)\n"
  ))
}
```

## SE-Constrained LGCM Validation

- Monte Carlo simulation
  (`r simulation.res$config$n_reps` replications
  at N = `r paste(simulation.res$config$sample_sizes, collapse = " and N = ")`)
  compared SE-constrained LGCM (residual variance
  fixed to SE²) against free estimation

```{r simulation-interpretation}
#| output: asis

stopifnot(!is.null(sim_conv))
fmt_sci <- function(x) {
  if (abs(x) < 1e-10) return("0")
  exp <- floor(log10(abs(x)))
  mant <- x / 10^exp
  sprintf("%.1f \u00d7 10^%d^", mant, exp)
}
for (i in seq_len(nrow(sim_conv))) {
  n_val <- sim_conv$n[i]
  cat(sprintf(
    paste0(
      "- N = %d: SE-constrained converged",
      " %d/500 (free: %d/500); bias = %s",
      " (free: %s); RMSE = %.3f",
      " (free: %.3f); 95%% CI coverage",
      " = %.1f%% (free: %.1f%%)\n"
    ),
    n_val,
    sim_conv$n_converged_se[i],
    sim_conv$n_converged_free[i],
    fmt_sci(sim_conv$bias_se[i]),
    fmt_sci(sim_conv$bias_free[i]),
    sim_conv$rmse_se[i],
    sim_conv$rmse_free[i],
    sim_conv$coverage_se[i] * 100,
    sim_conv$coverage_free[i] * 100
  ))
}
cat(paste0(
  "- Both approaches showed comparable",
  " performance, validating the",
  " SE-constrained specification used",
  " in the primary analyses\n"
))
```

## LME Covariate Effects

Full covariate effects from the sex-stratified
LME 3-way interaction models are presented for
the FRS model (@tbl-lme-covariates) and the
CVR~mimic~ model (@tbl-cvr-lme-covariates).

```{r tbl-lme-covariates}
#| tbl-cap: "Covariate effects from the FRS LME models, sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    frs_hvr, cvr_label = "FRS",
    title = "Covariate Effects (FRS Model)"
  ),
  "tbl-lme-covariates"
)
```

```{r tbl-cvr-lme-covariates}
#| tbl-cap: "Covariate effects from the CVR~mimic~ LME models, sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    cvr_hvr,
    cvr_label = "CVR~mimic~",
    title = md(
      "Covariate Effects (CVR~mimic~ Model)"
    )
  ),
  "tbl-cvr-lme-covariates"
)
```


:::{=latex}
\FloatBarrier
\newpage
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{C\arabic{table}}
\renewcommand{\thefigure}{C\arabic{figure}}
:::

# Sensitivity Analyses

All sensitivity checks converge on the same
pattern: FRS effects trace to age-weighting,
CVR~mimic~ effects are null or isolated, and the
HVR–cognition coupling (b-path) is consistent. Main
findings are stable across alternative brain
measures (HC volume), raw scores, attrition
adjustments, and EDT specification checks. Full
details follow.

## LME Sensitivity

### HC z-score (Total Hippocampal Volume)

- HC z-score effects were generally weaker than HVR
  z-score effects, with fewer significant 3-way
  interactions, supporting the hippocampus-to-ventricle
  ratio as a more sensitive brain aging biomarker
- LME 3-way interaction repeated with HC z-score
  instead of HVR
- Results in @tbl-hc-comparison; covariate effects in
  @tbl-hc-covariates and @tbl-hc-covariates-cvr

```{r tbl-hc-comparison}
#| tbl-cap: "Sensitivity analysis: LME 3-way interaction effects (YRS × CVR × HC z-score) for total hippocampal volume. FRS and CVR~mimic~ compared side-by-side. Bold indicates p < 0.05."

require_output(
  create_lme_comparison_table(
    frs_hc, cvr_hc,
    summary_dt = hc_summary.dt
  ),
  "tbl-hc-comparison"
)
```

```{r tbl-hc-covariates}
#| tbl-cap: "Covariate effects from the FRS HC z-score LME models, sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    frs_hc, cvr_label = "FRS",
    title = "HC Covariate Effects (FRS Model)"
  ),
  "tbl-hc-covariates"
)
```

```{r tbl-hc-covariates-cvr}
#| tbl-cap: "Covariate effects from the CVR~mimic~ HC z-score LME models, sex-stratified. Bold indicates p < 0.05."

require_output(
  create_lme_covariate_table(
    cvr_hc, cvr_label = "CVR~mimic~",
    title = md(
      "HC Covariate Effects (CVR~mimic~ Model)"
    )
  ),
  "tbl-hc-covariates-cvr"
)
```

```{r hc-interpretation}
#| output: asis

stopifnot(!is.null(hc_summary.dt))
  hc_p <- hc_summary.dt[
    Analysis == "Primary_Stratified"
  ]
  for (cvr_nm in c("FRS", "CVR_mimic")) {
    cvr_rows <- hc_p[CVR_Measure == cvr_nm]
    sig_rows <- cvr_rows[Significant == TRUE]
    cvr_lab <- ifelse(
      cvr_nm == "CVR_mimic",
      "CVR~mimic~", cvr_nm
    )
    if (nrow(sig_rows) > 0) {
      sig_desc <- paste(
        sprintf(
          "%s %s",
          sig_rows$Sex,
          DOMAIN_LABELS[sig_rows$Domain]
        ),
        collapse = ", "
      )
      cat(sprintf(
        "- %s × HC × YRS: significant for %s\n",
        cvr_lab, sig_desc
      ))
    } else {
      cat(sprintf(
        "- %s × HC × YRS: no significant effects\n",
        cvr_lab
      ))
    }
  }
  cat(
    "- HC z-score showed weaker effects than HVR\n",
    "  z-score, supporting use of the ratio measure\n"
  )
```

```{r hc-vs-hvr-synthesis}
#| output: asis

# Count significant HC vs HVR effects
hc_p <- hc_summary.dt[
  Analysis == "Primary_Stratified"
]
hvr_p <- lme_summary.dt[
  Analysis == "Primary_Stratified"
]
n_hc_sig <- sum(hc_p$Significant, na.rm = TRUE)
n_hvr_sig <- sum(
  hvr_p$Significant, na.rm = TRUE
)
cat(sprintf(
  paste0(
    "- HC z-score: %d/12 significant",
    " 3-way interactions vs HVR z-score",
    " %d/12 — weaker HC effects support",
    " the specificity of HVR: capturing",
    " ventricular expansion alongside",
    " hippocampal atrophy provides",
    " greater sensitivity to CVR",
    " moderation\n"
  ),
  n_hc_sig, n_hvr_sig
))
```

### Attrition Analysis

- Baseline characteristics compared between completers (≥4 visits)
  and dropouts (<4 visits)
- Results in @tbl-attrition-baseline

```{r tbl-attrition-baseline}
#| tbl-cap: "Attrition analysis: baseline characteristics of completers (≥4 visits) vs dropouts (<4 visits). Bold rows indicate p < 0.05. Completers = participants with sufficient visits for quadratic trend estimation."

require_output(
  create_attrition_table(attrition_analysis),
  "tbl-attrition-baseline"
)
```

```{r attrition-interpretation}
#| output: asis

stopifnot(
  !is.null(attrition_analysis),
  !is.null(attrition_analysis$comparison_table)
)
{
  comp_dt <- attrition_analysis$comparison_table
  sig_vars <- comp_dt[p_value < 0.05, Variable]
  # Map variable names for readability
  var_display <- c(
    AGE = "age", EDUC = "education",
    FRS = "FRS", HVR_z = "HVR z-score",
    HC_z = "HC z-score", SEX = "sex",
    APOE4 = "APOE \u03b54 carrier status"
  )
  sig_labels <- vapply(
    sig_vars, function(v) {
      if (v %in% names(var_display)) {
        var_display[v]
      } else v
    }, character(1)
  )
  if (length(sig_vars) > 0) {
    cat(sprintf(
      paste0(
        "- Significant baseline differences",
        " between completers and dropouts",
        " in %s (all p < 0.05;",
        " see @tbl-attrition-baseline)\n"
      ),
      paste(sig_labels, collapse = ", ")
    ))
  } else {
    cat(paste0(
      "- No significant baseline",
      " differences between completers",
      " and dropouts",
      " (see @tbl-attrition-baseline),",
      " suggesting attrition was random",
      " with respect to measured",
      " variables\n"
    ))
  }
}
```

```{r attrition-logistic-inline}
#| output: asis

stopifnot(
  !is.null(attrition_analysis),
  !is.null(attrition_analysis$logistic)
)
log_model <- attrition_analysis$logistic
or_dt <- as.data.table(log_model$odds_ratios)
setnames(
  or_dt,
  c("2.5 %", "97.5 %"),
  c("CI_lower", "CI_upper"),
  skip_absent = TRUE
)
or_dt <- or_dt[term != "(Intercept)"]
or_dt[, sig := CI_lower > 1 | CI_upper < 1]

# Readable term labels
term_display <- c(
  FRS = "FRS", AGE = "age",
  APOE4TRUE = "APOE \u03b54 carrier status",
  SEXMale = "male sex",
  HVR_z = "HVR z-score",
  HC_z = "HC z-score"
)
or_dt[, term_label := fifelse(
  term %in% names(term_display),
  term_display[term], term
)]

sig_preds <- or_dt[sig == TRUE]
if (nrow(sig_preds) > 0) {
  items <- sprintf(
    "%s (OR = %.2f, 95%% CI [%.2f, %.2f])",
    sig_preds$term_label,
    sig_preds$OR,
    sig_preds$CI_lower,
    sig_preds$CI_upper
  )
  cat(sprintf(
    paste0(
      "- Logistic regression predicting",
      " dropout: %d significant",
      " predictor%s — %s\n"
    ),
    nrow(sig_preds),
    ifelse(nrow(sig_preds) > 1, "s", ""),
    paste(items, collapse = "; ")
  ))
} else {
  cat(paste0(
    "- Logistic regression predicting",
    " dropout: no significant predictors",
    " (all 95%% CIs included 1)\n"
  ))
}

# FRS direction narrative
comp_dt <- attrition_analysis$comparison_table
frs_row <- comp_dt[
  grepl("FRS", Variable, ignore.case = TRUE)
]
if (nrow(frs_row) > 0) {
  c_mean <- frs_row$Completers_mean[1]
  d_mean <- frs_row$Dropouts_mean[1]
  if (!is.null(c_mean) && !is.null(d_mean) &&
      !is.na(c_mean) && !is.na(d_mean)) {
    direction <- if (d_mean > c_mean) {
      "higher"
    } else "lower"
    cat(sprintf(
      paste0(
        "- Dropouts had %s baseline FRS",
        " (M = %.2f) than completers",
        " (M = %.2f), consistent with",
        " selective loss of higher-risk",
        " participants — may attenuate",
        " observed CVR effects\n"
      ),
      direction, d_mean, c_mean
    ))
  }
}
```

### Raw Brain Measure Sensitivity

- LME 3-way interaction repeated using raw
  (un-normalized) HVR and raw HC (with TIV
  covariate); FRS predictor, full-sample, no sex
  stratification, no YRS² term

```{r raw-hvr-inline}
#| output: asis

stopifnot(
  !is.null(lme_raw_sens.res),
  !is.null(lme_raw_sens.res$results)
)
hvr_raw <- lme_raw_sens.res$results$HVR_raw_FRS
stopifnot(!is.null(hvr_raw))
raw_hvr_results <- list(results = hvr_raw)

hc_raw <- lme_raw_sens.res$results$HC_raw_FRS
stopifnot(!is.null(hc_raw))
raw_hc_results <- list(results = hc_raw)

for (meas_nm in c("Raw HVR", "Raw HC (+TIV)")) {
  res <- if (meas_nm == "Raw HVR") {
    raw_hvr_results
  } else raw_hc_results
  items <- c()
  for (d in c("MEM", "LAN", "EXF")) {
    r <- res$results[[d]]$interaction
    if (!is.null(r)) {
      items <- c(items, sprintf(
        "%s \u03b2 = %.4f, %s",
        DOMAIN_LABELS[d], r$beta,
        fmt_p_inline(r$p_value)
      ))
    }
  }
  cat(sprintf(
    "- %s: %s\n",
    meas_nm, paste(items, collapse = "; ")
  ))
}

# Concordance check
sig_hvr <- sapply(
  names(raw_hvr_results$results),
  function(d) {
    r <- raw_hvr_results$results[[d]]$interaction
    !is.null(r) && r$p_value < 0.05
  }
)
if (any(sig_hvr)) {
  cat(sprintf(
    paste0(
      "- Raw HVR \u00d7 FRS significant for",
      " %s, confirming normative z-scoring",
      " does not drive primary findings\n"
    ),
    paste(
      DOMAIN_LABELS[names(sig_hvr)[sig_hvr]],
      collapse = ", "
    )
  ))
} else {
  cat(paste0(
    "- Raw HVR \u00d7 FRS not significant",
    " for any domain, suggesting",
    " normative z-scoring enhances",
    " sensitivity\n"
  ))
}

# Raw HC (+TIV) note
sig_hc <- sapply(
  names(raw_hc_results$results),
  function(d) {
    r <- raw_hc_results$results[[d]]$interaction
    !is.null(r) && r$p_value < 0.05
  }
)
if (!any(sig_hc)) {
  cat(paste0(
    "- Raw HC (+TIV) not significant",
    " for any domain, reinforcing the",
    " added sensitivity of the ratio",
    " measure\n"
  ))
}
```

### EDT Moderation of LME Effects

- 4-way interaction (CVR × HVR × EDT × YRS) tested
  whether CVR effects vary by disease stage

```{r edt-moderation-inline}
#| output: asis

report_edt <- function(edt_res, cvr_label) {
  stopifnot(
    !is.null(edt_res),
    !is.null(edt_res$summary_continuous)
  )
  sc <- as.data.table(edt_res$summary_continuous)
  for (sx in c("Male", "Female")) {
    sx_rows <- sc[Sex == sx]
    sig <- sx_rows[Significant == TRUE]
    if (nrow(sig) > 0) {
      items <- sprintf(
        "%s (\u03b2 = %.4f, %s)",
        DOMAIN_LABELS[sig$Domain],
        sig$Beta,
        sapply(sig$p_value, fmt_p_inline)
      )
      cat(sprintf(
        "- %s \u00d7 HVR \u00d7 EDT (%s):",
        cvr_label, sx
      ))
      cat(sprintf(
        " significant for %s\n",
        paste(items, collapse = "; ")
      ))
    } else {
      min_p <- min(
        sx_rows$p_value, na.rm = TRUE
      )
      cat(sprintf(
        paste0(
          "- %s \u00d7 HVR \u00d7 EDT (%s):",
          " not significant (min %s)\n"
        ),
        cvr_label, sx,
        fmt_p_inline(min_p)
      ))
    }
  }
}
report_edt(frs_edt, "FRS")
report_edt(cvr_edt, "CVR~mimic~")
cat(paste0(
  "- Limited EDT moderation suggests",
  " CVR effects on HVR-cognition",
  " coupling are relatively stable",
  " across disease stages\n"
))
```

### EDT Stage Sensitivity

- Analyses restricted to early and middle EDT
  tertiles (excluding late-stage) to assess
  robustness

```{r edt-sensitivity-inline}
#| output: asis

stopifnot(!is.null(edt_sens.res))
rt <- as.data.table(edt_sens.res$results_table)

meas_labels <- c(
  HVR_z = "HVR z-score",
  HC_z = "HC z-score"
)
dom_labels <- c(
  MEM = "Memory", LAN = "Language",
  EXF = "Executive Function"
)

for (meas in unique(rt$Measure)) {
  mrt <- rt[Measure == meas]
  m_lab <- if (meas %in% names(meas_labels)) {
    meas_labels[meas]
  } else meas
  items <- c()
  n_concord <- 0L
  for (i in seq_len(nrow(mrt))) {
    row <- mrt[i]
    d_lab <- if (
      row$Domain %in% names(dom_labels)
    ) dom_labels[row$Domain] else row$Domain
    pct <- if (
      !is.null(row$Beta_change_pct) &&
      !is.na(row$Beta_change_pct)
    ) {
      sprintf(
        ", %.1f%% change",
        row$Beta_change_pct
      )
    } else ""
    if (!is.null(row$Consistent) &&
        isTRUE(row$Consistent)) {
      n_concord <- n_concord + 1L
    }
    items <- c(items, sprintf(
      "%s (\u03b2 = %.4f, %s%s)",
      d_lab, row$Beta_early_middle,
      fmt_p_inline(row$P_early_middle),
      pct
    ))
  }
  concord_note <- if (n_concord > 0) {
    sprintf(
      " (%d/%d concordant with full-sample)",
      n_concord, nrow(mrt)
    )
  } else ""
  cat(sprintf(
    "- %s: %s%s\n",
    m_lab,
    paste(items, collapse = "; "),
    concord_note
  ))
}
```

:::{=latex}
\FloatBarrier
:::

---

# References {.unnumbered}

::: {#refs}
:::
